---
title: "Take-home-Exercise 3"
author: "Tai QiuYan"
date: "October 31, 2025"
date-modified: "last-modified"
format:
  html:
    code-fold: show
editor: visual
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# 1 Introduction

This exercise develops a spatial, data-driven framework to explain and predict resale prices of 5-room HDB flats in Singapore by integrating structural attributes (floor area, floor level, remaining lease) with detailed measures of locational accessibility. Multiple open datasets are assembled and geocoded via OneMap, harmonised into a common SVY21/EPSG:3414 projection, and cleaned through geometry validation and removal of out-of-bounds features. Proximity features (nearest distance to CBD, MRT, hawker centres, parks, shopping malls, supermarkets, clinics, eldercare and good primary schools) and density features (counts of kindergartens, childcare centres, bus stops, preschools and schools within walkable buffers) are engineered to capture neighbourhood accessibility. Exploratory data analysis is used to examine distributions and spatial patterns, while correlation analysis and VIF diagnostics are applied to mitigate multicollinearity. Three modelling strategies are then calibrated on a rolling time split (train: Jul 2024–Jun 2025; test: Jul–Sep 2025): a global Multiple Linear Regression (MLR) hedonic model, a Geographically Weighted Random Forest (GWRF), and a Geographically Weighted Regression (GWR) with spatially varying coefficients. Model performance is evaluated using RMSE, prediction–observation plots, residual mapping, Moran’s I tests for spatial autocorrelation, in order to assess both predictive accuracy and the extent of spatial non-stationarity in price determinants.

# 2 Loading the R package

In this exercise, the following packages are used:

sf - Handles spatial vector data (points, lines, polygons) and performs spatial operations in R with a modern data frame approach.

sp - Provides classes and methods for spatial data handling (legacy package, predecessor to sf).

spdep - Performs spatial dependence analysis including spatial autocorrelation tests and spatial regression modeling.

olsrr - Offers comprehensive tools for building and diagnosing ordinary least squares regression models.

ggplot2 - Creates elegant data visualizations using a layered grammar of graphics approach.

plotly - Converts static ggplot2 graphics into interactive web-based visualizations.

SpatialML - Implements geographically weighted random forest and other spatial machine learning algorithms.

tmap - Produces thematic maps with a flexible layer-based syntax for spatial data visualization.

Metrics - Calculates common machine learning evaluation metrics including RMSE, MAE, and R-squared.

tidyverse - A collection of packages (dplyr, tidyr, etc.) for data manipulation, transformation, and analysis with consistent syntax.

httr - Provides tools for working with HTTP requests and APIs to retrieve web-based data.

jsonlite - Parses and generates JSON data for exchanging information with web services and APIs.

progress - Displays progress bars in R console for monitoring long-running operations.

ggpubr - Creates publication-ready plots with statistical annotations and easy customization of ggplot2 graphics.

stringr - Simplifies string manipulation with consistent functions for pattern matching, extraction, and replacement.

patchwork - Combines multiple ggplot2 plots into composite figures with simple and intuitive syntax.

ggstatsplot - Creates graphics with statistical details including correlation matrices and statistical test results.

GWmodel - Implements geographically weighted regression and other spatial analysis techniques for modeling local relationships.

```{r}
pacman::p_load(sf, sp, spdep, olsrr, ggplot2, plotly, SpatialML,
               tmap, Metrics, tidyverse, httr, jsonlite, progress,
               ggpubr, stringr,patchwork,ggstatsplot,GWmodel)
```

# 3 Data Importing and Wrangling

Before we start to explore, we need to import Master Plan 2019 Subzone Boundary data of Singapore.

```{r}
mpsz = read_rds("data/rds/mpsz_cl.rds")
```

## 3.1 Structural factors data

The code below imports HDB resale data.

```{r}
#| eval: false
#| output: false
HDBresale <- read_csv(
  "data/rawdata/HDBresale.csv") %>%
  filter(flat_type == "5 ROOM") %>%
  filter(month >= "2024-07" & month <= "2025-09")
```

The code below standardizes street name formatting by replacing all abbreviated "ST." (Street) with the full word "SAINT" in the street_name column of the HDBresale dataset.

```{r}
#| eval: false
#| output: false
HDBresale$street_name <- gsub(
  "ST\\.", "SAINT",
  HDBresale$street_name)
```

### Geocoding

The code below performs several key tasks to prepare and retrieve geospatial data. First, it combines the block number and street name into a full address. This address is then used as the search value (searchVal) in a query sent to the OneMapSG API. The API response, which is returned in JSON format, is converted into text and subsequently stored in a dataframe. From this dataframe, the latitude and longitude values are extracted and retained as the final output for further analysis or mapping.

```{r}
#| eval: false
#| output: false
geocode <- function(block, streetname) {
  base_url <- "https://onemap.gov.sg/api/common/elastic/search"
  address <- paste(block, streetname, sep = " ")
  query <- list("searchVal" = address, 
                "returnGeom" = "Y",
                "getAddrDetails" = "N",
                "pageNum" = "1")
  
  res <- GET(base_url, query = query)
  restext<-content(res, as="text")
  
  output <- fromJSON(restext)  %>% 
    as.data.frame %>%
    select(results.LATITUDE, results.LONGITUDE)

  return(output)
}
```

```{r}
#| eval: false
#| output: false
HDBresale$LATITUDE <- 0
HDBresale$LONGITUDE <- 0

for (i in 1:nrow(HDBresale)){
  temp_output <- geocode(HDBresale[i, 4], HDBresale[i, 5])
  
  HDBresale$LATITUDE[i] <- temp_output$results.LATITUDE
  HDBresale$LONGITUDE[i] <- temp_output$results.LONGITUDE
}
```

### Converting to an sf object

The code below converts the HDB resale dataset into an sf spatial object using longitude and latitude, transforms it to Singapore’s coordinate system (EPSG:3414), saves it as an .rds file, and reloads it for future use.

```{r}
#| eval: false
#| output: false
HDBresale <- HDBresale %>%
  st_as_sf(
    coords = c("LONGITUDE", "LATITUDE"),
    crs=4326) %>%
  st_transform(crs = 3414)
```

```{r}
#| eval: false
write_rds(HDBresale, "data/rds/HDBresale.rds")
```

```{r}
#| echo: false
HDBresale <- read_rds("data/rds/HDBresale.rds")
```

### Floor Level

The code below checks the storey_range column in HDBresale data.

```{r}
unique(HDBresale$storey_range)
```

To make the column more usable, a new numeric attribute is created to captures the midpoint of each range.

```{r}
HDBresale <- HDBresale %>%
  mutate(
    floor_level = (as.numeric(str_extract(storey_range, "^[0-9]+")) +
                     as.numeric(str_extract(storey_range, "[0-9]+$"))) / 2
  )
```

### Remaining Lease & Unit’s Age

This code extracts the remaining lease years and months, converts them into a decimal remaining lease value, calculates the unit age based on a 99-year lease, and removes intermediate columns.

```{r}
HDBresale <- HDBresale %>%
  mutate(
    remaining_lease_years = as.integer(str_sub(remaining_lease, 0, 2)),
    remaining_lease_months = as.integer(str_sub(remaining_lease, 9, 11)),
    
    # Replace NA in months with 0 if "months" part is missing
    remaining_lease_months = if_else(is.na(remaining_lease_months), 0, remaining_lease_months),
    
    # Calculate remaining lease in decimal years
    remaining_lease = remaining_lease_years + (remaining_lease_months / 12),
    
    # Calculate unit age based on a 99-year lease
    unit_age = 99 - remaining_lease
  ) %>%
  select(-remaining_lease_years, -remaining_lease_months)
```

### Keep relevant variables

This code keeps only the relevant variables in the HDB resale dataset for further analysis and removes unnecessary columns.

```{r}
HDBresale <- HDBresale %>%
  select(
    month,
    town,
    block,
    street_name,
    flat_type,
    floor_area_sqm,
    floor_level,
    remaining_lease,
    unit_age,
    resale_price,
    geometry
  )
```

## 3.2 Location factors data

This code loads various location factor datasets including eldercare centres, hawker/food centres, MRT stations, parks, schools (all, primary, and top primary schools), shopping malls, supermarkets, kindergartens, childcare centres, bus stops, CHAS clinics, and preschools, converts their coordinates to Singapore’s SVY21 system (EPSG:3414), identifies the CBD location, and prepares them for spatial analysis and mapping.

```{r}
# CBD

cbd_poly <- mpsz %>%
  filter(PLN_AREA_N == "DOWNTOWN CORE") %>%
  st_union()

cbd <- st_point_on_surface(cbd_poly)

# Eldercare

eldercare <- st_read("data/rawdata/EldercareServices.geojson") %>%
  st_transform(crs = 3414)

# Foodcourt/hawker centres

hawker <- st_read("data/rawdata/HawkerCentresGEOJSON.geojson") %>%
  st_transform(3414)

# MRT

mrt_station <- st_read("data/rawdata/RapidTransitSystemStation/RapidTransitSystemStation.shp") %>%
  st_transform(3414)

# Park

parks <- st_read("data/rawdata/Parks.geojson") %>%
  st_transform(3414)

# School

schools <- read_csv("data/rawdata/Generalinformationofschools.csv")

# Primary school

primary_schools <- schools %>%
  filter(mainlevel_code == "PRIMARY")

# Good primary school

good_primary_school_list <- c(
 "NANYANG PRIMARY SCHOOL", "TAO NAN SCHOOL", "Hong Wen School", "NAN HUA PRIMARY SCHOOL", "ST. HILDA'S PRIMARY SCHOOL", "HENRY PARK PRIMARY SCHOOL", "ANGLO-CHINESE SCHOOL (PRIMARY)", "RAFFLES GIRLS' PRIMARY SCHOOL", "PEI HWA PRESBYTERIAN PRIMARY SCHOOL", "Radin Mas Primary School", "ROSYTH SCHOOL", "KONG HWA SCHOOL", "POI CHING SCHOOL", "HOLY INNOCENTS' PRIMARY SCHOOL", "AI TONG SCHOOL", "RED SWASTIKA SCHOOL", "Maha Bodhi School", "RULANG PRIMARY SCHOOL", "PEI CHUN PUBLIC SCHOOL", "SINGAPORE CHINESE GIRLS' PRIMARY SCHOOL"
)

good_primary_schools <- primary_schools %>%
  filter(school_name %in% toupper(good_primary_school_list))

# Shopping mall

shopping <- read_csv("data/rawdata/mall_coordinates_updated.csv") %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(3414)

# Supermarket

supermarkets <- st_read("data/rawdata/SupermarketsGEOJSON.geojson") %>%
  st_transform(3414)

# Kindergartens

kindergartens <- st_read("data/rawdata/Kindergartens.geojson") %>%
  st_transform(3414)

# Childcare centres

childcare <- st_read("data/rawdata/ChildCareServices.geojson") %>%
  st_transform(3414)

# Bus stop

bus_stop <- st_read("data/rawdata/BusStop/BusStop.shp") %>%
  st_transform(3414)

# CHAS Clinics

clinics <- st_read("data/rawdata/CHASClinics.geojson") %>%
  st_transform(3414)

# Preschool

preschool = st_read(
  "data/rawdata/PreSchoolsLocation.kml") %>%
  st_transform(crs= 3414)
```

### Geocoding for school, primary school and good primary school

The code below performs geocoding using the Singapore OneMap API, transforming postal codes and address information of schools into geographic coordinates for spatial analysis.

```{r}
get_coords <- function(search_list){
  
  coord_df <- data.frame()
  
  for (item in search_list){
    r <- GET('https://www.onemap.gov.sg/api/common/elastic/search?',
             query=list(searchVal=item,
                        returnGeom='Y',
                        getAddrDetails='Y'))
    data <- fromJSON(rawToChar(r$content))
    found <- data$found
    res <- data$results
    
    new_row <- data.frame()
    
    if (found == 1){
      new_row <- data.frame(
        search_val = item,
        postal = res$POSTAL,
        latitude = res$LATITUDE,
        longitude = res$LONGITUDE
      )
    }
    
    else if (found > 1){
      res_sub <- res[res$POSTAL != "NIL", ]
      
      if (nrow(res_sub) == 0){
        new_row <- data.frame(search_val = item,
                              postal = NA,
                              latitude = NA,
                              longitude = NA)
      } else {
        top1 <- head(res_sub, 1)
        new_row <- data.frame(
          search_val = item,
          postal = top1$POSTAL,
          latitude = top1$LATITUDE,
          longitude = top1$LONGITUDE
        )
      }
    }
    
    else {
      new_row <- data.frame(search_val = item,
                            postal = NA,
                            latitude = NA,
                            longitude = NA)
    }
    
    coord_df <- rbind(coord_df, new_row)
  }
  
  return(coord_df)
}
```

```{r}
# School

school_postal_list <- schools$postal_code %>%
  as.character() %>%
  sprintf("%06s", .)

school_coords <- get_coords(school_postal_list)

schools <- bind_cols(
  schools,
  school_coords %>%
    transmute(
      latitude  = as.numeric(latitude),
      longitude = as.numeric(longitude)))


schools <- schools %>%
  filter(!is.na(longitude), !is.na(latitude)) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%  
  st_transform(3414) 

# Primary school

primary_postal_list <- primary_schools$postal_code %>%
  as.character() %>%
  sprintf("%06s", .)

primary_coords <- get_coords(primary_postal_list)

primary_schools <- bind_cols(
  primary_schools,
  primary_coords %>%
    transmute(
      latitude  = as.numeric(latitude),
      longitude = as.numeric(longitude)))

primary_schools <- primary_schools %>%
  filter(!is.na(longitude), !is.na(latitude)) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(3414)

# Good primary school

good_postal_list <- good_primary_schools$postal_code %>%
  as.character() %>%
  sprintf("%06s", .)

good_coords <- get_coords(good_postal_list)

good_primary_schools <- bind_cols(
  good_primary_schools,
  good_coords %>%
    transmute(
      latitude  = as.numeric(latitude),
      longitude = as.numeric(longitude)))

good_primary_schools <- good_primary_schools %>%
  filter(!is.na(longitude), !is.na(latitude)) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(3414)
```

### Z-Dimensions Removal

We’ll use the st_zm() function to remove the z-dimension from the data.

```{r}
eldercare <- st_zm(eldercare)
hawker <- st_zm(hawker)
mrt_station <- st_zm(mrt_station)
parks <- st_zm(parks)
schools <- st_zm(schools)
primary_schools <- st_zm(primary_schools)
good_primary_schools <- st_zm(good_primary_schools)
shopping <- st_zm(shopping)
supermarkets  <- st_zm(supermarkets)
kindergartens <- st_zm(kindergartens)
childcare <- st_zm(childcare)
bus_stop <- st_zm(bus_stop)
clinics <- st_zm(clinics)
preschool <- st_zm(preschool)
```

### Drop Unnecessary Columns

This code removes all unnecessary attribute columns from each dataset and keeps only the geometry column needed for spatial mapping.

```{r}
eldercare <- eldercare %>% select(c(1))
hawker <- hawker %>% select(c(1))
mrt_station <- mrt_station %>% select(c(1))
parks <- parks %>% select(c(1))
schools <- schools %>% select(c(1))
primary_schools <- primary_schools %>% select(c(1))
good_primary_schools <- good_primary_schools %>% select(c(1))
shopping <- shopping %>% select(c(1))
supermarkets <- supermarkets %>% select(c(1))
kindergartens <- kindergartens %>% select(c(1))
childcare <- childcare %>% select(c(1))
bus_stop <- bus_stop %>% select(c(1))
clinics <- clinics %>% select(c(1))
preschool <- preschool %>% select(c(1))
```

### Check Invalid Geometries

This code checks each spatial dataset for invalid geometries by counting the number of features with errors, ensuring data is clean for spatial analysis.

```{r}
length(which(st_is_valid(mpsz) == FALSE))
length(which(st_is_valid(cbd) == FALSE))
length(which(st_is_valid(eldercare) == FALSE))
length(which(st_is_valid(hawker) == FALSE))
length(which(st_is_valid(mrt_station) == FALSE))
length(which(st_is_valid(parks) == FALSE))
length(which(st_is_valid(schools) == FALSE))
length(which(st_is_valid(primary_schools) == FALSE))
length(which(st_is_valid(good_primary_schools) == FALSE))
length(which(st_is_valid(shopping) == FALSE))
length(which(st_is_valid(supermarkets) == FALSE))
length(which(st_is_valid(kindergartens) == FALSE))
length(which(st_is_valid(childcare) == FALSE))
length(which(st_is_valid(bus_stop) == FALSE))
length(which(st_is_valid(clinics) == FALSE))
length(which(st_is_valid(preschool) == FALSE))
```

As mpsz has invalid geometries, the code below fixes it using st_make_valid() function.

```{r}
mpsz <- st_make_valid(mpsz)
```

### MRT Geometry Correction

This code fixes invalid MRT station geometries by transforming them to point features, removing Z/M dimensions, handling empty polygons, and ensuring all geometries are valid point locations for mapping.

```{r}
sf_use_s2(FALSE)

mrt_station <- st_transform(mrt_station, 3414)
mrt_station <- st_zm(mrt_station, drop = TRUE, what = "ZM")

mrt_station <- st_collection_extract(mrt_station, "POLYGON", warn = FALSE)
mrt_station <- mrt_station[!st_is_empty(mrt_station), ]

safe_point <- function(g) {
  p <- try(st_point_on_surface(g), silent = TRUE)
  if (inherits(p, "try-error")) {
    bb <- st_bbox(g)
    st_point(c((bb["xmin"] + bb["xmax"]) / 2,
               (bb["ymin"] + bb["ymax"]) / 2))
  } else {
    st_geometry(p)[[1]]
  }
}

pts <- st_sfc(lapply(st_geometry(mrt_station), safe_point), crs = st_crs(mrt_station))
mrt_station <- st_set_geometry(mrt_station, pts)
mrt_station <- st_cast(mrt_station, "POINT", warn = FALSE)

# check result
unique(st_geometry_type(mrt_station))
```

### Remove Out-of-Bound Features

This code removes park and bus stop features that fall outside Singapore’s planning area boundaries (mpsz) by keeping only those located within the boundary.

```{r}
parks <- st_join(
  parks, mpsz, 
  join = st_within, 
  left = FALSE)

bus_stop <- st_join(
  bus_stop, mpsz, 
  join = st_within, 
  left = FALSE)
```

### Visualization

This code generates an interactive map to visualize multiple location including eldercare centres, hawker/food centres, MRT stations, parks, schools (all, primary, and top primary schools), shopping malls, supermarkets, kindergartens, childcare centres, bus stops, CHAS clinics, and preschools.

```{r}
tmap_mode("view")
tmap_style("classic")

tm_shape(eldercare) +
  tm_dots(col = "#1F77B4", size = 0.3, alpha = 0.7, legend.show = TRUE, group = "Eldercare") +
tm_shape(hawker) +
  tm_dots(col = "#FF7F0E", size = 0.3, alpha = 0.7, legend.show = TRUE, group = "Foodcourt/hawker centres") +
tm_shape(mrt_station) +
  tm_dots(col = "#2CA02C", size = 0.3, alpha = 0.7, legend.show = TRUE, group = "MRT") +
tm_shape(parks) +
  tm_dots(col = "#D62728", size = 0.3, alpha = 0.7, legend.show = TRUE, group = "Park") +
tm_shape(schools) +
  tm_dots(col = "#9467BD", size = 0.3, alpha = 0.7, legend.show = TRUE, group = "School") +
tm_shape(primary_schools) +
  tm_dots(col = "#8C564B", size = 0.3, alpha = 0.7, legend.show = TRUE, group = "Primary school") +
tm_shape(good_primary_schools) +
  tm_dots(col = "#E377C2", size = 0.3, alpha = 0.7, legend.show = TRUE, group = "Good primary school") +
tm_shape(shopping) +
  tm_dots(col = "#00FFFF", size = 0.3, alpha = 0.7, legend.show = TRUE, group = "Shopping mall") +
tm_shape(supermarkets) +
  tm_dots(col = "#7F7F7F", size = 0.3, alpha = 0.7, legend.show = TRUE, group = "Supermarket") +
tm_shape(kindergartens) +
  tm_dots(col = "#BCBD22", size = 0.3, alpha = 0.7, legend.show = TRUE, group = "Kindergarten") +
tm_shape(childcare) +
  tm_dots(col = "#17BECF", size = 0.3, alpha = 0.7, legend.show = TRUE, group = "Childcare centres") +
tm_shape(bus_stop) +
  tm_dots(col = "#FF00FF", size = 0.3, alpha = 0.7, legend.show = TRUE, group = "Bus stop") +
tm_shape(clinics) +
  tm_dots(col = "#39FF14", size = 0.3, alpha = 0.7, legend.show = TRUE, group = "CHAS clinics") +
tm_shape(preschool) +
  tm_dots(col = "#FFD300", size = 0.3, alpha = 0.7, legend.show = TRUE, group = "Preschool") +
  
tm_layout(legend.outside = TRUE,
          legend.title.size = 1.2,
          legend.text.size = 0.9) +
tm_view(set_zoom_limits = c(12,16))
```

### Proximity Calculation

The code below calculates the distance from each HDB resale flat to the nearest CBD, eldercare centre, hawker centre, MRT station, park, good primary school, shopping mall, supermarket, and clinic using spatial nearest-neighbour analysis.

```{r}
# CBD
dis_cbd <- st_nearest_feature(HDBresale, cbd)
HDBresale$dist_cbd <- as.numeric(st_distance(HDBresale, cbd[dis_cbd, ], by_element = TRUE))

# Eldercare
dis_eldercare <- st_nearest_feature(HDBresale, eldercare)
HDBresale$dist_eldercare <- as.numeric(st_distance(HDBresale, eldercare[dis_eldercare, ], by_element = TRUE))

# Foodcourt/hawker centres
dis_hawker <- st_nearest_feature(HDBresale, hawker)
HDBresale$dist_hawker <- as.numeric(st_distance(HDBresale, hawker[dis_hawker, ], by_element = TRUE))

# MRT
dis_mrt <- st_nearest_feature(HDBresale, mrt_station)
HDBresale$dist_mrt <- as.numeric(st_distance(HDBresale, mrt_station[dis_mrt, ], by_element = TRUE))

# Park
dis_parks <- st_nearest_feature(HDBresale, parks)
HDBresale$dist_parks <- as.numeric(st_distance(HDBresale, parks[dis_parks, ], by_element = TRUE))

# Good primary school
dis_goodps <- st_nearest_feature(HDBresale, good_primary_schools)
HDBresale$dist_goodps <- as.numeric(st_distance(HDBresale, good_primary_schools[dis_goodps, ], by_element = TRUE))

# Shopping mall
dis_shopping <- st_nearest_feature(HDBresale, shopping)
HDBresale$dist_shopping <- as.numeric(st_distance(HDBresale, shopping[dis_shopping, ], by_element = TRUE))

# Supermarket
dis_supermarkets <- st_nearest_feature(HDBresale, supermarkets)
HDBresale$dist_supermarkets <- as.numeric(st_distance(HDBresale, supermarkets[dis_supermarkets, ], by_element = TRUE))

# Clinics
dis_clinics <- st_nearest_feature(HDBresale, clinics)
HDBresale$dist_clinics <- as.numeric(st_distance(HDBresale, clinics[dis_clinics, ], by_element = TRUE))
```

### Facility Count within distance

This code counts the number of nearby facilities (e.g., kindergartens, childcare centres, bus stops, preschools, primary schools, other schools) within 350 m or 1 km of each HDB flat and adds these counts to the HDB resale dataset.

```{r}
within_350m_kindergartens <- st_is_within_distance(HDBresale, kindergartens, dist = 350)
within_350m_childcare <- st_is_within_distance(HDBresale, childcare, dist = 350)
within_350m_bus <- st_is_within_distance(HDBresale, bus_stop, dist = 350)
within_350m_preschool <- st_is_within_distance(HDBresale, preschool, dist = 350)
within_1km_schools <- st_is_within_distance(HDBresale, schools, dist = 1000)
within_1km_primary <- st_is_within_distance(HDBresale, primary_schools, dist = 1000)

HDBresale <- HDBresale %>%
  mutate(
    n_kindergartens_350m = map_int(within_350m_kindergartens, length),
    n_childcare_350m = map_int(within_350m_childcare, length),
    n_bus_350m = map_int(within_350m_bus, length),
    n_preschool_350m = map_int(within_350m_preschool, length))

HDBresale <- HDBresale %>%
  mutate(
    n_primary_1km = map_int(within_1km_primary, length),
    n_schools_1km = map_int(within_1km_schools, length))
```

# 4 Exploratory Data Analysis

## Distribution of Resale price

```{r}
ggplot(HDBresale, aes(x=resale_price)) +
  geom_histogram(bins = 20, color = "black", fill = "lightblue")
```

From the histogram above, we can see that the distribution of HDB resale prices is right-skewed, with most transactions concentrated in the lower to mid-price range. The majority of resale flats are sold between approximately \$600,000 and \$800,000, indicating this is the typical market value for most units. As prices increase, the number of transactions declines steadily, reflecting fewer high-priced sales. Nonetheless, there are notable outliers, with some flats being transacted above \$1 million and even reaching beyond \$1.5 million. These high-value units likely represent larger flats, prime locations, or recently renovated units. Overall, the HDB resale market shows a dominant clustering around mid-range prices, with a gradual tapering towards premium-priced properties.

```{r}
tmap_mode("plot")
tmap_style("white")

tm_shape(mpsz)+
  tm_polygons(col = "white") +
tm_shape(HDBresale) +  
  tm_dots(col = "resale_price",
          alpha = 0.6,
          style="quantile",
          size = 0.1,
          palette = "Blues") +
  tm_layout(legend.position = c("right", "bottom"))
```

The map illustrates the spatial distribution of HDB resale prices across Singapore, with darker shades of blue representing higher-priced flats and lighter shades indicating lower-priced ones. From the visualization, it is evident that higher resale prices are concentrated in central and mature estates such as Queenstown, Bukit Merah, Kallang/Whampoa, Bishan, and parts of Bukit Timah—areas that are closer to the city centre, well-connected by transport, and known for limited housing supply. In contrast, lighter blue areas, indicating lower-priced flats, are more common in non-mature or peripheral estates such as Sembawang, Yishun, Woodlands, Jurong West, and Punggol. These areas are typically farther from the Central Business District (CBD) and have more abundant supply of newer flats. Overall, the map highlights a clear spatial pattern—HDB resale prices tend to be higher in centrally located and mature towns, and lower in newer or peripheral regions of Singapore.

## Distribution of Locational factors

```{r}
location_factors <- c(
  "dist_cbd", "dist_eldercare", "dist_hawker", "dist_mrt", "dist_parks",
  "dist_goodps", "dist_shopping", "dist_supermarkets", "dist_clinics",
  "n_kindergartens_350m", "n_childcare_350m", "n_bus_350m",
  "n_preschool_350m", "n_primary_1km", "n_schools_1km"
)

plots <- lapply(location_factors, function(v) {
  ggplot(HDBresale, aes_string(x = v)) +
    geom_histogram(bins = 20, color = "black", fill = "lightblue") +
    labs(title = v, x = NULL, y = NULL) +
    theme_minimal() +
    theme(
      plot.title = element_text(size = 9, hjust = 0.5),
      axis.text = element_text(size = 6)
    )
})

EDA_Location <- wrap_plots(plots, ncol = 3) +
  plot_annotation(title = "EDA of Location & Accessibility Factors")

EDA_Location
```

The histograms provide an overview of how HDB resale flats are spatially distributed with respect to various amenities and accessibility factors. Most flats are located between 10 km to 18 km from the Central Business District (CBD), indicating that the dataset includes many non-central residential areas. For amenities such as MRT stations, hawker centres, supermarkets, parks, clinics, and eldercare centres, the majority of flats are situated within 500–1,000 meters, showing generally good accessibility for daily needs. Distances to good primary schools and shopping malls show wider variation, with some flats located much further away. In terms of facility counts within a walkable 350-meter radius, most flats have only 0–1 kindergartens or preschools nearby, but typically 3–8 childcare centres and 5–12 bus stops. Within a 1 km radius, most flats have 2–5 primary schools and 4–8 schools overall.

## Distribution of Structural factors

```{r}
# Create individual plots for structural factors
UNIT_AREA <- ggplot(data = HDBresale, aes(x = floor_area_sqm)) +
  geom_histogram(bins = 20, color = "black", fill = "lightblue")

UNIT_FLOOR_LEVEL <- ggplot(data = HDBresale, aes(x = floor_level)) +
  geom_histogram(bins = 20, color = "black", fill = "lightblue")

UNIT_REMAINING_LEASE <- ggplot(data = HDBresale, aes(x = remaining_lease)) +
  geom_histogram(bins = 20, color = "black", fill = "lightblue")

UNIT_AGE <- ggplot(data = HDBresale, aes(x = unit_age)) +
  geom_histogram(bins = 20, color = "black", fill = "lightblue")

# Arrange plots in a 2x2 grid
ggarrange(UNIT_AREA, UNIT_FLOOR_LEVEL, UNIT_REMAINING_LEASE, UNIT_AGE,
          ncol = 2, nrow = 2)
```

The histograms provide an overview of the physical characteristics of HDB resale flats. Most units have a floor area between 110 and 130 square meters, with a sharp concentration indicating the common sizes of 5-room flats or executive apartments. Floor levels are right-skewed, with the majority of flats located between the 1st and 12th floors, and relatively few units situated above the 20th floor. The remaining lease distribution is fairly symmetrical, mostly ranging from 60 to 90 years, showing that many of the flats were built between the late 1990s and early 2000s. Correspondingly, unit age displays a roughly mirrored pattern of remaining lease, with most flats being between 20 and 40 years old.

# 5 Multicollinearity Analysis

The code below analyzes the dataset for multicollinearity, where predictor variables are highly correlated.

```{r fig.height=10, fig.width=12}
hdb_resale_nogeo <- HDBresale %>%
  st_drop_geometry()%>%
  select(
    -resale_price
  )
ggcorrmat(hdb_resale_nogeo, names = hdb_resale_nogeo)
```

There is a perfect negative correlation of -1.0 between unit_age and remaining_lease, which represents a critical multicollinearity issue. This perfect correlation is expected since these variables measure essentially the same characteristic from opposite perspectives. Thus, unit_age will be removed.

The analysis also identifies a strong positive correlation of 0.88 between n_schools_1km and n_primary_1km, indicating that areas with many schools within 1km also tend to have many primary schools in the same radius. This high correlation suggests that these variables capture overlapping information about the educational facility density in the neighborhood. Thus, n_schools_1km will be dropped as primary school distance is more important for parents to choose houses. Similarly, n_preschool_350m and n_childcare_350m show strong positive correlation of 0.97, suggesting that they tend to cluster together. Thus, n_childcare_350m will be dropped.

```{r fig.height=10, fig.width=12}
re_hdb_nogeo <- HDBresale %>%
  st_drop_geometry() %>%
  select(
    -n_schools_1km,
    -unit_age,
    -n_childcare_350m,
    -resale_price
  )

ggcorrmat(re_hdb_nogeo, names = re_hdb_nogeo)
```

The refined dataset now exhibits substantially improved characteristics for regression modeling, with multicollinearity concerns largely resolved. The correlation values are predominantly in the low to moderate range, with the vast majority falling below 0.6, which will facilitate stable coefficient estimation and reduce variance inflation in subsequent hedonic pricing models.

# 6 Multiple Linear Regression

## Prepare Train and Test Data

The code below splits the HDB resale dataset into training and testing subsets based on transaction dates — using data from July 2024 to June 2025 for model training and July 2025 to September 2025 for model testing.

```{r}
# Training data: July 2024 - June 2025
hdb_resale_train <- HDBresale %>%
  filter((str_sub(month, 1, 7) >= "2024-07") & (str_sub(month, 1, 7) <= "2025-06"))

# Test data: July 2025 - September 2025
hdb_resale_test <- HDBresale %>%
  filter((str_sub(month, 1, 7) >= "2025-07") & (str_sub(month, 1, 7) <= "2025-09"))
```

## Calibrate Multiple Linear Regression Model

The code below builds a multiple linear regression model to predict HDB resale prices using distance to CBD (dist_cbd), distance to the nearest eldercare centre (dist_eldercare), distance to hawker/food centres (dist_hawker), distance to MRT station (dist_mrt), distance to parks (dist_parks), distance to good primary schools (dist_goodps), distance to shopping malls (dist_shopping), distance to supermarkets (dist_supermarkets), distance to clinics (dist_clinics), number of kindergartens within 350 m (n_kindergartens_350m), number of bus stops within 350 m (n_bus_350m), number of preschools within 350 m (n_preschool_350m), number of schools within 1 km (n_schools_1km), floor area in square meters (floor_area_sqm), floor level of the unit (floor_level), remaining lease in years (remaining_lease).

```{r}
resale_mlr <- lm(
  formula = resale_price ~ 
    dist_cbd +
    dist_eldercare +
    dist_hawker +
    dist_mrt +
    dist_parks +
    dist_goodps +
    dist_shopping +
    dist_supermarkets +
    dist_clinics +
    n_kindergartens_350m +
    n_bus_350m +
    n_preschool_350m +
    n_primary_1km +
    floor_area_sqm +
    floor_level +
    remaining_lease,
  data = hdb_resale_train
)

ols_regress(resale_mlr)
```

The final multiple linear regression (MLR) model successfully explains variation in HDB resale prices, with all 16 variables now statistically significant (p \< 0.05). The model achieves an R-squared of 0.730 and an adjusted R-squared of 0.729, indicating strong explanatory power while maintaining model parsimony. The ANOVA F-statistic of 1056.217 with a p-value \< 0.0001 confirms the model’s overall significance.

Among structural factors, floor area emerges as the most influential predictor, with each additional square meter contributing \$7,828 to resale value. Remaining lease adds \$7,240 per year, and floor level contributes \$6,589 per floor, confirming buyers’ preference for larger, newer, and higher-floor units. Locational attributes also play a critical role. Proximity to hawker centres shows the strongest impact, with each additional meter reducing value by \$45.57. Distance to MRT stations (-\$41.56) and the CBD (-\$25.94) further highlight the importance of accessibility to daily necessities, transit, and employment. Additional negative effects are observed for distance to parks, shopping centres, and top primary schools. In terms of facility density, the number of kindergartens within 350 meters contributes the highest positive effect (\$18,147), indicating strong appeal to young families. However, other count-based variables such as nearby preschools and primary schools show negative effects, suggesting potential oversupply or congestion.

Overall, the model captures key structural and locational determinants of price, offering valuable insights for urban planning, valuation, and household decision-making.

## Multicollinearity Check

This code checks for multicollinearity in the regression model by calculating the Variance Inflation Factor (VIF) and tolerance values for each predictor.

```{r}
ols_vif_tol(resale_mlr)
```

The Variance Inflation Factor (VIF) values in the output are all below 5, which suggests that multicollinearity is not a significant issue in the model. Each predictor variable has a relatively low VIF, indicating minimal correlation with the other predictors. This means the model's coefficients are likely to be stable and reliable, with each variable contributing independently to the prediction of the resale price.

## Test for Spatial Autocorrelation

The code below maps the residuals of the multiple linear regression model onto Singapore’s spatial map to visually check for spatial patterns or clustering in prediction errors.

```{r}
mlr_res <- as.data.frame(resale_mlr$residuals)

resale_res <- cbind(hdb_resale_train, mlr_res) %>%
  rename(MLR_RES = resale_mlr.residuals)

tm_shape(mpsz) +
  tm_polygons(col = "white") +
tm_shape(resale_res) +
  tm_dots(col = "MLR_RES",
          alpha = 0.6,
          style = "quantile",
          size = 0.1) +  # Adjust dot size (smaller)
  tm_layout(main.title = "Multiple Linear Regression Residuals",
            main.title.position = "center",
            main.title.size = 1,
            legend.position = c("right", "bottom"))  # Position legend at bottom right
```

The map shows spatial patterns in the model’s residuals, suggesting potential spatial autocorrelation.

To statistically confirm our observation of spatial autocorrelation, we’ll perform Moran’s I Test.

H₀: The regression residuals are randomly distributed in space (no spatial autocorrelation).

H₁: The regression residuals exhibit positive spatial autocorrelation.

```{r}
resale_sp <- as_Spatial(resale_res)

nb <- dnearneigh(coordinates(resale_sp), 0, 1500, longlat = FALSE)
nb_lw <- nb2listw(nb, style = 'W')

lm.morantest(resale_mlr, nb_lw)
```

The Global Moran's I test for residual spatial autocorrelation shows a p-value far below the alpha value of 0.05, so we reject the null hypothesis that the residuals are randomly distributed across space. The observed Moran's I statistic is approximately 0.322, which is substantially greater than the expected value of approximately -0.001 under spatial randomness. Since the observed Global Moran's I is positive and significantly greater than zero, we can infer that the residuals exhibit a spatially clustered distribution, with similar residual values (either positive or negative) tending to occur near each other. This indicates the presence of significant positive spatial autocorrelation.

# 7 Geographically Weighted Random Forest

## Data Preparation

The code below extracts spatial coordinates from the training dataset and removes the geometry column to prepare the data for use in the SpatialML package.

```{r}
coords_train <- st_coordinates(hdb_resale_train)

hdb_resale_train_nogeo <- hdb_resale_train %>% 
  st_drop_geometry()
```

## Compute Optimum Bandwidth

The code below computes the optimum adaptive bandwidth using cross-validation to achieve the best local model performance.Since both GWR and GWRF are geographically weighted methods that define local neighborhoods through bandwidth selection, the cross-validation approach in bw.gwr() provides a computationally efficient and statistically sound starting point for GWRF.

```{r}
#| eval: false
#| output: false
bw_adaptive <- bw.gwr(
  formula = resale_price ~ 
    dist_cbd +
    dist_eldercare +
    dist_hawker +
    dist_mrt +
    dist_parks +
    dist_goodps +
    dist_shopping +
    dist_supermarkets +
    n_kindergartens_350m +
    n_bus_350m +
    n_preschool_350m +
    n_primary_1km +
    floor_area_sqm +
    floor_level +
    remaining_lease,
  data = resale_res,
  approach = "CV",
  kernel = "gaussian",
  adaptive = TRUE,
  longlat = FALSE
)
```

```{r}
#| eval: false
write_rds(bw_adaptive, "data/model/bw_adaptive.rds")
```

```{r}
#| echo: false
bw_adaptive <- read_rds("data/model/bw_adaptive.rds")
```

```{r}
bw_adaptive
```

The optimal bandwidth for this dataset is 53.

## Calibrate Geographically Weighted Random Forest Model

The code below calibrates a Geographically Weighted Random Forest model to predict HDB resale prices using the following predictors: distance to CBD (dist_cbd), distance to the nearest eldercare centre (dist_eldercare), distance to hawker/food centres (dist_hawker), distance to MRT station (dist_mrt), distance to parks (dist_parks), distance to good primary schools (dist_goodps), distance to shopping malls (dist_shopping), distance to supermarkets (dist_supermarkets), distance to clinics (dist_clinics), number of kindergartens within 350 m (n_kindergartens_350m), number of bus stops within 350 m (n_bus_350m), number of preschools within 350 m (n_preschool_350m), number of schools within 1 km (n_schools_1km), floor area in square meters (floor_area_sqm), floor level of the unit (floor_level), and remaining lease in years (remaining_lease). The model incorporates spatial coordinates and uses the optimal adaptive bandwidth (bw_adaptive) to account for spatial non-stationarity in feature importance and predictions.

```{r}
#| eval: false
#| output: false
set.seed(1234)

gwRF_adaptive <- grf(
  formula = resale_price ~
    dist_cbd +
    dist_eldercare +
    dist_hawker +
    dist_mrt +
    dist_parks +
    dist_goodps +
    dist_shopping +
    dist_supermarkets +
    n_kindergartens_350m +
    n_bus_350m +
    n_preschool_350m +
    n_primary_1km +
    floor_area_sqm +
    floor_level +
    remaining_lease,
  dframe = hdb_resale_train_nogeo,
  bw = bw_adaptive,
  ntree = 300,
  kernel = "adaptive",
  verbose = TRUE,
  coords = coords_train
)

write_rds(gwRF_adaptive, "data/model/gwRF_adaptive.rds")
```

```{r}
#| echo: false
gwRF_adaptive <- read_rds("data/rds/gwRF_adaptive.rds")
```

The code below checks the importance of parameters from the geographically weighted random forest model.

```{r}
variable_importance <- gwRF_adaptive$Global.Model$variable.importance
sort(variable_importance, decreasing = TRUE)
```

The variable importance analysis from the Geographically Weighted Random Forest (GWRF) model highlights the relative contribution of each predictor to model performance. Distance to CBD (dist_cbd) remains the most influential feature with an importance score of 9.79×10¹³, reaffirming the premium placed on centrality in HDB resale pricing. This is followed by remaining lease in months (remaining_lease_mths) at 7.02×10¹³, confirming the strong influence of lease duration on valuation.

The next tier of important predictors includes storey_mean (2.85×10¹³), distance to good primary schools (dist_good_primary, 2.34×10¹³), floor area (floor_area_sqm, 1.08×10¹³), and distance to MRT stations (dist_mrt, 9.92×10¹²). These variables emphasize the value buyers place on higher floor levels, educational access, larger units, and connectivity to transit.

Moderate importance is observed for other proximity-based variables such as distance to eldercare (8.29×10¹²), hawker centres (6.83×10¹²), parks (6.57×10¹²), shopping malls (5.46×10¹²), and supermarkets (3.37×10¹²), suggesting that accessibility to lifestyle and care amenities significantly contributes to perceived value.

At the lower end are facility density variables, including number of kindergartens, childcare centres, bus stops within 350m, and schools within 1km, with scores ranging from 1.26×10¹² to 3.37×10¹². These results suggest that proximity to amenities has a stronger influence on pricing than the sheer number of nearby facilities, aligning with previous MLR findings and underscoring the role of both spatial location and structural features in shaping resale value.

# 8 Geographically Weighted Regression

## Calibrate Geographically Weighted Regression Model

The code below fits an adaptive-bandwidth Geographically Weighted Regression (GWR) model for HDB resale prices using spatial predictors, then saves the fitted model as an RDS file for later use.

```{r}
#| eval: false
#| output: false
gwr_adaptive <- gwr.basic(
  formula = resale_price ~ 
    dist_cbd +
    dist_eldercare +
    dist_hawker +
    dist_mrt +
    dist_parks +
    dist_goodps +
    dist_shopping +
    dist_supermarkets +
    n_kindergartens_350m +
    n_bus_350m +
    n_preschool_350m +
    n_primary_1km +
    floor_area_sqm +
    floor_level +
    remaining_lease,
  data    = resale_sp,
  bw      = bw_adaptive, 
  kernel  = "gaussian",
  adaptive = TRUE,
  longlat = FALSE
)

write_rds(gwr_adaptive, "data/model/gwr_adaptive.rds")
```

```{r}
#| echo: false
gwr_adaptive <- read_rds("data/model/gwr_adaptive.rds")
```

```{r}
gwr_adaptive
resale_sp$gwr_pred  <- gwr_adaptive$SDF$pred     
resale_sp$gwr_resid <- gwr_adaptive$SDF$gwr.e  
```

The Geographically Weighted Regression (GWR) model delivers a substantial improvement over the global regression, achieving an R-squared of 0.941, compared to 0.730 from the global model. This represents an additional 21% of explained variation, confirming the presence of strong spatial non-stationarity in the determinants of HDB resale prices. In other words, the influence of predictors such as accessibility, amenities, and physical attributes varies meaningfully across different neighbourhoods in Singapore.

The model uses an adaptive bandwidth of 53 nearest neighbours with a Gaussian kernel. This bandwidth is large enough to ensure stable estimation of the 15 predictors, while remaining small enough to preserve local specificity. Such a bandwidth is consistent with recommendations in GWR literature for datasets of this size.

The GWR coefficient summary reveals notable spatial variation in predictor effects. For example, the coefficient for distance to the CBD ranges from approximately –72 to –86, indicating that the price penalty for being farther from the CBD differs across locations. Similarly, the floor area premium ranges from –14,550 to +70,720, reflecting how the value of larger flats varies between mature estates and newer towns. These variations illustrate how GWR successfully captures geographically varying housing market dynamics that a single global model cannot represent.

Model diagnostics highlight the complexity inherent in GWR. The effective number of parameters is approximately 878, far larger than the 15 parameters of the global model, resulting in a corresponding reduction of effective degrees of freedom to 5386. Despite this complexity, the adjusted R-squared remains high at 0.931, indicating that the improved fit is not merely due to overfitting. The AICc value of 150,292 provides an appropriate complexity-penalized benchmark for comparing GWR against alternative spatial models.

# 9 Model Evaluation

We will evaluate the models we calibrated earlier (Multiple Linear Regression and Geographically Weighted Random Forest).

## Prepare Test Data

The code below extracts the spatial coordinates and removes geometry from the HDB resale test dataset to prepare it for prediction using the GWRF model.

```{r}
coords_test <- st_coordinates(hdb_resale_test)

hdb_resale_test_nogeo <- hdb_resale_test %>%
  st_drop_geometry()
```

## Model Predictions on Test Data

The code below uses the Multiple Linear Regression model to make predictions on the test data.

```{r}
hdb_resale_test$PREDICT_MLR <- predict(object = resale_mlr, newdata = hdb_resale_test)
```

The code below uses the Geographically Weighted Random Forest model to make predictions.

```{r}
#| eval: false
#| output: false
test_data_for_prediction <- hdb_resale_test_nogeo %>%
  rename(
    storey_mean = floor_level,
    remaining_lease_mths = remaining_lease,
    dist_park = dist_parks,
    dist_good_primary = dist_goodps,
    dist_supermarket = dist_supermarkets,
    dist_mall = dist_shopping
  )

test_data_for_prediction$X <- coords_test[, 1]
test_data_for_prediction$Y <- coords_test[, 2]

test_data_for_prediction <- as.data.frame(test_data_for_prediction)

predictions_gwrf <- predict.grf(
  gwRF_adaptive,
  test_data_for_prediction,
  x.var.name = "X",
  y.var.name = "Y",
  local.w = 1,
  global.w = 0
)

hdb_resale_test$PREDICT_GWRF <- predictions_gwrf
```

```{r}
#| eval: false
write_rds(hdb_resale_test, "data/rds/hdb_resale_test.rds")
```

```{r}
#| echo: false
hdb_resale_test <- read_rds("data/rds/hdb_resale_test.rds")
```

The code below uses the Geographically Weighted Linear Regression to make predictions.

```{r}
#| eval: false
#| output: false
coords_test <- st_coordinates(hdb_resale_test)
resale_sp_test <- as_Spatial(hdb_resale_test)

predictions_gwr <- gwr.predict(
  formula = resale_price ~ 
    dist_cbd +
    dist_eldercare +
    dist_hawker +
    dist_mrt +
    dist_parks +
    dist_goodps +
    dist_shopping +
    dist_supermarkets +
    n_kindergartens_350m +
    n_bus_350m +
    n_preschool_350m +
    n_primary_1km +
    floor_area_sqm +
    floor_level +
    remaining_lease,
  data = resale_sp_test,
  predictdata = resale_sp_test,
  bw = bw_adaptive,
  kernel = "gaussian",
  adaptive = TRUE,
  longlat = FALSE
)

hdb_resale_test$PREDICT_GWR <- predictions_gwr$SDF$prediction


write_rds(predictions_gwr, "data/rds/predictions_gwr.rds")

write_rds(hdb_resale_test, "data/rds/hdb_resale_test.rds")
```

```{r}
#| echo: false
predictions_gwr <- read_rds("data/rds/predictions_gwr.rds")
hdb_resale_test <- read_rds("data/rds/hdb_resale_test.rds")
```

## Calculate Root Mean Square Error (RMSE)

The code below calculates the Root Mean Square Error (RMSE) of the Multiple Linear Regression and Geographically Weighted Random Forest models.

```{r}
MLR_RMSE <- rmse(hdb_resale_test$resale_price, hdb_resale_test$PREDICT_MLR)
gwRF_RMSE <- rmse(hdb_resale_test$resale_price, hdb_resale_test$PREDICT_GWRF)
GWR_RMSE  <- rmse(hdb_resale_test$resale_price, hdb_resale_test$PREDICT_GWR)

RMSE_results <- data.frame(
  Model = c("Multiple Linear Regression", "Geographically Weighted Random Forest", "Geographically Weighted Regression"),
  RMSE = c(MLR_RMSE, gwRF_RMSE, GWR_RMSE))

knitr::kable(RMSE_results, caption = "RMSE Comparison of Model Predictions")
```

The results indicate that Geographically Weighted Regression (GWR) achieves the lowest prediction error (RMSE = 46,448.19), outperforming both the Multiple Linear Regression model (RMSE = 90,545.65) and the Geographically Weighted Random Forest model (RMSE = 178,805.25). This suggests that incorporating spatially varying coefficients through GWR is highly effective for this dataset, capturing local price dynamics that global models cannot.

In contrast, the MLR model, despite being a simple global model, performs moderately well and substantially better than gwRF. The relatively strong performance of MLR implies that a significant portion of price variation can still be explained by global linear relationships without requiring complex spatial interactions.The gwRF model, however, performs the worst, indicating potential overfitting to local training patterns or insufficient spatial heterogeneity to justify its high model flexibility.

## Model Predictions Visualization

The code below visualizes the prediction performance of both models by plotting actual resale prices against predicted prices.

```{r fig.height = 8, fig.width=14}
plot_mlr <- ggplot(data = hdb_resale_test,
                   aes(x = PREDICT_MLR,
                       y = resale_price)) +
  geom_point() +
  ggtitle("Resale Price vs Predicted Price (MLR)") +
  theme(plot.title = element_text(size = 10))

plot_gwrf <- ggplot(data = hdb_resale_test,
                    aes(x = PREDICT_GWRF,
                        y = resale_price)) +
  geom_point() +
  ggtitle("Resale Price vs Predicted Price (GWRF)") +
  theme(plot.title = element_text(size = 10)) 


plot_gwr <- ggplot(data = hdb_resale_test,
                   aes(x = PREDICT_GWR,
                       y = resale_price)) +
  geom_point() + 
  ggtitle("Resale Price vs Predicted Price (GWR)") +
  theme(plot.title = element_text(size = 10))


ggarrange(plot_mlr, plot_gwrf, plot_gwr, ncol = 3)
```

The scatterplots reveal clear differences in predictive performance among the three models. GWR (right) exhibits the strongest alignment between predicted and actual resale prices, with points clustering most tightly along the diagonal across the entire price spectrum. This visual pattern corresponds to its lowest RMSE and indicates that GWR captures spatially varying price relationships effectively, yielding highly accurate and well-calibrated predictions.

The MLR model (left) also shows a strong linear relationship, with most points concentrated around the diagonal. However, compared to GWR, the spread is wider and deviations from the trend line become more noticeable at higher price levels (above approximately \$1,200,000). This suggests that while MLR performs reasonably well—especially in the dominant \$600,000–\$1,000,000 transaction range—it lacks the spatial flexibility to fully capture localized price variations.

In contrast, the gwRF model (middle) shows the greatest scatter and the weakest alignment with the diagonal, particularly in the upper price range. Several predictions deviate substantially from actual values, reflecting inconsistent performance and greater prediction error. This visual behavior aligns with gwRF's much higher RMSE and indicates probable overfitting to local patterns that do not generalize well to the test dataset.

## Visualization of Model Residuals

The code below calculates the residuals of best model(Geographically Weighted Regression), and visualizes them on a map.

```{r}
hdb_resale_test <- hdb_resale_test %>%
  mutate(GWR_RES = resale_price - PREDICT_GWR)

common_breaks <- c(-Inf, seq(-400000, 400000, by = 100000), Inf)

base_map <- tm_shape(mpsz) +
  tm_polygons(col = "white", border.col = "gray70", border.alpha = 0.5)

gwrf_res_map <- base_map +
  tm_shape(hdb_resale_test) +
  tm_dots(
    col = "GWR_RES",
    breaks = common_breaks,
    palette = "-RdBu",
    size = 0.25,
    alpha = 0.7,
    border.col = "black",
    border.lwd = 0.5,
    border.alpha = 0.3,
    title = "GWR Residuals ($)") +
  tm_layout(
    main.title = "Geographically Weighted Regression Residuals",
    main.title.size = 1,
    legend.outside = TRUE,
    legend.outside.position = "right")

tmap_mode("plot")

gwrf_res_map
```

The spatial distribution of GWR residuals illustrates how well the model captures local housing price dynamics across Singapore. Red indicates over-prediction (model predicts higher than actual price), blue indicates under-prediction, and lighter tones represent predictions close to the true values.

Overall, the GWR residual map shows mostly light-colored points, indicating that prediction errors are generally small and well-controlled across the island—consistent with GWR’s strong RMSE performance. The residuals do not form large, contiguous clusters, suggesting that the model successfully accounts for much of the spatial heterogeneity in resale prices.

However, some localized patterns remain. Mild under-predictions (light blue) appear in parts of the northeast and eastern regions, suggesting that actual prices in those neighborhoods are slightly higher than what the model expects. Conversely, small pockets of over-prediction (light to medium red) are visible in selected central and northwestern subzones, possibly reflecting micro-market premiums or recent price movements not fully captured by the training data.

Notably, the map lacks the large, concentrated errors seen in more rigid global models or overly complex spatial models. GWR avoids extreme residuals (very deep blue or red), demonstrating that its spatially varying coefficients adapt well to neighborhood-level market conditions. The absence of strong spatial clustering of errors further confirms that GWR provides stable, geographically consistent prediction performance, reinforcing why it achieved the lowest RMSE among the three models.

# 9 Conclusion

The analysis reveals a distinctly right-skewed distribution of 5-room HDB resale prices and a clear spatial gradient: most transactions cluster between \$600,000 and \$800,000, with high-value sales above \$1 million concentrated in central and mature towns such as Queenstown, Bukit Merah, Kallang/Whampoa and Bishan, while lower prices dominate in peripheral estates like Sembawang, Yishun, Woodlands and Punggol. Across all models, the same core value drivers emerge. Structurally, larger floor area, longer remaining lease and higher floor level consistently command substantial premiums. Locationally, proximity to the CBD, MRT stations, hawker centres, parks, shopping malls and top primary schools exhibits strong capitalization effects, confirming that accessibility to employment, transit and amenities is deeply embedded in HDB pricing.

Out-of-sample evaluation shows a clear ranking among the three modelling approaches. Geographically Weighted Regression (GWR) achieves the lowest prediction error (RMSE ≈ \$46,448), substantially outperforming the global Multiple Linear Regression (RMSE ≈ \$90,546) and the Geographically Weighted Random Forest (RMSE ≈ \$178,805). Scatterplots of predicted versus actual prices and the GWR residual map confirm this: GWR predictions align most closely with the 45-degree line and residuals are generally small, with limited spatial clustering. In contrast, while the parsimonious MLR still performs reasonably well and demonstrates that a large share of price variation can be captured by global linear relationships, its residuals exhibit significant positive spatial autocorrelation, indicating unmodeled spatial processes. The GWRF model performs worst, with highly scattered predictions and pockets of large errors, suggesting overfitting to local training patterns without corresponding gains in generalization.

GWR diagnostics further highlight meaningful spatial non-stationarity. The model attains an R² of about 0.94 (adjusted R² ≈ 0.93), and local coefficients show substantial geographic variation—for example, the price penalty of distance to the CBD and the premium for additional floor area differ markedly across neighbourhoods, reflecting heterogeneous sub-markets between mature central estates and newer peripheral towns. Although this flexibility comes at the cost of a high effective parameter count and reduced degrees of freedom, the strong adjusted R² and improved residual behaviour suggest that the additional complexity captures genuine spatial structure rather than mere noise.

This study remains constrained by a short evaluation window, reliance on Euclidean rather than network distances, potential geocoding inaccuracies, and limited treatment of temporal effects such as policy changes or market shocks. Future work could extend the time horizon, incorporate network-based accessibility measures, explore spatio-temporal or multilevel models, and refine spatial machine-learning approaches. Nonetheless, the results demonstrate that while a well-specified global hedonic MLR provides a solid baseline, Geographically Weighted Regression offers the best balance of interpretability and predictive accuracy by explicitly modelling the spatially varying nature of Singapore’s HDB resale market.
