---
title: "Take home exercise 2"
author: "Tai Qiu Yan"
date: "October 8, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# 1 Introduction

# 2 Loading the R package

In this exercise, the following packages are used:

tidyverse – A collection of data science packages (readr, dplyr, tidyr, ggplot2, etc.) with a unified grammar for importing, wrangling, and visualising data.

sf – Simple Features for R: create, read, write, transform, and analyse vector geospatial data (points/lines/polygons) with modern CRS support.

sfdep – Provides functions for spatial dependence analysis, including tools for computing spatial weights, spatial autocorrelation measures, and emerging hotspot analysis, extending the functionality of the sf package.

tmap – A flexible thematic mapping package that supports both static and interactive maps.

knitr – An engine for dynamic report generation, allowing integration of R code and results into documents.

kableExtra – Extends knitr::kable() by providing advanced table styling features. It allows users to create publication-quality tables in HTML and LaTeX.

DT – An R interface to the DataTables JavaScript library, enabling interactive tables with features such as searching, filtering, sorting, and pagination.

Kendall – Implements the Mann-Kendall trend test and related functions for detecting monotonic trends in time series data.

```{r}
pacman::p_load(tidyverse, sf, sfdep, tmap, knitr, kableExtra, DT, Kendall, scales)
```

# 3 Data Importing and Wrangling

## Importing Data

The following datasets are imported:

- Bus Stop Location shapefile downloaded from LTA DataMall

- Master Plan 2019 Subzone Boundary (No Sea) from Singapore’s open data portal

- Passenger Volume by Origin Destination Bus Stops downloaded from LTA DataMall

```{r}
#Bus Stop Location
BusStop = st_read(dsn = "data/LTADataMall/", 
                  layer = "BusStop") %>%
  st_transform(crs = 3414)

#Master Plan 2019 Subzone
mpsz = read_rds("data/rds/mpsz_sf.rds")

#O-D Bus Trips
odbus <- read_csv("data/LTADataMall/origin_destination_bus_202508.csv")
```

## Extracting Bus Stops located within Singapore

The code below assigns each bus stop to the planning subzone it belongs to, creating a dataset of bus stops (BusStop_in_SG) linked to their respective subzones.

```{r}
BusStop_in_SG <- st_join(
  BusStop, mpsz, 
  join = st_within, 
  left = FALSE)
```

### Visualize

```{r}
tm_shape(mpsz) +
  tm_polygons() +
tm_shape(BusStop_in_SG) +
  tm_dots()
```

## Deriving Analytical Hexagon

The code below generates a hexagonal tessellation over Singapore’s planning subzones to standardize spatial units for analysis.

```{r}
hexagon <- st_make_grid(mpsz, 
                        cellsize = 750,
                        what = "polygon",
                        square = FALSE) %>%
  st_sf()
```

## Selecting hexagons with bus stops

The code below links bus stops to the analytical hexagons, counts how many fall inside each, and filters the grid to include only meaningful hexagons with bus stops.

```{r}
hexagon$busstop_count =
  lengths(st_intersects(
    hexagon, BusStop_in_SG))

hexagon <- filter(
  hexagon, busstop_count > 0)
```

## Assigning ids to each hexagon

The code below cleans the hexagon dataset by dropping the temporary count column and assigns each hexagon a unique identifier (HEX_ID) for consistent reference in spatial analysis.

```{r}
hexagon <- hexagon %>% 
  select(, -busstop_count)

hexagon$HEX_ID <- sprintf(
  "H%04d", seq_len(
    nrow(hexagon))) %>% 
  as.factor()
```

## Preparing Trip Generation Data

The code below extracts the relevant bus trip information, standardizes the column names into intuitive labels (BUS_STOP_N, HOUR_OF_DAY, TRIPS), and prepares the bus stop IDs as categorical values for later aggregation and spatial analysis.

```{r}
trips <- odbus %>%
  select(c(ORIGIN_PT_CODE, DAY_TYPE, TIME_PER_HOUR, TOTAL_TRIPS)) %>%
  rename(BUS_STOP_N = ORIGIN_PT_CODE) %>%
  rename(HOUR_OF_DAY = TIME_PER_HOUR) %>%
  rename(TRIPS = TOTAL_TRIPS)
trips$BUS_STOP_N <- as.factor(trips$BUS_STOP_N)
```

## Populating Hexagon IDs into BusStop data frame

The code below creates a lookup table that links each bus stop to the hexagon it belongs to. It is essential for aggregating bus trips at the hexagon level later on.

```{r}
bs_hex <- st_intersection(
  BusStop, hexagon) %>%
  st_drop_geometry() %>%
  select(c(BUS_STOP_N, HEX_ID))
```

## Adding HEX_ID into bus trips data

The code below enriches the trip dataset by adding hexagon IDs to each bus trip record, enabling aggregation and spatial analysis of trips at the hexagon level.

```{r}
trips <- inner_join(trips, bs_hex)
```

## Aggregating TRIPS based on HEX_ID

The code below converts raw trip records into spatial-temporal summaries, making the data easier to analyze by hexagon location, type of day, and hour of day.

```{r}
trips <- trips %>%
  group_by(
    HEX_ID,
    DAY_TYPE,
    HOUR_OF_DAY) %>%
  summarise(TRIPS = sum(TRIPS))
```


# 4 Weekday morning peak

## 4.1 Overview

### Choropleth map of total trips

```{r}
weekday_morning <- trips %>%
  filter(DAY_TYPE == "WEEKDAY", HOUR_OF_DAY %in% 6:9) %>%
  group_by(HEX_ID) %>%
  summarise(TRIPS = sum(TRIPS), .groups = "drop") %>%
  left_join(hexagon, by = "HEX_ID") %>%
  st_as_sf()

tmap_mode("plot")

tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  
  tm_shape(weekday_morning) +
  tm_polygons(
    fill = "TRIPS",
    fill.scale = tm_scale_intervals(
      style = "quantile",  # Use style = "quantile" instead
      n = 5,
      values = "viridis"
    ),
    fill.legend = tm_legend(
      title = "Total Trips\n(Weekday Morning)",
      position = tm_pos_in("right", "bottom")
    )
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_title("Distribution of Bus Trips - Weekday Morning Peak (6am-9am)") +
  tm_compass(position = tm_pos_in("left", "top")) +
  tm_scale_bar(position = tm_pos_in("left", "bottom"))
```

The visualisation shows that bus trips during the weekday morning peak (6–9am) are heavily concentrated in large residential towns such as Tampines, Pasir Ris, Yishun, and Woodlands, as well as the western corridor around Jurong East and Clementi. Moderate activity is seen in areas like Hougang, Sengkang, and Punggol, while peripheral industrial and coastal zones record low trip counts. Overall, the map highlights a polycentric commuting pattern, where dense housing estates generate significant trips feeding into central employment areas, reinforcing the key role of buses in connecting residential towns with work and study destinations.

### Trip Distribution

```{r}
mean_trips <- mean(weekday_morning$TRIPS)
median_trips <- median(weekday_morning$TRIPS)
q75 <- quantile(weekday_morning$TRIPS, 0.75)
q95 <- quantile(weekday_morning$TRIPS, 0.95)

# Create improved histogram
ggplot(weekday_morning, aes(x = TRIPS)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(aes(xintercept = mean_trips), 
             color = "red", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = median_trips), 
             color = "orange", linetype = "dashed", size = 1) +
  # Add text labels for mean and median
  annotate("text", x = mean_trips, y = Inf, 
           label = paste0("Mean: ", round(mean_trips, 0)), 
           vjust = 2, hjust = -0.1, color = "red", size = 3.5) +
  annotate("text", x = median_trips, y = Inf, 
           label = paste0("Median: ", round(median_trips, 0)), 
           vjust = 4, hjust = -0.1, color = "orange", size = 3.5) +
  # Format x-axis with comma separators
  scale_x_continuous(
    labels = label_number(scale = 1/1000, suffix = "k"),
    breaks = seq(0, max(weekday_morning$TRIPS), by = 20000)
  ) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Distribution of Trip Counts - Weekday Morning Peak (6am-9am)",
    subtitle = paste0("n = ", nrow(weekday_morning), " hexagons | 95th percentile: ", 
                     round(q95, 0), " trips"),
    x = "Number of Trips",
    y = "Frequency (Number of Hexagons)") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10, color = "gray40"),
    axis.title = element_text(size = 11),
    panel.grid.minor = element_blank()
  )
```
The histogram of trip counts during the weekday morning peak (6–9am) shows a clear right-skewed distribution, where most hexagons generate relatively few trips and only a small number experience very high volumes. The median trip count is 14,934, while the mean is 30,884, reflecting the influence of extreme outliers. The 95th percentile is much higher at 110,640 trips, with a few hexagons exceeding even this threshold, reaching nearly 400,000 trips. This indicates that while the majority of areas have modest bus activity, a small set of high-demand locations—likely major housing estates or transport hubs—contribute disproportionately to overall flows. The skew highlights an uneven distribution of bus demand, with most places generating low-to-moderate trips, but a few nodes acting as critical centers of commuter activity.

## 4.2 Local Indicators of Spatial Association(LISA)

```{r}
critical_dist <- critical_threshold(st_geometry(weekday_morning))

nb <- st_dist_band(
  st_geometry(weekday_morning),
  upper = critical_dist)

n_neighbors <- lengths(nb)

wt <- st_weights(nb, style = "W")

# Compute Local Moran's I
set.seed(1234)
lisa_morning <- weekday_morning %>%
  mutate(
    local_moran = local_moran(
      TRIPS, nb, wt, nsim = 99), 
    .before = 1) %>%
  unnest(local_moran)
```

### Local Moran’s I values and p-values

```{r}
ii_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +  # base layer

  tm_shape(lisa_morning) +
  tm_polygons(fill = "ii",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 5,
                values = "brewer.RdBu"
              ),
              fill.legend = tm_legend(
                title = "Local Moran's I",
                position = tm_pos_in("right", "bottom")
              )
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_title("Local Moran's I - morning peak")

p_ii_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +

  tm_shape(lisa_morning) +
  tm_polygons(fill = "p_ii",
              fill.scale = tm_scale_intervals(
                breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
                values = "-brewer.Reds"
              ),
              fill.legend = tm_legend(
                title = "p-value",
                position = tm_pos_in("right", "bottom")
              )
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_title("p-values of Local Moran's I")

tmap_arrange(ii_map, p_ii_map, asp = 1, ncol = 2)
```

The Local Moran’s I geovisualisation for the weekday morning peak reveals distinct spatial clustering of bus trip activity across Singapore.

On the left map, areas shaded in dark blue (high positive Local Moran’s I) represent clusters of high trip counts located near other high-trip areas, forming significant commuting hotspots. These are mainly seen in the western corridor (Jurong East, Clementi), northeastern estates (Hougang, Sengkang, Punggol), and northern towns (Woodlands, Yishun), which are key residential areas generating large commuter flows. In contrast, areas shaded in brown to red indicate negative values, suggesting spatial outliers—locations with high or low trip counts surrounded by the opposite trend. These tend to occur in industrial and peripheral regions, where demand is patchy.

The right map shows the p-values of Local Moran’s I, highlighting the statistical significance of these clusters. Darker red shades (<0.01) point to highly significant clusters concentrated in central, northern, and northeastern Singapore, confirming that spatial clustering of trip demand in these regions is not random but systematic. Areas with pale shades (>0.1) indicate no statistically significant clustering, reflecting more dispersed or random travel demand patterns.

### Moran scatterplot with standardised variable

```{r}
lisa_morning <- lisa_morning %>%
  mutate(lag_TRIPS = st_lag(
    TRIPS, nb, wt),
    .before = 1) %>%
  unnest(lag_TRIPS)

lisa_morning <- lisa_morning %>%
  mutate(z_TRIPS = scale(TRIPS),
         z_lag_TRIPS = scale(lag_TRIPS),
         .before = 1)

ggplot(data = lisa_morning,
       aes(x = z_TRIPS,
           y = z_lag_TRIPS,
           color = mean)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm",
              se = FALSE,
              color = "black") +
  geom_hline(yintercept = mean(lisa_morning$z_lag_TRIPS), lty = 2) +
  geom_vline(xintercept = mean(lisa_morning$z_TRIPS), lty = 2) +
  scale_color_manual(
    values = c("High-High" = "red",
               "Low-Low" = "blue",
               "Low-High" = "lightblue",
               "High-Low" = "pink")) +
  labs(x = "Standardised TRIPS",
       y = "Standardised Spatial Lag of TRIPS",
       title = "Standardised Moran Scatterplot with LISA Quadrants") +
  theme_minimal()
```

The scatter plot provides a statistical interpretation of the spatial autocorrelation in weekday morning trips. The strong clustering of points in the High-High quadrant (red) shows areas where hexagons with high trip counts are surrounded by other high-trip hexagons, suggesting distinct hotspots of morning bus activity. These clusters are often linked to dense residential or employment zones with high commuting demand. Conversely, the Low-Low quadrant (blue) indicates cold spots, where low-trip hexagons are spatially clustered, likely in peripheral or less populated regions.

The smaller presence of High-Low (pink) and Low-High (light blue) points reflects outliers. High-Low points represent isolated high-trip hexagons surrounded by low-trip areas, suggesting unique hubs, while Low-High points indicate underperforming hexagons embedded within busier regions. The regression line slope being positive supports the presence of overall positive spatial autocorrelation, meaning that similar trip intensities tend to cluster together.

In summary, the scatterplot confirms that weekday morning bus trips are not randomly distributed, but spatially concentrated into significant hotspot and coldspot clusters, with limited outliers. This provides statistical evidence of strong spatial dependence in travel demand patterns during morning peak hours.

### LISA Map

```{r}
lisa_sig <- lisa_morning %>%
  filter(p_ii < 0.05)

tmap_mode("plot")

tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  
  tm_shape(lisa_morning) +
  tm_polygons(fill = "TRIPS",
              fill.scale = tm_scale_continuous(
                values = "greys"
              ),
              fill.legend = tm_legend(
                title = "Total TRIPS",
                position = tm_pos_in("right", "bottom")
              )
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  
  tm_shape(lisa_sig) +
  tm_polygons(fill = "mean",
              fill.scale = tm_scale_categorical(
                values = c("High-High" = "red",
                          "Low-Low" = "blue",
                          "Low-High" = "lightblue",
                          "High-Low" = "pink")
              ),
              fill.legend = tm_legend(
                title = "LISA Quadrants (p < 0.05)",
                position = tm_pos_in("left", "bottom")
              )
  ) +
  tm_borders(col = "black", lwd = 0.5) +
  tm_title("LISA Map - Weekday Morning Peak (6am-9am)")
```

The LISA map for weekday morning peak reveals clear clusters of significant spatial autocorrelation in bus trip volumes across Singapore. High-High clusters (in red) appear prominently in the central and eastern regions, including dense residential and mixed-use areas, indicating concentrations of high trip volumes surrounded by other high-volume hexagons. These represent strong hotspots of demand during the morning peak. Conversely, Low-Low clusters (in dark blue) are mostly located in the western and peripheral regions, where low trip counts are surrounded by other low-volume areas, reflecting weaker bus demand in suburban or less connected zones.

Notably, some Low-High and High-Low outliers (in light blue and pink) emerge along the boundaries of clusters, suggesting transitional zones where high-demand cells are adjacent to low-demand cells and vice versa. The presence of these outliers indicates local heterogeneity and the influence of bordering subzones. Overall, the spatial patterns confirm that bus trips are not randomly distributed but exhibit strong clustering aligned with Singapore’s urban structure, with demand hotspots concentrated in central/eastern hubs and cold spots at the urban periphery.

## 4.3 Hot Spots and Cold Spots Analysis (HCSA)

### Gi* values and p-values

```{r}
# Compute fixed distance weight matrix
ct <- critical_threshold(st_geometry(weekday_morning))

weekday_morning_fdw <- weekday_morning %>%
  mutate(nb = include_self(
    st_dist_band(
      st_geometry(geometry),
      upper = ct)),
    wt = st_weights(nb,
                    style = "W"),
    .before = 1)

# Compute local Gi*

set.seed(1234)

HCSA_fdw <- weekday_morning_fdw %>%
  mutate(
    gistar = local_gstar_perm(
      TRIPS, nb, wt, nsim = 99),
    .before = 1) %>%
  unnest(gistar)

# Plot Gi* values
Gi_star_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  
  tm_shape(HCSA_fdw) +
  tm_polygons(fill = "gi_star",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 5,
                values = "-brewer.rd_bu"),
              fill.legend = tm_legend(
                title = "local Gi*",
                position = tm_pos_in(
                  "right", "bottom"))) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_title("Local Gi* of TRIPS")

# Plot p-values
p_values_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  
  tm_shape(HCSA_fdw) +
  tm_polygons(fill = "p_sim",
              fill.scale = tm_scale_intervals(
                breaks = c(0, 0.001, 0.01, 0.05, 0.1, 1),
                values = "-brewer.reds"),
              fill.legend = tm_legend(
                title = "p-value",
                position = tm_pos_in("right", "bottom"))) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_title("p-values of local Gi* of TRIPS")

# Display both maps side by side
tmap_arrange(Gi_star_map, p_values_map, asp = 1, ncol = 2)
```

The Local Gi* map (left) highlights concentrated hotspots in central and eastern regions, shown in red, where bus trip intensity is significantly higher than expected. Conversely, coldspots appear in the west and northwest, marked in blue, where trip counts are consistently lower relative to their neighbours. This contrast reflects commuting dynamics, with central and eastern hubs acting as major destinations.

The p-value map (right) supports the robustness of these patterns. Many hotspots and coldspots overlap with areas of low p-values (<0.05, dark red shades), confirming their statistical significance and suggesting that the clustering is unlikely to be due to random variation. Meanwhile, lighter shades (higher p-values) indicate regions where clustering is weaker or not statistically reliable.

### HCSA Map

```{r}
# HCSA cluster classification
HCSA_fdw <- HCSA_fdw %>%
  mutate(HCSA_cluster = case_when(
    p_sim > 0.05 ~ "Insignificant",
    p_sim <= 0.05 & cluster == "High" ~ "Hot spot",
    p_sim <= 0.05 & cluster == "Low" ~ "Cold spot",
    TRUE ~ "Other"),
    HCSA_cluster = factor(
      HCSA_cluster,
      levels = c("Insignificant", "Hot spot", "Cold spot")
    ),
    .before = 1
  )

# Plot HCSA cluster map
HCSA_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  
  tm_shape(HCSA_fdw) +
  tm_polygons(
    fill = "HCSA_cluster",
    fill.scale = tm_scale_categorical(
      values = c(
        "grey80",  # Insignificant
        "red",     # Hot spot
        "blue"     # Cold spot
      )
    ),
    fill.legend = tm_legend(
      title = "HCSA Cluster",
      position = tm_pos_in("right", "bottom"))
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_title("HCSA Clusters (p < 0.05)")

# Gi* map and HCSA map
tmap_arrange(Gi_star_map, HCSA_map, asp = 1, ncol = 2)
```

The right-hand map, showing Hot Spot and Cold Spot Analysis (HCSA) clusters, highlights statistically significant spatial clusters of bus trips during the weekday morning peak (6–9am). Red areas represent hot spots, where hexagons with high trip counts are surrounded by other high trip hexagons. These are concentrated in central and eastern Singapore, indicating consistently high demand in dense urban and employment hubs. Blue areas represent cold spots, where low trip counts cluster together, and these are more evident in peripheral regions such as the western and north-western areas. Grey areas indicate insignificant clusters, where trip patterns do not form clear spatial groupings. This contrast emphasizes that bus trip demand is not evenly distributed but spatially clustered: strong activity occurs in central commuting corridors and urban nodes, while peripheral regions consistently experience low demand.

Together, the two maps complement each other: the Local Gi* shows where clustering intensity is strongest, while the HCSA map strengthens interpretation by focusing only on statistically robust clusters.

## 4.4 Emerging Hot Spot Analysis (EHSA)

```{r}
hex_ids <- hexagon$HEX_ID

trips_morning_hourly <- trips %>%
  filter(DAY_TYPE == "WEEKDAY", HOUR_OF_DAY %in% 5:10) %>%
  group_by(HEX_ID, HOUR_OF_DAY) %>%
  summarise(TRIPS = sum(TRIPS), .groups = "drop") %>%
  tidyr::complete(HEX_ID = hex_ids, HOUR_OF_DAY = 5:10,
                  fill = list(TRIPS = 0)) %>%
  arrange(HEX_ID, HOUR_OF_DAY)

trips_morning_hourly$HEX_ID <- factor(trips_morning_hourly$HEX_ID, levels = hex_ids)
trips_morning_hourly$HOUR_OF_DAY <- as.integer(trips_morning_hourly$HOUR_OF_DAY)

# Build space–time cube
TRIPS_morning_st <- spacetime(
  trips_morning_hourly,
  hexagon %>% select(HEX_ID), 
  .loc_col  = "HEX_ID",
  .time_col = "HOUR_OF_DAY"
)

is_spacetime_cube(TRIPS_morning_st) 
```

```{r}
# Add neighbours and weights
TRIPS_morning_nb <- TRIPS_morning_st %>%
  activate("geometry") %>%
  mutate(
    nb = include_self(st_contiguity(geometry)), 
    wt = st_inverse_distance(nb, geometry, scale = 1, alpha = 1),
    .before = 1
  ) %>%
  set_nbs("nb") %>%
  set_wts("wt")

# Compute Gi* 
set.seed(1234)

gi_morning <- TRIPS_morning_nb %>%
  group_by(HOUR_OF_DAY) %>%
  mutate(gi_star = local_gstar_perm(TRIPS, nb, wt, nsim = 99)) %>%
  tidyr::unnest(gi_star) %>%
  ungroup()
```

### Mann-Kendall Test

H0: No monotonic trend

H1: Monotonic trend is present

```{r}
mk_hotspots <- gi_morning %>%
  st_drop_geometry() %>%
  group_by(HEX_ID) %>%
  summarise(
    mk = if(n() >= 3) list(unclass(Kendall::MannKendall(gi_star))) else NA,
    .groups = "drop"
  ) %>%
  tidyr::unnest_wider(mk) %>%
  rename(tau = tau, p_value = sl)

mk_hotspots_ranked <- mk_hotspots %>%
  filter(!is.na(p_value)) %>%                
  arrange(p_value, desc(abs(tau)))        

mk_hotspots_ranked %>% slice(1:10)
```

With reference to the results, we will reject the null hypothesis and infer that there is an upward trend trend.

The Hex H0276 is selected to view the Gi* Trend for morning peak:

```{r}
hex_trend <- gi_morning %>%
  st_drop_geometry() %>%
  filter(HEX_ID == "H0276") %>%
  select(HEX_ID, HOUR_OF_DAY, gi_star)

ggplot(data = hex_trend, aes(x = HOUR_OF_DAY, y = gi_star)) +
  geom_line(color = "black") +
  labs(
    title = "Gi* Trend across Morning Hours — H0276",
    x = "Hour of Day",
    y = "Gi*"
  ) +
  theme_light()
```

The plot for Hex H0276 during the morning 5–10am shows a positive trend in the Gi* statistic. At the start of the time window (5–7am), the Gi* values remain relatively low and negative, indicating that this location is associated with a cold spot of bus trips, where trip intensity is below its surrounding areas. However, from 7am onward, there is a clear upward movement, with Gi* values steadily increasing and approaching zero by 10am.

It confirms the presence of a monotonic increasing trend, suggesting that spatial clustering of trips at this location becomes stronger through the later morning hours. While still remaining below zero (indicating weaker clustering compared to strong hotspots), the improvement signals that demand in this area becomes more aligned with busier surrounding hexagons as the peak progresses.

### Performing Emerging Hotspot Analysis

```{r}
set.seed(1234)

ehsa_morning <- emerging_hotspot_analysis(
  x     = TRIPS_morning_nb,   
  .var  = "TRIPS",     
  k     = 1,      
  nsim  = 99   
)

ehsa_morning_sf <- hexagon %>%
  select(HEX_ID) %>%
  left_join(ehsa_morning, by = c("HEX_ID" = "location")) %>%
  filter(p_value < 0.05) %>%
  st_as_sf()
```

```{r}
ehsa_levels <- c(
  "consecutive coldspot", "consecutive hotspot",
  "new coldspot", "new hotspot",
  "no pattern detected",
  "persistent coldspot", "persistent hotspot",
  "sporadic coldspot", "sporadic hotspot"
)

ehsa_colors <- c(
  "consecutive coldspot" = "#4a90e2",   # dark blue
  "consecutive hotspot"  = "#e31a1c",   # dark red
  "new coldspot"         = "#20b2aa",   # light blue
  "new hotspot"          = "#ff69b4",   # light red/orange
  "no pattern detected"  = "grey80",    # neutral grey
  "persistent coldspot"  = "#08519c",   # very dark blue
  "persistent hotspot"   = "#8b0000",   # very dark red
  "sporadic coldspot"    = "#c6dbef",   # pastel blue
  "sporadic hotspot"     = "#fcae91"    # pastel orange
)

ehsa_morning$classification   <- factor(ehsa_morning$classification,   levels = ehsa_levels)
ehsa_morning_sf$classification <- factor(ehsa_morning_sf$classification, levels = ehsa_levels)
```


```{r}
ggplot(data = ehsa_morning, 
       aes(x = classification, fill = classification)) +
  geom_bar(color = "black", alpha = 0.8) +
  scale_fill_manual(values = ehsa_colors, drop = FALSE) +
  labs(
    title = "Distribution of EHSA Classes – Weekday Morning (5–10am)",
    x = "EHSA Classification", 
    y = "Number of Hexagons"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", size = 14)
  )

```

The EHSA classification distribution for weekday mornings (5–10am) reveals that the majority of hexagons fall under no pattern detected, indicating that most areas do not exhibit significant or consistent hotspot or coldspot behaviour. However, a substantial number are classified as sporadic coldspots and sporadic hotspots, reflecting localized and irregular clustering of both low and high bus trip counts. Persistent coldspots are also evident, though persistent hotspots are largely absent, suggesting that consistently high-demand areas are rare in this period. Overall, the results highlight a spatially uneven demand pattern, with significant variability across different parts of the network.

```{r}
tmap_mode("plot")
tm_shape(mpsz) + tm_polygons(col = "grey90", border.col = "white") +
tm_shape(ehsa_morning_sf) +
  tm_polygons(
    fill = "classification",
    fill.scale = tm_scale_categorical(values = ehsa_colors),
    fill.legend = tm_legend(
      title = "EHSA Classification",
      position = tm_pos_in("right", "bottom"))
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_title("Emerging Hot Spot Analysis (EHSA) – Weekday Morning (5–10am)")
```

In the EHSA map, most hexagons are classified as no pattern detected, indicating stable or insignificant fluctuations in trip counts across much of the city. However, several areas show distinct hotspot or coldspot dynamics.

In the western and northwestern regions, there is a clear concentration of persistent coldspots (dark blue), where consistently low trip counts are clustered across time. This suggests peripheral estates or less connected areas with structurally lower bus demand. Conversely, sporadic hotspots (peach) and consecutive hotspots (red) emerge in central-western areas, particularly near major transport corridors. These areas likely experience occasional or sustained surges in bus demand, reflecting commuting activity into employment or mixed-use hubs.

A smaller number of new hotspots (pink) and sporadic coldspots (light blue) also appear, pointing to dynamic shifts in travel intensity at localized scales. These classifications are critical because they indicate not just where demand is high or low, but whether such patterns are emerging, persistent, or episodic.


# 5 Weekday afternoon peak

## 5.1 Overview

### Choropleth map of total trips

```{r}
weekday_afternoon <- trips %>%
  filter(DAY_TYPE == "WEEKDAY", HOUR_OF_DAY %in% 17:20) %>%
  group_by(HEX_ID) %>%
  summarise(TRIPS = sum(TRIPS), .groups = "drop") %>%
  left_join(hexagon, by = "HEX_ID") %>%
  st_as_sf()

tmap_mode("plot")

tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  
  tm_shape(weekday_afternoon) +
  tm_polygons(
    fill = "TRIPS",
    fill.scale = tm_scale_intervals(
      style = "quantile",  # Use style = "quantile" instead
      n = 5,
      values = "viridis"
    ),
    fill.legend = tm_legend(
      title = "Total Trips\n(Weekday Afternoon)",
      position = tm_pos_in("right", "bottom")
    )
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_title("Distribution of Bus Trips - Weekday Afternoon Peak (5pm-8pm)") +
  tm_compass(position = tm_pos_in("left", "top")) +
  tm_scale_bar(position = tm_pos_in("left", "bottom"))
```

The visualisation shows that bus trips during the weekday afternoon peak (5–8pm) are highly concentrated in central and western regions such as Jurong East, Clementi, and Queenstown, as well as large residential towns including Tampines, Yishun, and Woodlands. These areas reflect strong homeward-bound commuting flows as people return from central business districts, universities, and industrial hubs. Moderate activity is evident in northern and northeastern towns such as Sengkang, Hougang, and Punggol, which act as important residential catchments. In contrast, peripheral zones, especially coastal and industrial areas, record relatively low trip counts. The overall spatial pattern highlights the dominance of evening return journeys, with trips dispersing across multiple residential clusters rather than converging on central areas.

### Trip Distribution

```{r}
mean_trips <- mean(weekday_afternoon$TRIPS)
median_trips <- median(weekday_afternoon$TRIPS)
q75 <- quantile(weekday_afternoon$TRIPS, 0.75)
q95 <- quantile(weekday_afternoon$TRIPS, 0.95)

ggplot(weekday_afternoon, aes(x = TRIPS)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(aes(xintercept = mean_trips), 
             color = "red", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = median_trips), 
             color = "orange", linetype = "dashed", size = 1) +
  annotate("text", x = mean_trips * 1.15, y = Inf, 
           label = paste0("Mean: ", round(mean_trips, 0)), 
           vjust = 1.5, color = "red", size = 3.5, fontface = "bold") +
  annotate("text", x = median_trips * 0.85, y = Inf, 
           label = paste0("Median: ", round(median_trips, 0)), 
           vjust = 3, color = "orange", size = 3.5, fontface = "bold") +
  coord_cartesian(xlim = c(0, 240000)) +
  scale_x_continuous(
    labels = label_number(scale = 1/1000, suffix = "k"),
    breaks = seq(0, 240000, by = 40000)  # 0k, 40k, 80k, 120k, 160k, 200k, 240k
  ) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Distribution of Trip Counts - Weekday Afternoon Peak (5pm-8pm)",
    subtitle = paste0("n = ", nrow(weekday_afternoon), " hexagons | 95th percentile: ", 
                     round(q95, 0), " trips"),
    x = "Number of Trips (Thousands)",
    y = "Frequency (Number of Hexagons)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10, color = "gray40"),
    axis.title = element_text(size = 11),
    panel.grid.minor = element_blank()
  )
```

The histogram of trip counts during the weekday afternoon peak (5–8pm) also shows a pronounced right-skewed distribution, similar to the morning peak. Most hexagons record relatively low trip volumes, while a small number account for very high values. The median trip count is 14,872, but the mean is much higher at 30,263, reflecting the influence of a few extreme outliers. The 95th percentile reaches 100,900 trips, and some hexagons exceed this, with the maximum observed count surpassing 200,000 trips.

This skewed pattern indicates that while the majority of areas contribute modestly to afternoon bus usage, a limited set of high-demand zones—likely large residential towns, key transfer nodes, or major employment areas—generate disproportionately high passenger flows. Compared to the morning peak, the afternoon distribution suggests similarly uneven demand, with commuter flows dispersing across many neighborhoods but still dominated by a small number of critical hubs.

## 5.2 Local Indicators of Spatial Association(LISA)

```{r}
critical_dist <- critical_threshold(st_geometry(weekday_afternoon))

nb <- st_dist_band(
  st_geometry(weekday_afternoon),
  upper = critical_dist)

n_neighbors <- lengths(nb)

wt <- st_weights(nb, style = "W")

# Compute Local Moran's I
set.seed(1234)
lisa_afternoon <- weekday_afternoon %>%
  mutate(
    local_moran = local_moran(
      TRIPS, nb, wt, nsim = 99), 
    .before = 1) %>%
  unnest(local_moran)

glimpse(lisa_afternoon)
```

### Local Moran’s I values and p-values

```{r}
ii_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +  

  tm_shape(lisa_afternoon) +
  tm_polygons(fill = "ii",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 5,
                values = "brewer.RdBu"
              ),
              fill.legend = tm_legend(
                title = "Local Moran's I",
                position = tm_pos_in("right", "bottom")
              )
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_title("Local Moran's I - afternoon peak")

p_ii_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +

  tm_shape(lisa_afternoon) +
  tm_polygons(fill = "p_ii",
              fill.scale = tm_scale_intervals(
                breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
                values = "-brewer.Reds"
              ),
              fill.legend = tm_legend(
                title = "p-value",
                position = tm_pos_in("right", "bottom")
              )
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_title("p-values of Local Moran's I")

tmap_arrange(ii_map, p_ii_map, asp = 1, ncol = 2)
```

On the left map, positive values (blue shades) indicate clusters where high-trip hexagons are located near other high-trip hexagons, forming strong evening hotspots. These are mainly concentrated in the western corridor, central regions, and northeastern towns, reflecting strong residential demand as commuters return home in the evening. In contrast, negative values (brown to red) represent spatial outliers, where high-trip areas are surrounded by low-trip neighbors or vice versa. These appear in more peripheral or industrial areas, where trip demand is uneven.

The right map of p-values shows the statistical significance of these clusters. Darker red areas (<0.01) highlight significant clustering, particularly in central, northern, and northeastern Singapore, confirming that evening trip concentrations in these regions are systematic rather than random. Lighter shades (>0.1) denote regions with weaker or insignificant clustering, where travel patterns are more dispersed.

Overall, the maps indicate that weekday afternoon trips are strongly clustered in key residential and transport hubs, reinforcing their role as destinations during the evening return commute, while peripheral areas remain less significant in shaping overall trip patterns.

### Moran scatterplot with standardised variable

```{r}
lisa_afternoon <- lisa_afternoon %>%
  mutate(lag_TRIPS = st_lag(
    TRIPS, nb, wt),
    .before = 1) %>%
  unnest(lag_TRIPS)

lisa_afternoon <- lisa_afternoon %>%
  mutate(z_TRIPS = scale(TRIPS),
         z_lag_TRIPS = scale(lag_TRIPS),
         .before = 1)

ggplot(data = lisa_afternoon,
       aes(x = z_TRIPS,
           y = z_lag_TRIPS,
           color = mean)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm",
              se = FALSE,
              color = "black") +
  geom_hline(yintercept = mean(lisa_afternoon$z_lag_TRIPS), lty = 2) +
  geom_vline(xintercept = mean(lisa_afternoon$z_TRIPS), lty = 2) +
  scale_color_manual(
    values = c("High-High" = "red",
               "Low-Low" = "blue",
               "Low-High" = "lightblue",
               "High-Low" = "pink")) +
  labs(x = "Standardised TRIPS",
       y = "Standardised Spatial Lag of TRIPS",
       title = "Standardised Moran Scatterplot with LISA Quadrants") +
  theme_minimal()
```

The scatter plot for the weekday afternoon peak shows a clear dominance of High-High clusters (red) and Low-Low clusters (blue), confirming that bus trips remain strongly spatially autocorrelated during the evening commute. The regression line indicates positive spatial dependence, meaning similar trip levels tend to group together. Outliers in the High-Low and Low-High quadrants are limited, suggesting that most demand patterns are consistent across neighboring areas.

### LISA Map

```{r}
lisa_sig <- lisa_afternoon %>%
  filter(p_ii < 0.05)

tmap_mode("plot")

tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  
  tm_shape(lisa_afternoon) +
  tm_polygons(fill = "TRIPS",
              fill.scale = tm_scale_continuous(
                values = "greys"
              ),
              fill.legend = tm_legend(
                title = "Total TRIPS",
                position = tm_pos_in("right", "bottom")
              )
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  
  tm_shape(lisa_sig) +
  tm_polygons(fill = "mean",
              fill.scale = tm_scale_categorical(
                values = c("High-High" = "red",
                          "Low-Low" = "blue",
                          "Low-High" = "lightblue",
                          "High-Low" = "pink")
              ),
              fill.legend = tm_legend(
                title = "LISA Quadrants (p < 0.05)",
                position = tm_pos_in("left", "bottom")
              )
  ) +
  tm_borders(col = "black", lwd = 0.5) +
  tm_title("LISA Map - Weekday Afternoon Peak (6am-9am)")
```

The LISA map for the weekday afternoon peak (5–8pm) reveals distinct spatial clustering of bus trip volumes across Singapore. High-High clusters (red) are concentrated in central, northeastern, and selected western areas, where hexagons with high trip counts are surrounded by other high-demand cells. These represent major evening hotspots associated with residential estates and transport nodes, reflecting strong commuter flows returning home during the afternoon peak. In contrast, Low-Low clusters (dark blue) appear in western and peripheral regions, where low trip volumes cluster together, highlighting areas with consistently weaker bus demand.

A few Low-High and High-Low outliers (light blue and pink) emerge at the edges of clusters, suggesting transitional zones where high-demand cells border low-demand areas and vice versa. These outliers point to localized heterogeneity, possibly shaped by differences in land use or accessibility within adjacent subzones.

Overall, the afternoon LISA map confirms that bus trips are not randomly distributed but form clear hotspot and coldspot clusters, consistent with Singapore’s urban commuting structure. Hotspots are aligned with key residential hubs and employment-linked zones, while coldspots dominate peripheral areas, reinforcing the spatial divide between dense commuter corridors and low-demand regions.


## 5.3 Hot Spots and Cold Spots Analysis (HCSA)

### Gi* values and p-values

```{r}
# Compute fixed distance weight matrix
ct <- critical_threshold(st_geometry(weekday_afternoon))

weekday_afternoon_fdw <- weekday_afternoon %>%
  mutate(nb = include_self(
    st_dist_band(
      st_geometry(geometry),
      upper = ct)),
    wt = st_weights(nb,
                    style = "W"),
    .before = 1)

# Compute local Gi*

set.seed(1234)

HCSA_fdw <- weekday_afternoon_fdw %>%
  mutate(
    gistar = local_gstar_perm(
      TRIPS, nb, wt, nsim = 99),
    .before = 1) %>%
  unnest(gistar)

# Plot Gi* values
Gi_star_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  
  tm_shape(HCSA_fdw) +
  tm_polygons(fill = "gi_star",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 5,
                values = "-brewer.rd_bu"),
              fill.legend = tm_legend(
                title = "local Gi*",
                position = tm_pos_in(
                  "right", "bottom"))) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_title("Local Gi* of TRIPS")

# Plot p-values
p_values_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  
  tm_shape(HCSA_fdw) +
  tm_polygons(fill = "p_sim",
              fill.scale = tm_scale_intervals(
                breaks = c(0, 0.001, 0.01, 0.05, 0.1, 1),
                values = "-brewer.reds"),
              fill.legend = tm_legend(
                title = "p-value",
                position = tm_pos_in("right", "bottom"))) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_title("p-values of local Gi* of TRIPS")

# Display both maps side by side
tmap_arrange(Gi_star_map, p_values_map, asp = 1, ncol = 2)
```

The Local Gi* map (left) for the weekday afternoon peak shows clear clustering of bus trip intensity across Singapore. Hotspots (red shades) are concentrated in the central, northeastern, and selected western areas, where hexagons register significantly higher trip volumes than their neighbours. These reflect strong evening commuting flows, as passengers return to major residential towns and transport hubs. Conversely, coldspots (blue shades) are evident in the western and northwestern regions, where bus trip counts are consistently lower than surrounding areas, likely reflecting peripheral industrial or less populated zones with weaker demand.

The p-value map (right) reinforces the statistical significance of these patterns. Dark red areas (<0.05) overlap with many identified hotspots and coldspots, confirming that clustering is unlikely to have occurred by chance. In contrast, lighter shades represent areas where spatial clustering is weaker or not statistically significant.

Together, the maps confirm that afternoon bus trips are highly clustered, with strong hotspots forming around dense residential and commuting corridors, while peripheral areas act as stable coldspots. These results highlight the spatially uneven but systematic nature of evening travel demand.

### HCSA Map

```{r}
# HCSA cluster classification
HCSA_fdw <- HCSA_fdw %>%
  mutate(HCSA_cluster = case_when(
    p_sim > 0.05 ~ "Insignificant",
    p_sim <= 0.05 & cluster == "High" ~ "Hot spot",
    p_sim <= 0.05 & cluster == "Low" ~ "Cold spot",
    TRUE ~ "Other"),
    HCSA_cluster = factor(
      HCSA_cluster,
      levels = c("Insignificant", "Hot spot", "Cold spot")
    ),
    .before = 1
  )

# Plot HCSA cluster map
HCSA_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  
  tm_shape(HCSA_fdw) +
  tm_polygons(
    fill = "HCSA_cluster",
    fill.scale = tm_scale_categorical(
      values = c(
        "grey80",  # Insignificant
        "red",     # Hot spot
        "blue"     # Cold spot
      )
    ),
    fill.legend = tm_legend(
      title = "HCSA Cluster",
      position = tm_pos_in("right", "bottom"))
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_title("HCSA Clusters (p < 0.05)")

# Gi* map and HCSA map
tmap_arrange(Gi_star_map, HCSA_map, asp = 1, ncol = 2)
```

The right HCSA map for the weekday afternoon peak shows statistically significant cold spots (blue) dominate the western and northern peripheries, reflecting consistently low trip activity in less densely populated residential or industrial zones. In contrast, hot spots (red) are fewer and more scattered but appear in selected central and northeastern areas, aligning with high-demand nodes such as transport interchanges or dense residential–employment corridors. Grey areas represent insignificant clusters, indicating that many locations do not form statistically robust groupings, with trip activity distributed more heterogeneously.

When viewed alongside the Local Gi* map (left), which highlights the intensity of clustering, the HCSA map strengthens interpretation by filtering out noise and showing only statistically robust patterns. For the afternoon peak, the HCSA results confirm that significant clustering of high trips is limited compared to the morning, while cold spots remain more widespread and consistent. This contrast suggests a more dispersed commuting demand in the evening, as workers and students return to residential areas from multiple destinations. Overall, the maps together reveal that afternoon bus trip patterns, while still spatially clustered, are less dominated by strong hotspots and instead emphasize widespread cold spot clustering at the periphery.

## 5.4 Emerging Hot Spot Analysis (EHSA)

```{r}
hex_ids <- hexagon$HEX_ID

trips_afternoon_hourly <- trips %>%
  filter(DAY_TYPE == "WEEKDAY", HOUR_OF_DAY %in% 16:21) %>%
  group_by(HEX_ID, HOUR_OF_DAY) %>%
  summarise(TRIPS = sum(TRIPS), .groups = "drop") %>%
  tidyr::complete(HEX_ID = hex_ids, HOUR_OF_DAY = 16:21,
                  fill = list(TRIPS = 0)) %>%
  arrange(HEX_ID, HOUR_OF_DAY)

trips_afternoon_hourly$HEX_ID <- factor(trips_afternoon_hourly$HEX_ID, levels = hex_ids)
trips_afternoon_hourly$HOUR_OF_DAY <- as.integer(trips_afternoon_hourly$HOUR_OF_DAY)

# Build space–time cube
TRIPS_afternoon_st <- spacetime(
  trips_afternoon_hourly,
  hexagon %>% select(HEX_ID), 
  .loc_col  = "HEX_ID",
  .time_col = "HOUR_OF_DAY"
)

is_spacetime_cube(TRIPS_afternoon_st) 
```

```{r}
# Add neighbours and weights
TRIPS_afternoon_nb <- TRIPS_afternoon_st %>%
  activate("geometry") %>%
  mutate(
    nb = include_self(st_contiguity(geometry)), 
    wt = st_inverse_distance(nb, geometry, scale = 1, alpha = 1),
    .before = 1
  ) %>%
  set_nbs("nb") %>%
  set_wts("wt")

# Compute Gi* 
set.seed(1234)

gi_afternoon <- TRIPS_afternoon_nb %>%
  group_by(HOUR_OF_DAY) %>%
  mutate(gi_star = local_gstar_perm(TRIPS, nb, wt, nsim = 99)) %>%
  tidyr::unnest(gi_star) %>%
  ungroup()
```

### Mann-Kendall Test

H0: No monotonic trend

H1: Monotonic trend is present

```{r}
mk_hotspots_afternoon <- gi_afternoon %>%
  st_drop_geometry() %>%
  group_by(HEX_ID) %>%
  summarise(
    mk = if(n() >= 3) list(unclass(Kendall::MannKendall(gi_star))) else NA,
    .groups = "drop"
  ) %>%
  tidyr::unnest_wider(mk) %>%
  rename(tau = tau, p_value = sl)

mk_hotspots_afternoon_ranked <- mk_hotspots_afternoon %>%
  filter(!is.na(p_value)) %>%
  arrange(p_value, desc(abs(tau)))        

mk_hotspots_afternoon_ranked %>% slice(1:10)
```

With reference to the results, we will reject the null hypothesis and infer that there is an upward trend.

The Hex H0012 is selected to view the Gi* Trend for afternoon peak:

```{r}
hex_trend <- gi_afternoon %>%
  st_drop_geometry() %>%
  filter(HEX_ID == "H0012") %>%
  select(HEX_ID, HOUR_OF_DAY, gi_star)

ggplot(data = hex_trend, aes(x = HOUR_OF_DAY, y = gi_star)) +
  geom_line(color = "black") +
  labs(
    title = "Gi* Trend across Afternoon Hours — H0012",
    x = "Hour of Day",
    y = "Gi*"
  ) +
  theme_light()
```

The plot for Hex H0012 during the afternoon peak 4–9pm shows a clear upward trend in the Gi* statistic. At the start (around 4–5pm), Gi* values are more negative, indicating that this location is associated with a cold spot of bus trips, where trip intensity is significantly lower compared to its neighbors. However, as time progresses, the Gi* values steadily rise, especially between 6–7pm, and continue increasing until 9pm.

Although the Gi* values remain negative overall, the consistent upward trajectory suggests a monotonic increasing trend. This means that bus trip clustering in this location is gradually becoming less cold and more aligned with surrounding activity levels during the afternoon peak. In practical terms, demand here may start from relatively low levels but strengthens toward the later hours, reflecting increasing alignment with busier adjacent hexagons as commuters travel during evening peak hours.

### Performing Emerging Hotspot Analysis

```{r}
set.seed(1234)

ehsa_afternoon <- emerging_hotspot_analysis(
  x     = TRIPS_afternoon_nb,   
  .var  = "TRIPS",     
  k     = 1,      
  nsim  = 99   
)

ehsa_afternoon_sf <- hexagon %>%
  select(HEX_ID) %>%
  left_join(ehsa_afternoon, by = c("HEX_ID" = "location")) %>%
  filter(p_value < 0.05) %>%
  st_as_sf()
```


```{r}
ehsa_afternoon$classification   <- factor(ehsa_afternoon$classification,   levels = ehsa_levels)
ehsa_afternoon_sf$classification <- factor(ehsa_afternoon_sf$classification, levels = ehsa_levels)
```


```{r}
ggplot(data = ehsa_afternoon, 
       aes(x = classification, fill = classification)) +
  geom_bar(color = "black", alpha = 0.8) +
  scale_fill_manual(values = ehsa_colors, drop = FALSE) +
  labs(
    title = "Distribution of EHSA Classes – Weekday Afternoon (4–9pm)",
    x = "EHSA Classification", 
    y = "Number of Hexagons"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", size = 14)
  )
```

The EHSA distribution for the weekday afternoon peak 4–9pm shows that most hexagons fall under no pattern detected, suggesting that clustering of demand is limited across much of the city. Among significant categories, consecutive coldspots and persistent coldspots are notable, indicating stable zones of consistently low demand, particularly in less connected areas. Sporadic hotspots also appear, reflecting irregular but short-lived surges in demand at specific locations. By contrast, persistent and consecutive hotspots are nearly absent, underscoring the lack of sustained high-demand clusters during this time window. Overall, the afternoon peak is defined by widespread coldspot dominance, with hotspots emerging only intermittently.

```{r}
tmap_mode("plot")
tm_shape(mpsz) + tm_polygons(col = "grey90", border.col = "white") +
tm_shape(ehsa_afternoon_sf) +
  tm_polygons(
    fill = "classification",
    fill.scale = tm_scale_categorical(values = ehsa_colors),
    fill.legend = tm_legend(
      title = "EHSA Classification",
      position = tm_pos_in("right", "bottom"))
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_title("Emerging Hot Spot Analysis (EHSA) – Weekday Afternoon (4–9pm)")
```

The EHSA map for weekday afternoon peak (4–9pm) shows that consecutive hotspots (red) are concentrated in the central zone, suggesting consistent high bus demand during the afternoon peak, likely linked to commuting flows toward employment and mixed-use hubs. Surrounding these, sporadic hotspots (peach) appear, reflecting irregular but recurring surges in demand. On the other hand, new coldspots (teal) are scattered along the periphery, marking emerging zones of consistently lower demand. Importantly, no consecutive coldspots are observed, implying that while some low-demand areas exist, they are not stable or spatially persistent. Overall, the map highlights a pattern where central urban areas serve as focal points of afternoon demand, while peripheral regions show weaker and less structured clustering.


# 6 Weekend/holiday

## 6.1 Overview

### Choropleth map of total trips

```{r}
weekend <- trips %>%
  filter(DAY_TYPE == "WEEKENDS/HOLIDAY", HOUR_OF_DAY %in% 11:20) %>%
  group_by(HEX_ID) %>%
  summarise(TRIPS = sum(TRIPS), .groups = "drop") %>%
  left_join(hexagon, by = "HEX_ID") %>%
  st_as_sf()

tmap_mode("plot")

tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  
  tm_shape(weekend) +
  tm_polygons(
    fill = "TRIPS",
    fill.scale = tm_scale_intervals(
      style = "quantile",  # Use style = "quantile" instead
      n = 5,
      values = "viridis"
    ),
    fill.legend = tm_legend(
      title = "Total Trips\n(Weekend/holiday)",
      position = tm_pos_in("right", "bottom")
    )
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_title("Distribution of Bus Trips - Weekend/holiday (11am-8pm)") +
  tm_compass(position = tm_pos_in("left", "top")) +
  tm_scale_bar(position = tm_pos_in("left", "bottom"))
```

The visualisation of weekend and holiday bus trips (11am–8pm) shows a more dispersed and leisure-oriented travel pattern compared to weekday peaks. High trip volumes are concentrated across large residential towns such as Tampines, Pasir Ris, Woodlands, and Yishun, as well as around central areas, reflecting strong demand for shopping, recreation, and social visits. Notably, trips also cluster around regional hubs like Jurong East and Tampines, which function as retail and entertainment centres.

Moderate activity is spread across northern and northeastern towns including Hougang, Sengkang, and Punggol, while the western corridor shows balanced demand, likely linked to weekend mall and park visits. In contrast, peripheral coastal and industrial zones record relatively low activity, as they are less associated with weekend travel purposes.

Overall, the spatial distribution suggests that weekend and holiday bus demand is polycentric and more evenly spread across residential and leisure destinations, rather than being heavily dominated by work-related commuting flows. This highlights the role of buses in supporting recreational mobility and connecting suburban households to key shopping, social, and leisure hubs across the island.

### Trip Distribution

```{r}
mean_trips <- mean(weekend$TRIPS)
median_trips <- median(weekend$TRIPS)
q75 <- quantile(weekend$TRIPS, 0.75)
q95 <- quantile(weekend$TRIPS, 0.95)

ggplot(weekend, aes(x = TRIPS)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(aes(xintercept = mean_trips), 
             color = "red", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = median_trips), 
             color = "orange", linetype = "dashed", size = 1) +
  annotate("text", x = mean_trips * 1.15, y = Inf, 
           label = paste0("Mean: ", round(mean_trips, 0)), 
           vjust = 1.5, color = "red", size = 3.5, fontface = "bold") +
  annotate("text", x = median_trips * 0.85, y = Inf, 
           label = paste0("Median: ", round(median_trips, 0)), 
           vjust = 3, color = "orange", size = 3.5, fontface = "bold") +
  coord_cartesian(xlim = c(0, 240000)) +
  scale_x_continuous(
    labels = label_number(scale = 1/1000, suffix = "k"),
    breaks = seq(0, 240000, by = 40000)  # 0k, 40k, 80k, 120k, 160k, 200k, 240k
  ) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Distribution of Trip Counts - Weekend/holiday (11am-8pm)",
    subtitle = paste0("n = ", nrow(weekend), " hexagons | 95th percentile: ", 
                     round(q95, 0), " trips"),
    x = "Number of Trips (Thousands)",
    y = "Frequency (Number of Hexagons)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10, color = "gray40"),
    axis.title = element_text(size = 11),
    panel.grid.minor = element_blank()
  )
```

The histogram of trip counts during weekends and holidays (11am–8pm) shows a strongly right-skewed distribution. The majority of hexagons record relatively low trip volumes, with the median trip count at 11,742, indicating that half of the areas experience modest demand. However, the mean is much higher at 25,297, reflecting the influence of a small number of locations with very high passenger volumes. The 95th percentile reaches 89,450 trips, and a few hexagons far exceed this threshold, with maximum counts surpassing 200,000 trips.

This pattern highlights that while most areas have low-to-moderate weekend travel activity, a limited number of high-demand hubs dominate overall bus flows. The skew underscores an uneven demand distribution, where weekend and holiday travel is more dispersed compared to weekday commuting but still driven by concentrated hotspots of activity.

## 6.2 Local Indicators of Spatial Association(LISA)

```{r}
critical_dist <- critical_threshold(st_geometry(weekend))

nb <- st_dist_band(
  st_geometry(weekend),
  upper = critical_dist)

n_neighbors <- lengths(nb)

wt <- st_weights(nb, style = "W")

# Compute Local Moran's I
set.seed(1234)
lisa_weekend <- weekend %>%
  mutate(
    local_moran = local_moran(
      TRIPS, nb, wt, nsim = 99), 
    .before = 1) %>%
  unnest(local_moran)

glimpse(lisa_weekend)
```

### Local Moran’s I values and p-values

```{r}
ii_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +  

  tm_shape(lisa_weekend) +
  tm_polygons(fill = "ii",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 5,
                values = "brewer.RdBu"
              ),
              fill.legend = tm_legend(
                title = "Local Moran's I",
                position = tm_pos_in("right", "bottom")
              )
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_title("Local Moran's I - weekend/holiday peak")

p_ii_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +

  tm_shape(lisa_weekend) +
  tm_polygons(fill = "p_ii",
              fill.scale = tm_scale_intervals(
                breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
                values = "-brewer.Reds"
              ),
              fill.legend = tm_legend(
                title = "p-value",
                position = tm_pos_in("right", "bottom")
              )
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_title("p-values of Local Moran's I")

tmap_arrange(ii_map, p_ii_map, asp = 1, ncol = 2)
```

On the left map, positive values (blue shades) show clusters of high-trip hexagons adjacent to other high-trip hexagons, indicating areas of consistently strong demand. These clusters appear mainly in the central, northern, and northeastern regions, reflecting the role of major residential towns and popular activity hubs as key weekend trip generators. In contrast, negative values (brown shades) highlight spatial outliers—locations where demand patterns differ from their surroundings, such as pockets of low trips near busier zones or isolated hotspots in low-demand areas.

The right map of p-values reveals the statistical significance of these patterns. Darker red shades (<0.01) confirm highly significant clusters, concentrated in central and northern parts of the island. These indicate systematic spatial clustering of weekend bus activity, likely linked to shopping centres, recreational spaces, or well-connected transit nodes. Lighter shades (>0.1) mark areas with weaker or insignificant clustering, where demand is more dispersed and less structured.

Overall, the maps suggest that weekend and holiday bus trips are not randomly distributed but exhibit strong clustering in central and northern Singapore, reinforcing the importance of residential and recreational hubs in shaping travel demand during non-working days.

### Moran scatterplot with standardised variable

```{r}
lisa_weekend <- lisa_weekend %>%
  mutate(lag_TRIPS = st_lag(
    TRIPS, nb, wt),
    .before = 1) %>%
  unnest(lag_TRIPS)

lisa_weekend <- lisa_weekend %>%
  mutate(z_TRIPS = scale(TRIPS),
         z_lag_TRIPS = scale(lag_TRIPS),
         .before = 1)

ggplot(data = lisa_weekend,
       aes(x = z_TRIPS,
           y = z_lag_TRIPS,
           color = mean)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm",
              se = FALSE,
              color = "black") +
  geom_hline(yintercept = mean(lisa_weekend$z_lag_TRIPS), lty = 2) +
  geom_vline(xintercept = mean(lisa_weekend$z_TRIPS), lty = 2) +
  scale_color_manual(
    values = c("High-High" = "red",
               "Low-Low" = "blue",
               "Low-High" = "lightblue",
               "High-Low" = "pink")) +
  labs(x = "Standardised TRIPS",
       y = "Standardised Spatial Lag of TRIPS",
       title = "Standardised Moran Scatterplot with LISA Quadrants") +
  theme_minimal()
```

The scatter plot for weekend/holiday trips (11am–8pm) shows strong evidence of spatial autocorrelation, with two dominant clusters: High-High (red) and Low-Low (blue). The High-High clusters indicate hexagons with high bus trip volumes surrounded by similarly high-demand neighbors, reflecting key residential and activity hubs that attract sustained weekend travel. Conversely, the Low-Low clusters highlight peripheral or less connected areas where consistently low trip counts group together.

The regression line has a clear positive slope, reinforcing that trip intensities are spatially dependent—high values tend to cluster with high, and low with low. Outliers in the High-Low (pink) and Low-High (light blue) quadrants are relatively limited, suggesting that demand patterns during weekends and holidays are broadly consistent across neighboring hexagons

### LISA Map

```{r}
lisa_sig <- lisa_weekend %>%
  filter(p_ii < 0.05)

tmap_mode("plot")

tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  
  tm_shape(lisa_weekend) +
  tm_polygons(fill = "TRIPS",
              fill.scale = tm_scale_continuous(
                values = "greys"
              ),
              fill.legend = tm_legend(
                title = "Total TRIPS",
                position = tm_pos_in("right", "bottom")
              )
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  
  tm_shape(lisa_sig) +
  tm_polygons(fill = "mean",
              fill.scale = tm_scale_categorical(
                values = c("High-High" = "red",
                          "Low-Low" = "blue",
                          "Low-High" = "lightblue",
                          "High-Low" = "pink")
              ),
              fill.legend = tm_legend(
                title = "LISA Quadrants (p < 0.05)",
                position = tm_pos_in("left", "bottom")
              )
  ) +
  tm_borders(col = "black", lwd = 0.5) +
  tm_title("LISA Map - Weekend/Holiday (11am-8pm)")
```

The LISA map for weekend/holiday trips (11am–8pm) highlights clear clustering patterns in bus demand across Singapore. High-High clusters (red) are concentrated in central, northeastern, and selected western regions, where hexagons with high trip counts are surrounded by other high-demand cells. These hotspots likely reflect popular destinations such as shopping centres, entertainment hubs, and large residential estates that attract strong weekend activity. In contrast, Low-Low clusters (dark blue) are evident in the western and peripheral areas, indicating consistent zones of low demand, often linked to industrial or less connected estates.

A few outliers, including Low-High and High-Low hexagons, appear along cluster edges, showing transitional zones where demand changes abruptly between adjacent areas. Their presence reflects local heterogeneity, where differences in land use or accessibility create contrasting demand within small spatial boundaries.

Overall, the weekend/holiday LISA map confirms that bus trips are not evenly distributed but exhibit strong spatial clustering. Hotspots align with leisure, retail, and dense housing areas, while coldspots dominate less central zones. This pattern underscores the shift in travel behaviour on weekends and holidays, where demand is shaped more by leisure and retail trips than by work-related commuting flows.

## 6.3 Hot Spots and Cold Spots Analysis (HCSA)

### Gi* values and p-values

```{r}
# Compute fixed distance weight matrix
ct <- critical_threshold(st_geometry(weekend))

weekend_fdw <- weekend %>%
  mutate(nb = include_self(
    st_dist_band(
      st_geometry(geometry),
      upper = ct)),
    wt = st_weights(nb,
                    style = "W"),
    .before = 1)

# Compute local Gi*
set.seed(1234)

HCSA_fdw <- weekend_fdw %>%
  mutate(
    gistar = local_gstar_perm(
      TRIPS, nb, wt, nsim = 99),
    .before = 1) %>%
  unnest(gistar)

# Plot Gi* values
Gi_star_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  
  tm_shape(HCSA_fdw) +
  tm_polygons(fill = "gi_star",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 5,
                values = "-brewer.rd_bu"),
              fill.legend = tm_legend(
                title = "local Gi*",
                position = tm_pos_in(
                  "right", "bottom"))) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_title("Local Gi* of TRIPS")

# Plot p-values
p_values_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  
  tm_shape(HCSA_fdw) +
  tm_polygons(fill = "p_sim",
              fill.scale = tm_scale_intervals(
                breaks = c(0, 0.001, 0.01, 0.05, 0.1, 1),
                values = "-brewer.reds"),
              fill.legend = tm_legend(
                title = "p-value",
                position = tm_pos_in("right", "bottom"))) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_title("p-values of local Gi* of TRIPS")

tmap_arrange(Gi_star_map, p_values_map, asp = 1, ncol = 2)
```

The Local Gi* analysis for weekend and holiday trips (left map) reveals clear clustering of demand across Singapore. Strong hotspots (red shades) are concentrated in the central, northeastern, and selected western regions, where trip volumes are significantly higher than in surrounding areas. These likely correspond to major residential estates, recreational hubs, or transport interchanges that attract higher passenger flows during non-working days. In contrast, coldspots (blue shades) appear mainly in the west and northwestern periphery, where consistently lower trip counts suggest weaker demand, possibly reflecting less populated or industrial zones.

The p-value map (right) confirms the statistical robustness of these clusters. Many of the identified hotspots and coldspots overlap with dark red areas (<0.05), meaning their clustering is unlikely to be random. Peripheral areas with lighter shades show weaker or insignificant clustering, indicating more dispersed or inconsistent demand patterns.

Overall, the weekend/holiday maps highlight that trip demand is unevenly distributed but strongly structured around key hubs. Hotspots align with central and residential corridors where leisure and social activities drive bus usage, while coldspots dominate the periphery. This reflects a spatially concentrated but systematic pattern of non-work travel demand

### HCSA Map

```{r}
# HCSA cluster classification
HCSA_fdw <- HCSA_fdw %>%
  mutate(HCSA_cluster = case_when(
    p_sim > 0.05 ~ "Insignificant",
    p_sim <= 0.05 & cluster == "High" ~ "Hot spot",
    p_sim <= 0.05 & cluster == "Low" ~ "Cold spot",
    TRUE ~ "Other"),
    HCSA_cluster = factor(
      HCSA_cluster,
      levels = c("Insignificant", "Hot spot", "Cold spot")
    ),
    .before = 1
  )

# Plot HCSA cluster map
HCSA_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  
  tm_shape(HCSA_fdw) +
  tm_polygons(
    fill = "HCSA_cluster",
    fill.scale = tm_scale_categorical(
      values = c(
        "grey80",  # Insignificant
        "red",     # Hot spot
        "blue"     # Cold spot
      )
    ),
    fill.legend = tm_legend(
      title = "HCSA Cluster",
      position = tm_pos_in("right", "bottom"))
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_title("HCSA Clusters (p < 0.05)")

# Gi* map and HCSA map
tmap_arrange(Gi_star_map, HCSA_map, asp = 1, ncol = 2)
```

On the right map, significant cold spots (blue) are concentrated in the western and northern peripheries, reflecting consistently low demand in less populated or industrial zones. In contrast, hot spots (red) appear in central and northeastern areas, as well as selected pockets in the east, aligning with high-demand hubs such as residential estates, shopping districts, and transport interchanges that attract leisure and family trips during weekends. Grey hexagons dominate much of the map, indicating insignificant clusters where trip patterns are more dispersed and not statistically robust.

When considered with the Local Gi* map on the left, which depicts clustering intensity, the HCSA analysis provides a sharper focus by filtering for only statistically significant hot and cold spots. This combination reveals that while weekend demand is more spatially dispersed compared to weekday peaks, notable hotspots still emerge in urban centres and high-traffic nodes, whereas low-demand clusters are consistently found in peripheral regions. Overall, the results underscore that weekend bus trips are less uniformly tied to commuting patterns but are still structured around key leisure, residential, and transport activity zones.

## 6.4 Emerging Hot Spot Analysis (EHSA)

```{r}
hex_ids <- hexagon$HEX_ID

trips_weekend_hourly <- trips %>%
  filter(DAY_TYPE == "WEEKENDS/HOLIDAY", HOUR_OF_DAY %in% 11:20) %>%
  group_by(HEX_ID, HOUR_OF_DAY) %>%
  summarise(TRIPS = sum(TRIPS), .groups = "drop") %>%
  tidyr::complete(HEX_ID = hex_ids, HOUR_OF_DAY = 11:20,
                  fill = list(TRIPS = 0)) %>%
  arrange(HEX_ID, HOUR_OF_DAY)

trips_weekend_hourly$HEX_ID <- factor(trips_weekend_hourly$HEX_ID, levels = hex_ids)
trips_weekend_hourly$HOUR_OF_DAY <- as.integer(trips_weekend_hourly$HOUR_OF_DAY)

# Build space–time cube
TRIPS_weekend_st <- spacetime(
  trips_weekend_hourly,
  hexagon %>% select(HEX_ID), 
  .loc_col  = "HEX_ID",
  .time_col = "HOUR_OF_DAY"
)

is_spacetime_cube(TRIPS_weekend_st) 
```

```{r}
# Add neighbours and weights
TRIPS_weekend_nb <- TRIPS_weekend_st %>%
  activate("geometry") %>%
  mutate(
    nb = include_self(st_contiguity(geometry)), 
    wt = st_inverse_distance(nb, geometry, scale = 1, alpha = 1),
    .before = 1
  ) %>%
  set_nbs("nb") %>%
  set_wts("wt")

# Compute Gi* 
set.seed(1234)

gi_weekend <- TRIPS_weekend_nb %>%
  group_by(HOUR_OF_DAY) %>%
  mutate(gi_star = local_gstar_perm(TRIPS, nb, wt, nsim = 99)) %>%
  tidyr::unnest(gi_star) %>%
  ungroup()
```

### Mann-Kendall Test

H0: No monotonic trend

H1: Monotonic trend is present

```{r}
mk_hotspots_weekend <- gi_weekend %>%
  st_drop_geometry() %>%
  group_by(HEX_ID) %>%
  summarise(
    mk = if(n() >= 3) list(unclass(Kendall::MannKendall(gi_star))) else NA,
    .groups = "drop"
  ) %>%
  tidyr::unnest_wider(mk) %>%
  rename(tau = tau, p_value = sl)

mk_hotspots_weekend_ranked <- mk_hotspots_weekend %>%
  filter(!is.na(p_value)) %>%
  arrange(p_value, desc(abs(tau)))        

mk_hotspots_weekend_ranked %>% slice(1:10)
```

With reference to the results, we will reject the hypothesis null and infer that there is a downward trend.

The Hex H0166 is selected to view the Gi* Trend for afternoon peak:

```{r}
hex_trend <- gi_weekend %>%
  st_drop_geometry() %>%
  filter(HEX_ID == "H0166") %>%
  select(HEX_ID, HOUR_OF_DAY, gi_star)

ggplot(data = hex_trend, aes(x = HOUR_OF_DAY, y = gi_star)) +
  geom_line(color = "black") +
  labs(
    title = "Gi* Trend across Weekend/Holiday Hours — H0166",
    x = "Hour of Day",
    y = "Gi*"
  ) +
  theme_light()
```

The plot for Hex H0166 shows a clear downward trend in the Gi* statistic. At the start of the period (around 11am–12pm), Gi* values are high and positive, close to 2.0, indicating that this location is initially part of a strong hotspot with trip counts significantly higher than its surrounding areas. However, from midday onward, the Gi* values steadily decline, first stabilising between 1.0 and 0.8 around 2–4pm, before continuing to drop sharply after 5pm. By 8pm, the Gi* value reaches close to zero, suggesting the hotspot effect diminishes entirely.

This declining trend indicates a monotonic decrease in spatial clustering strength at this location over the course of the day. The area starts as a robust hotspot during midday, likely linked to leisure, shopping, or recreational travel demand, but gradually loses its clustering significance into the evening. In practical terms, this suggests that bus trip intensity at this location aligns less with high-demand neighbours later in the day, reflecting a dispersal of travel patterns typical of weekends and holidays when activities become more spread across multiple destinations rather than concentrated in a single area.

### Performing Emerging Hotspot Analysis

```{r}
set.seed(1234)

ehsa_weekend <- emerging_hotspot_analysis(
  x     = TRIPS_weekend_nb,   
  .var  = "TRIPS",     
  k     = 1,      
  nsim  = 99   
)

ehsa_weekend_sf <- hexagon %>%
  select(HEX_ID) %>%
  left_join(ehsa_weekend, by = c("HEX_ID" = "location")) %>%
  filter(p_value < 0.05) %>%
  st_as_sf()
```


```{r}
ehsa_weekend$classification   <- factor(ehsa_weekend$classification,   levels = ehsa_levels)
ehsa_weekend_sf$classification <- factor(ehsa_weekend_sf$classification, levels = ehsa_levels)
```


```{r}
ggplot(data = ehsa_weekend, 
       aes(x = classification, fill = classification)) +
  geom_bar(color = "black", alpha = 0.8) +
  scale_fill_manual(values = ehsa_colors, drop = FALSE) +
  labs(
    title = "Distribution of EHSA Classes – Weekend/Holiday (11am–8pm)",
    x = "EHSA Classification", 
    y = "Number of Hexagons"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", size = 14)
  )
```

Most hexagons fall under no pattern detected, suggesting that demand clustering is absent across much of the city during these hours. However, a notable share of hexagons are identified as persistent coldspots and sporadic coldspots, indicating consistent or irregularly low trip counts. Sporadic hotspots also appear in meaningful numbers, pointing to localized and irregular bursts of higher demand that may be tied to leisure activities, shopping trips, or recreational destinations. Meanwhile, persistent hotspots are few, and consecutive hotspots are almost absent, showing that stable high-demand clusters are rare during weekends and holidays.

Overall, the distribution suggests a spatially uneven pattern where low-demand clusters dominate, but isolated hotspots emerge in specific areas. This aligns with the idea that weekend and holiday travel is less structured by daily commuting flows and more influenced by discretionary trips, creating irregular surges in activity at certain nodes while much of the network remains stable at low demand.

```{r}
tmap_mode("plot")
tm_shape(mpsz) + tm_polygons(col = "grey90", border.col = "white") +
tm_shape(ehsa_weekend_sf) +
  tm_polygons(
    fill = "classification",
    fill.scale = tm_scale_categorical(values = ehsa_colors), # use custom palette
    fill.legend = tm_legend(
      title = "EHSA Classification",
      position = tm_pos_in("right", "bottom"))
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_title("Emerging Hot Spot Analysis (EHSA) – Weekend/Holiday (11am–8pm)")

```

The EHSA map for weekends and holidays shows a mixed spatial pattern of hotspots and coldspots across Singapore. Consecutive hotspots (red) are concentrated in the central region and along some transport corridors, indicating areas of consistently high demand. Sporadic hotspots (peach) appear across several dispersed locations, reflecting irregular surges in bus activity that may correspond to event-based or recreational travel.

On the other hand, consecutive and persistent coldspots (dark blue) dominate the western and northern peripheries, suggesting stable zones of consistently low bus demand, particularly in less connected or industrial areas. New coldspots (teal) are also scattered, highlighting emerging low-demand locations during weekends. Grey areas classified as “no pattern detected” cover a large share of the map, signifying that many regions do not show statistically strong or consistent clustering.

Overall, the map suggests that weekend and holiday demand is more dispersed compared to weekday peaks, with central and activity-rich areas sustaining stronger hotspots, while peripheral estates remain coldspots. This highlights the recreational and leisure-driven nature of weekend travel, where demand is concentrated in specific urban nodes rather than broadly distributed across the network.