---
title: "Take home exercise 2"
author: "Tai Qiu Yan"
date: "October 8, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# 1 Introduction

This report explores the space–time structure of bus trip demand in Singapore using stop-level OD data aggregated to a 750 m hexagonal grid. I integrate geospatial datasets (bus stops and MP19 subzones) with tidy workflows, then build distance-band neighbourhoods and row-standardised weights to study spatial dependence. Descriptive maps and distributions establish baseline patterns for three periods: weekday morning, weekday afternoon, and weekend/holiday. I then apply Local Moran’s I (LISA), the Getis–Ord Gi* statistic, Hot/Cold Spot Analysis (HCSA), and Emerging Hot Spot Analysis (EHSA), together with the Mann–Kendall test, to detect clusters, outliers, and temporal trends in demand.

# 2 Loading the R package

In this exercise, the following packages are used:

tidyverse – A collection of data science packages (readr, dplyr, tidyr, ggplot2, etc.) with a unified grammar for importing, wrangling, and visualising data.

sf – Simple Features for R: create, read, write, transform, and analyse vector geospatial data (points/lines/polygons) with modern CRS support.

sfdep – Provides functions for spatial dependence analysis, including tools for computing spatial weights, spatial autocorrelation measures, and emerging hotspot analysis, extending the functionality of the sf package.

tmap – A flexible thematic mapping package that supports both static and interactive maps.

knitr – An engine for dynamic report generation, allowing integration of R code and results into documents.

kableExtra – Extends knitr::kable() by providing advanced table styling features. It allows users to create publication-quality tables in HTML and LaTeX.

DT – An R interface to the DataTables JavaScript library, enabling interactive tables with features such as searching, filtering, sorting, and pagination.

Kendall – Implements the Mann-Kendall trend test and related functions for detecting monotonic trends in time series data.

scales – Provides tools for scaling data and controlling axis, legend, and color breaks in visualisations.

patchwork - A lightweight toolkit for composing multiple ggplot2 plots into a single layout using “+” syntax; supports flexible grids, nested layouts, shared guides, and plot annotations/titles.

cowplot – A ggplot2 extension focused on publication-ready figures; provides functions like plot_grid() and ggdraw() to arrange plots, add titles/labels, and fine-tune themes with minimal whitespace.

```{r}
pacman::p_load(tidyverse, sf, sfdep, tmap, knitr, kableExtra, DT, Kendall, scales, patchwork, cowplot)
```

# 3 Data Importing and Wrangling

## Importing Data

The following datasets are imported:

- Bus Stop Location shapefile downloaded from LTA DataMall

- Master Plan 2019 Subzone Boundary (No Sea) from Singapore’s open data portal

- Passenger Volume by Origin Destination Bus Stops downloaded from LTA DataMall

```{r}
#Bus Stop Location
BusStop = st_read(dsn = "data/LTADataMall/", 
                  layer = "BusStop") %>%
  st_transform(crs = 3414)

#Master Plan 2019 Subzone
mpsz = read_rds("data/rds/mpsz_sf.rds")

#O-D Bus Trips
odbus <- read_csv("data/LTADataMall/origin_destination_bus_202508.csv")
```

## Extracting Bus Stops located within Singapore

The code below assigns each bus stop to the planning subzone it belongs to, creating a dataset of bus stops (BusStop_in_SG) linked to their respective subzones.

```{r}
BusStop_in_SG <- st_join(
  BusStop, mpsz, 
  join = st_within, 
  left = FALSE)
```

### Visualize

```{r}
tm_shape(mpsz) +
  tm_polygons() +
tm_shape(BusStop_in_SG) +
  tm_dots()
```

## Deriving Analytical Hexagon

The code below generates a hexagonal tessellation over Singapore’s planning subzones to standardize spatial units for analysis.

```{r}
hexagon <- st_make_grid(mpsz, 
                        cellsize = 750,
                        what = "polygon",
                        square = FALSE) %>%
  st_sf()
```

## Selecting hexagons with bus stops

The code below links bus stops to the analytical hexagons, counts how many fall inside each, and filters the grid to include only meaningful hexagons with bus stops.

```{r}
hexagon$busstop_count =
  lengths(st_intersects(
    hexagon, BusStop_in_SG))

hexagon <- filter(
  hexagon, busstop_count > 0)
```

## Assigning ids to each hexagon

The code below cleans the hexagon dataset by dropping the temporary count column and assigns each hexagon a unique identifier (HEX_ID) for consistent reference in spatial analysis.

```{r}
hexagon <- hexagon %>% 
  select(, -busstop_count)

hexagon$HEX_ID <- sprintf(
  "H%04d", seq_len(
    nrow(hexagon))) %>% 
  as.factor()
```

## Preparing Trip Generation Data

The code below extracts the relevant bus trip information, standardizes the column names into intuitive labels (BUS_STOP_N, HOUR_OF_DAY, TRIPS), and prepares the bus stop IDs as categorical values for later aggregation and spatial analysis.

```{r}
trips <- odbus %>%
  select(c(ORIGIN_PT_CODE, DAY_TYPE, TIME_PER_HOUR, TOTAL_TRIPS)) %>%
  rename(BUS_STOP_N = ORIGIN_PT_CODE) %>%
  rename(HOUR_OF_DAY = TIME_PER_HOUR) %>%
  rename(TRIPS = TOTAL_TRIPS)
trips$BUS_STOP_N <- as.factor(trips$BUS_STOP_N)
```

## Populating Hexagon IDs into BusStop data frame

The code below creates a lookup table that links each bus stop to the hexagon it belongs to. It is essential for aggregating bus trips at the hexagon level later on.

```{r}
bs_hex <- st_intersection(
  BusStop, hexagon) %>%
  st_drop_geometry() %>%
  select(c(BUS_STOP_N, HEX_ID))
```

## Adding HEX_ID into bus trips data

The code below enriches the trip dataset by adding hexagon IDs to each bus trip record, enabling aggregation and spatial analysis of trips at the hexagon level.

```{r}
trips <- inner_join(trips, bs_hex)
```

## Aggregating TRIPS based on HEX_ID

The code below converts raw trip records into spatial-temporal summaries, making the data easier to analyze by hexagon location, type of day, and hour of day.

```{r}
trips <- trips %>%
  group_by(
    HEX_ID,
    DAY_TYPE,
    HOUR_OF_DAY) %>%
  summarise(TRIPS = sum(TRIPS))
```

## Building Distance-Based Spatial Weights

The code below sets up spatial neighbors and weights using a critical distance, then checks neighbor statistics to ensure the spatial structure is valid.

```{r}
# Calculate critical distance for all hexagons with bus stops
critical_dist <- critical_threshold(st_geometry(hexagon))

# Create distance-based neighbors using the critical distance
nb <- st_dist_band(
  st_geometry(hexagon),
  upper = critical_dist)

# Check the number of neighbors for each hexagon
n_neighbors <- lengths(nb)

# Create spatial weights
wt <- st_weights(nb, style = "W", allow_zero = TRUE)

cat("Critical distance:", critical_dist, "meters\n")
cat("Total hexagons:", nrow(hexagon), "\n")
cat("Hexagons with neighbors:", sum(n_neighbors > 0), "\n")
cat("Min neighbors:", min(n_neighbors), "\n")
cat("Max neighbors:", max(n_neighbors), "\n")
```


# 4 Weekday morning 5 - 10 am

## 4.1 Overview

### Choropleth map of total trips

```{r}
weekday_morning <- trips %>%
  filter(DAY_TYPE == "WEEKDAY", HOUR_OF_DAY %in% 6:9) %>%
  group_by(HEX_ID) %>%
  summarise(TRIPS = sum(TRIPS), .groups = "drop") %>%
  left_join(hexagon, by = "HEX_ID") %>%
  st_as_sf()

# Aggregate mpsz by PLN_AREA_N to get boundaries
mpsz_area <- mpsz %>%
  group_by(PLN_AREA_N) %>%
  summarise(geometry = st_union(geometry))

tmap_mode("plot")

tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  
  tm_shape(weekday_morning) +
  tm_polygons(
    fill = "TRIPS",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      n = 5,
      values = "viridis"
    ),
    fill.legend = tm_legend(
      title = "Total Trips\n(Weekday Morning)",
      position = tm_pos_in("right", "bottom")
    )
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_shape(mpsz_area) +
  tm_borders(col = "black", lwd = 0.2) +
  tm_title("Distribution of Bus Trips - Weekday Morning Peak (6-9am)") +
  tm_compass(position = tm_pos_in("left", "top")) +
  tm_scale_bar(position = tm_pos_in("left", "bottom"))
```

The visualisation shows that bus trips during the weekday morning peak (6–9am) are heavily concentrated in large residential towns such as Tampines, Pasir Ris, Yishun, and Woodlands, as well as the western corridor around Jurong East and Clementi. Moderate activity is seen in areas like Hougang, Sengkang, and Punggol, while peripheral industrial and coastal zones record low trip counts. Overall, the map highlights a polycentric commuting pattern, where dense housing estates generate significant trips feeding into central employment areas, reinforcing the key role of buses in connecting residential towns with work and study destinations.

### Trip Distribution

```{r}
# Filter out zero trips
weekday_morning_filtered <- weekday_morning %>% filter(TRIPS > 0)

# Calculate statistics for filtered data
mean_trips <- mean(weekday_morning_filtered$TRIPS)
median_trips <- median(weekday_morning_filtered$TRIPS)
q95 <- quantile(weekday_morning_filtered$TRIPS, 0.95)

# Plot 1: Original scale histogram (cut off at 95th percentile)
x_limit <- q95
p1 <- ggplot(weekday_morning_filtered, aes(x = TRIPS)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(aes(xintercept = mean_trips), 
             color = "red", linetype = "dashed", linewidth = 0.8) +
  geom_vline(aes(xintercept = median_trips), 
             color = "orange", linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = mean_trips, y = Inf, 
           label = paste0("Mean: ", format(round(mean_trips, 0), big.mark = ",")), 
           vjust = 1.5, hjust = -0.1, color = "red", size = 3.5, fontface = "bold") +
  annotate("text", x = median_trips, y = Inf, 
           label = paste0("Median: ", format(round(median_trips, 0), big.mark = ",")), 
           vjust = 3, hjust = -0.1, color = "orange", size = 3.5, fontface = "bold") +
  scale_x_continuous(
    labels = label_number(scale = 1/1000, suffix = "k"),
    breaks = seq(0, x_limit, by = 20000),
    limits = c(0, x_limit),
    expand = expansion(mult = c(0.02, 0.05))
  ) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "0-95th Percentile",
    x = "Number of Trips",
    y = "Frequency",
    caption = paste0("95th percentile: ", format(round(q95, 0), big.mark = ","), " trips")
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
    axis.title = element_text(size = 10),
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 9),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = "gray90"),
    plot.margin = margin(10, 5, 10, 10),
    plot.caption = element_text(size = 8, color = "gray50", hjust = 1, face = "italic")
  )

# Plot 2: Bar chart for 95th-100th percentile (top 5%)
top5_data <- weekday_morning_filtered %>% filter(TRIPS >= q95)

p2 <- ggplot(top5_data, aes(x = TRIPS)) +
  geom_histogram(bins = 25, fill = "coral", color = "white", alpha = 0.8) +
  scale_x_continuous(
    labels = label_number(scale = 1/1000, suffix = "k"),
    expand = expansion(mult = c(0.02, 0.05))
  ) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Top 5%",
    x = "Number of Trips",
    y = "Frequency",
    caption = paste0("n = ", nrow(top5_data), " hexagons | Max: ", 
                    format(max(top5_data$TRIPS), big.mark = ","))
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
    axis.title = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = "gray90"),
    plot.margin = margin(10, 5, 10, 5),
    plot.caption = element_text(size = 8, color = "gray50", hjust = 0.5)
  )

# Plot 3: Log transformation
p3 <- ggplot(weekday_morning_filtered, aes(x = log10(TRIPS))) +
  geom_histogram(bins = 50, fill = "darkgreen", color = "white", alpha = 0.8) +
  scale_x_continuous(
    breaks = 0:5,
    labels = c("1", "10", "100", "1k", "10k", "100k"),
    expand = expansion(mult = c(0.02, 0.05))
  ) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Log10 Transformed",
    x = "Number of Trips (Log Scale)",
    y = "Frequency",
    caption = "Distribution shape across all magnitudes"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
    axis.title = element_text(size = 10),
    axis.text.x = element_text(size = 9),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = "gray90"),
    plot.margin = margin(10, 10, 10, 5),
    plot.caption = element_text(size = 8, color = "gray50", hjust = 0.5)
  )

# Combine plots with adjusted widths
combined_plot <- p1 + p2 + p3 + 
  plot_layout(widths = c(2.5, 1.5, 2.5)) +
  plot_annotation(
    title = "Distribution of Trip Counts - Weekday Morning Peak (6-9am)",
    subtitle = paste0("n = ", format(nrow(weekday_morning_filtered), big.mark = ","), 
                     " hexagons (zeros excluded)"),
    theme = theme(
      plot.title = element_text(face = "bold", size = 15, hjust = 0.5),
      plot.subtitle = element_text(size = 11, color = "gray40", hjust = 0.5)
    )
  )

combined_plot
```

This three-panel visualization reveals a highly right-skewed distribution of weekday morning peak trips (6am-9am) across 816 hexagonal zones in the study area. The main distribution (0-95th percentile) shows that the majority of hexagons have relatively low trip counts, with a median of 14,934 trips and mean of 30,884 trips. The substantial gap between median and mean indicates strong positive skew, suggesting most areas experience moderate traffic while a few zones generate disproportionately high trip volumes. The distribution shows a sharp peak near zero with a long tail extending rightward.

The top 5% panel highlights the extreme values in the distribution, where only 41 hexagons exceed the 95th percentile threshold of 110,640 trips, with the maximum reaching 396,146 trips. These outliers likely represent major transportation hubs, central business districts, or key transit corridors that concentrate morning commute activity. 

The log-transformed view provides additional insight by revealing a more normalized, approximately bell-shaped distribution. This transformation suggests the underlying data may follow a log-normal distribution, which is typical for spatial phenomena where multiplicative processes drive variation. The log scale demonstrates that trip generation patterns span multiple orders of magnitude, reflecting the diverse nature of urban zones ranging from quiet residential neighborhoods to major activity centers and employment hubs.

## 4.2 Local Indicators of Spatial Association(LISA)

The code below computes Local Moran’s I for TRIPS in the weekday-morning hexagons (6–9am), using a distance-band neighbor list at 1984.313 meters.

```{r}
nb <- st_dist_band(st_geometry(weekday_morning), upper = critical_dist)
wt <- st_weights(nb, style = "W", allow_zero = TRUE)

set.seed(1234)
lisa_morning <- weekday_morning %>%
  mutate(
    local_moran = local_moran(
      TRIPS, nb, wt, nsim = 99), 
    .before = 1) %>%
  unnest(local_moran)
```

### Local Moran’s I values

```{r}
# Filter for significant Local Moran's I (p-value < 0.1)
lisa_morning_sig <- lisa_morning %>%
  filter(p_ii < 0.1)

# Create the Local Moran's I map
ii_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  tm_shape(lisa_morning_sig) +
  tm_polygons(fill = "ii",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 5,
                values = "brewer.RdBu"
              ),
              fill.legend = tm_legend(
                title = "Local Moran's I",
                position = tm_pos_in("right", "bottom"),
                title.size = 0.8,
                text.size = 0.7
              )
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_shape(mpsz_area) +
  tm_borders(col = "grey30", lwd = 0.2) +
  tm_title("Local Moran's I - Morning Peak (p-value < 0.1)") +
  tm_compass(position = tm_pos_in("left", "top")) +
  tm_scale_bar(position = tm_pos_in("left", "bottom"))

# Display the map
ii_map
```

The Local Moran's I map reveals significant spatial clustering patterns of bus trip generation during weekday afternoon peak hours (5pm-8pm), with only statistically significant results (p < 0.1) displayed. Positive spatial autocorrelation (blue hexagons, values 0.5-1.5) appears more scattered compared to morning patterns, concentrated primarily in western, southwestern, and selected eastern planning areas. These high-high clusters indicate areas where high trip volumes are surrounded by similarly high-activity neighbors, representing residential zones receiving concentrated return commuter flows during evening peak hours.

Moderate positive clusters (light blue, 0-0.5) appear dispersed across various planning areas, suggesting more fragmented evening clustering patterns than morning peaks. Notably, negative spatial autocorrelation (orange to red, -1.0 to 0) appears extensively throughout the island, particularly in central, northern, and eastern regions. These represent spatial outliers—either high-activity hexagons surrounded by low-activity neighbors (high-low outliers) or vice versa (low-high outliers)—indicating transitional boundaries, edge effects between different land uses, or isolated activity centers.

The afternoon pattern differs markedly from morning peaks, showing more dispersed clustering as commuters return from centralized employment hubs to diverse residential locations. The fragmented spatial structure reflects Singapore's polycentric urban form and varied evening travel patterns, with some residential areas forming return-commute clusters while others exhibit irregular demand. 

### Moran scatterplot with standardised variable

```{r fig.width=10, fig.height=6}
lisa_morning <- lisa_morning %>%
  mutate(lag_TRIPS = st_lag(
    TRIPS, nb, wt),
    .before = 1) %>%
  unnest(lag_TRIPS)

lisa_morning <- lisa_morning %>%
  mutate(z_TRIPS = scale(TRIPS),
         z_lag_TRIPS = scale(lag_TRIPS),
         .before = 1)

# Create the Moran scatterplot
moran_plot <- ggplot(data = lisa_morning,
       aes(x = z_TRIPS,
           y = z_lag_TRIPS,
           color = mean)) +
  geom_point(size = 2, alpha = 0.6) +
  geom_smooth(method = "lm",
              se = FALSE,
              color = "black") +
  geom_hline(yintercept = mean(lisa_morning$z_lag_TRIPS), lty = 2) +
  geom_vline(xintercept = mean(lisa_morning$z_TRIPS), lty = 2) +
  scale_color_manual(
    values = c("High-High" = "red",
               "Low-Low" = "blue",
               "Low-High" = "lightblue",
               "High-Low" = "pink")) +
  labs(x = "Standardised TRIPS",
       y = "Standardised Spatial Lag of TRIPS",
       title = "Standardised Moran Scatterplot with LISA Quadrants") +
  theme_minimal() +
  theme(legend.position = "bottom")

# Create the bar chart for quadrant counts
quadrant_counts <- lisa_morning %>%
  st_drop_geometry() %>%
  count(mean) %>%
  mutate(mean = factor(mean, levels = c("High-High", "Low-High", 
                                         "Low-Low", "High-Low")))

bar_plot <- ggplot(quadrant_counts, aes(x = mean, y = n, fill = mean)) +
  geom_bar(stat = "identity", alpha = 0.8) +
  geom_text(aes(label = n), vjust = -0.5, size = 4, fontface = "bold") +
  scale_fill_manual(
    values = c("High-High" = "red",
               "Low-Low" = "blue",
               "Low-High" = "lightblue",
               "High-Low" = "pink")) +
  labs(x = "LISA Quadrant",
       y = "Count of Hexagons",
       title = "Count by Quadrant") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

# Combine plots side by side
combined_plot <- moran_plot + bar_plot + 
  plot_layout(widths = c(2, 1)) +
  plot_annotation(
    title = "Local Moran's I Analysis - Weekday Morning Peak",
    theme = theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5))
  )

combined_plot
```

The Moran scatterplot reveals distinct spatial patterns in weekday morning bus trips, with the bar chart quantifying the distribution across LISA quadrants. The Low-Low quadrant (dark blue, 346 hexagons) dominates, representing cold spots where low-trip hexagons cluster together in peripheral or less populated regions, exhibiting strong positive spatial autocorrelation. The High-High quadrant (red, 239 hexagons) shows significant hot spots—high-trip hexagons surrounded by similarly high-trip neighbors in dense residential zones with strong morning commuting demand, forming positive spatial clusters.

The Low-High quadrant (light blue, 178 hexagons) represents underperforming hexagons within busier regions, indicating negative spatial autocorrelation where low-activity areas are surrounded by high-activity neighbors. The High-Low quadrant (pink, 53 hexagons) contains the fewest outliers—isolated high-trip hexagons surrounded by low-trip areas, suggesting unique transit hubs or edge locations between different activity zones.

The positive slope of the regression line confirms overall positive spatial autocorrelation, meaning similar trip intensities systematically cluster together. The dominance of Low-Low and High-High quadrants (585 of 816 hexagons, 72%) validates strong spatial dependence in morning trip patterns, with clear segregation between high-activity residential/employment zones and low-activity peripheral areas, while spatial outliers remain minimal.

### LISA Map

```{r fig.width=12, fig.height=5}
# Count significant quadrants
quadrant_counts <- lisa_morning_sig %>%
  st_drop_geometry() %>%
  count(mean) %>%
  mutate(mean = factor(mean, levels = c("High-High", "Low-High", 
                                         "Low-Low", "High-Low")))

# Bar chart
bar_chart <- ggplot(quadrant_counts, aes(x = mean, y = n, fill = mean)) +
  geom_bar(stat = "identity", width = 0.7) +
  geom_text(aes(label = n), vjust = -0.3, size = 5, fontface = "bold") +
  scale_fill_manual(
    values = c("High-High" = "red",
               "Low-Low" = "blue",
               "Low-High" = "lightblue",
               "High-Low" = "pink")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(x = "LISA Quadrant", 
       y = "Count of Hexagons", 
       title = "LISA Quadrant Counts") +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    axis.text.y = element_text(size = 11),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 13, face = "bold", hjust = 0.5),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.margin = margin(10, 10, 10, 10)
  )

# LISA map
tmap_mode("plot")
lisa_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  
  tm_shape(lisa_morning_sig) +
  tm_polygons(fill = "mean",
              fill.scale = tm_scale_categorical(
                values = c("High-High" = "red",
                          "Low-Low" = "blue",
                          "Low-High" = "lightblue",
                          "High-Low" = "pink")
              ),
              fill.legend = tm_legend(
                title = "LISA Quadrants",
                position = tm_pos_in("right", "bottom")
              )
  ) +
  tm_borders(col = "black", lwd = 0.5) +
  tm_shape(mpsz_area) +
  tm_borders(col = "grey30", lwd = 1.5) +
  tm_layout(
    main.title = "LISA Map (p-value < 0.1)",
    main.title.size = 1.2,
    main.title.fontface = "bold",
    legend.text.size = 0.8,
    legend.title.size = 1,
    inner.margins = c(0.05, 0.05, 0.05, 0.05)
  )

# Convert tmap to grob
lisa_map_grob <- tmap_grob(lisa_map)

# Combine
combined_plot <- plot_grid(
  lisa_map_grob, 
  bar_chart, 
  ncol = 2, 
  rel_widths = c(2,1.5),
  labels = NULL
)

title <- ggdraw() + 
  draw_label(
    "Local Indicators of Spatial Association (LISA) - Weekday Morning Peak",
    fontface = 'bold',
    size = 15,
    x = 0.5,
    hjust = 0.5
  )

final_plot <- plot_grid(
  title, 
  combined_plot,
  ncol = 1,
  rel_heights = c(0.06, 0.94)
)

final_plot
```

The LISA map reveals significant spatial clustering of weekday morning bus trips across Singapore. Low-Low clusters (dark blue, 150 hexagons) slightly dominate, appearing in peripheral, southern, and eastern regions where low trip volumes are surrounded by similarly low-activity neighbors, reflecting weaker demand in less densely populated or non-residential zones. High-High clusters (red, 145 hexagons) are nearly equal in number, concentrated across northern, central-eastern, and western planning areas, representing dense residential zones with concentrated morning commuter demand.

Low-High outliers (light blue, 78 hexagons) indicate underperforming hexagons within high-activity areas, typically at cluster edges or transitional zones between different land uses. High-Low outliers (pink, 8 hexagons) are minimal but mark isolated high-demand locations like transit hubs or interchanges surrounded by low-activity areas.

The near-equal distribution of High-High and Low-Low clusters confirms strong positive spatial autocorrelation and clear spatial segregation between high-demand residential zones and low-demand peripheral areas. The pattern aligns with Singapore's polycentric urban structure, where established residential planning areas generate morning trip hotspots feeding commuters toward employment centers, while outliers mark transitional boundaries and strategic transit nodes. Overall, bus trips exhibit systematic spatial clustering rather than random distribution.

## 4.3 Hot Spots and Cold Spots Analysis (HCSA)

The code below builds a fixed distance–band neighbourhood for each weekday-morning hexagon, creates row-standardised spatial weights, and then computes the Local Getis-Ord Gi* statistic for TRIPS.

```{r}
weekday_morning_fdw <- weekday_morning %>%
  mutate(nb = include_self(
    st_dist_band(
      st_geometry(geometry),
      upper = critical_dist)),
    wt = st_weights(nb,
                    style = "W"),
    .before = 1)

# Compute local Gi*
set.seed(1234)

HCSA_fdw <- weekday_morning_fdw %>%
  mutate(
    gistar = local_gstar_perm(
      TRIPS, nb, wt, nsim = 99),
    .before = 1) %>%
  unnest(gistar)
```


```{r}
# Filter for significant Gi* values (p < 0.1)
HCSA_sig <- HCSA_fdw %>%
  filter(p_sim < 0.1)

# Plot Gi* values for significant hexagons only
Gi_star_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  
  tm_shape(HCSA_sig) +
  tm_polygons(fill = "gi_star",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 5,
                values = "-brewer.rd_bu"),
              fill.legend = tm_legend(
                title = "local Gi*",
                position = tm_pos_in("right", "bottom")
              )
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_shape(mpsz_area) +
  tm_borders(col = "grey30", lwd = 0.2) +
  tm_title("Local Gi* of TRIPS (p-value < 0.1)") +
  tm_compass(position = tm_pos_in("left", "top")) +
  tm_scale_bar(position = tm_pos_in("left", "bottom"))

# HCSA cluster classification with p < 0.1
HCSA_fdw <- HCSA_fdw %>%
  mutate(HCSA_cluster = case_when(
    p_sim <= 0.1 & cluster == "High" ~ "Hot spot",
    p_sim <= 0.1 & cluster == "Low" ~ "Cold spot",
    TRUE ~ "Other"),
    HCSA_cluster = factor(
      HCSA_cluster,
      levels = c("Hot spot", "Cold spot")  # Reordered levels
    ),
    .before = 1
  )

# Filter to keep only significant hexagons
HCSA_sig <- HCSA_fdw %>%
  filter(HCSA_cluster != "Insignificant")

# Aggregate mpsz by PLN_AREA_N to get planning area boundaries
mpsz_planning <- mpsz %>%
  group_by(PLN_AREA_N) %>%
  summarise(geometry = st_union(geometry))

# HCSA cluster map
HCSA_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  
  tm_shape(HCSA_sig) +
  tm_polygons(
    fill = "HCSA_cluster",
    fill.scale = tm_scale_categorical(
      values = c(
        "Hot spot" = "red",
        "Cold spot" = "blue"
      )
    ),
    fill.legend = tm_legend(
      title = "HCSA Cluster",
      position = tm_pos_in("right", "bottom"))
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_shape(mpsz_area) +
  tm_borders(col = "grey30", lwd = 0.2) +
  tm_title("HCSA Clusters (p-value < 0.1)")

tmap_arrange(Gi_star_map, HCSA_map, asp = 1, ncol = 2)
```

The Local Gi* map identifies scattered hot spots (red/orange) across northern, central, and eastern regions indicating high trip clustering intensity, while cold spots (blue) dominate western and peripheral areas with consistently low trips. Higher absolute Gi* values indicate stronger concentration, though the pattern shows localized variations rather than contiguous clusters.

The HCSA Clusters map reveals statistically significant spatial clustering of weekday morning bus trips (p < 0.1). Hot spots (red) form extensive contiguous clusters across northern, central-eastern, and western planning areas, representing dense residential zones generating concentrated morning commuter flows from established housing estates with strong transit connectivity. Cold spots (blue) appear across western, northwestern, southern, and peripheral regions, reflecting lower residential density, industrial land uses, or limited bus connectivity. These form contiguous clusters confirming genuine spatial patterns of reduced transit demand in non-residential areas.

Together, the maps demonstrate clear spatial segregation in morning bus trip generation. The Gi* map shows clustering intensity with localized variations, while the HCSA map emphasizes broader, statistically robust spatial patterns. Both confirm strong spatial autocorrelation, with high-demand residential zones geographically distinct from low-demand peripheral areas, and transitional boundaries revealing spatial heterogeneity where different land use functions meet.

## 4.4 Emerging Hot Spot Analysis (EHSA) - 5 to 10 am

The code below prepares weekday morning (5–10h) trips for Emerging Hot Spot/space–time analysis. It aggregates trips by hexagon × hour, then ensure every hexagon has all hours (filling missing cells with 0) and orders IDs/time consistently. It converts IDs and hours to the right types, then builds a space–time cube (spacetime) by linking the hourly table to the hexagon geometries, and finally checks that the result is a valid space–time cube.

```{r}
hex_ids <- hexagon$HEX_ID

trips_morning_hourly <- trips %>%
  filter(DAY_TYPE == "WEEKDAY", HOUR_OF_DAY %in% 5:10) %>%
  group_by(HEX_ID, HOUR_OF_DAY) %>%
  summarise(TRIPS = sum(TRIPS), .groups = "drop") %>%
  tidyr::complete(HEX_ID = hex_ids, HOUR_OF_DAY = 5:10,
                  fill = list(TRIPS = 0)) %>%
  arrange(HEX_ID, HOUR_OF_DAY)

trips_morning_hourly$HEX_ID <- factor(trips_morning_hourly$HEX_ID, levels = hex_ids)
trips_morning_hourly$HOUR_OF_DAY <- as.integer(trips_morning_hourly$HOUR_OF_DAY)

# Build space–time cube
TRIPS_morning_st <- spacetime(
  trips_morning_hourly,
  hexagon %>% select(HEX_ID), 
  .loc_col  = "HEX_ID",
  .time_col = "HOUR_OF_DAY"
)

is_spacetime_cube(TRIPS_morning_st) 
```
The code below adds spatial neighbours/weights to the morning space–time cube and computes hourly Local Getis-Ord Gi* for TRIPS. It then flattens the results so each hexagon–hour has a Gi* score and significance.

```{r}
# Add neighbours and weights
TRIPS_morning_nb <- TRIPS_morning_st %>%
  activate("geometry") %>%
  mutate(
    nb = include_self(st_contiguity(geometry)), 
    wt = st_inverse_distance(nb, geometry, scale = 1, alpha = 1),
    .before = 1
  ) %>%
  set_nbs("nb") %>%
  set_wts("wt")

# Compute Gi* 
set.seed(1234)

gi_morning <- TRIPS_morning_nb %>%
  group_by(HOUR_OF_DAY) %>%
  mutate(gi_star = local_gstar_perm(TRIPS, nb, wt, nsim = 99)) %>%
  tidyr::unnest(gi_star) %>%
  ungroup()
```

### Mann-Kendall Test

The code below tests whether each hexagon’s Gi* values show a monotonic time trend (consistently increasing or decreasing) across the selected hours.

H0: No monotonic trend

H1: Monotonic trend is present

alpha: 0.1

```{r}
mk_hotspots <- gi_morning %>%
  st_drop_geometry() %>%
  group_by(HEX_ID) %>%
  summarise(
    mk = if(n() >= 3) list(unclass(Kendall::MannKendall(gi_star))) else NA,
    .groups = "drop"
  ) %>%
  tidyr::unnest_wider(mk) %>%
  rename(tau = tau, p_value = sl)

mk_hotspots_ranked <- mk_hotspots %>%
  filter(!is.na(p_value)) %>%                
  arrange(p_value, desc(abs(tau)))        

mk_hotspots_ranked %>% slice(1:10)
```

With reference to the results, as p-value < alpha, we will reject the null hypothesis and infer that there is a significant trend.

### Performing Emerging Hotspot Analysis

The code below runs EHSA on the weekday–morning space-time cube to classify each hexagon into temporal hotspot/coldspot types (e.g., persistent, consecutive, new, sporadic).

```{r}
set.seed(1234)

ehsa_morning <- emerging_hotspot_analysis(
  x     = TRIPS_morning_nb,   
  .var  = "TRIPS",     
  k     = 1,      
  nsim  = 99   
)
```

The code below attaches EHSA results back to the hexagon layer and keep only statistically significant cells for mapping.

```{r}
ehsa_morning_sf <- hexagon %>%
  select(HEX_ID) %>%
  left_join(ehsa_morning, by = c("HEX_ID" = "location")) %>%
  filter(p_value < 0.1) %>%
  st_as_sf()
```

The code below sets a consistent legend order and color scheme for EHSA classes and apply them to the data before plotting.

```{r}
ehsa_levels <- c(
  "consecutive coldspot", "consecutive hotspot",
  "new coldspot", "new hotspot",
  "no pattern detected",
  "persistent coldspot", "persistent hotspot",
  "sporadic coldspot", "sporadic hotspot"
)

ehsa_colors <- c(
  "consecutive coldspot" = "#4a90e2",   
  "consecutive hotspot"  = "#e31a1c",   
  "new coldspot"         = "#20b2aa",
  "new hotspot"          = "#ff69b4",  
  "no pattern detected"  = "grey80", 
  "persistent coldspot"  = "#08519c", 
  "persistent hotspot"   = "#8b0000",  
  "sporadic coldspot"    = "#c6dbef", 
  "sporadic hotspot"     = "#fcae91" 
)

ehsa_morning$classification   <- factor(ehsa_morning$classification,   levels = ehsa_levels)
ehsa_morning_sf$classification <- factor(ehsa_morning_sf$classification, levels = ehsa_levels)
```

#### Distribution of EHSA Classes

```{r}
ggplot(data = ehsa_morning, 
       aes(x = classification, fill = classification)) +
  geom_bar(color = "black", alpha = 0.8) +
  scale_fill_manual(values = ehsa_colors, drop = FALSE) +
  labs(
    title = "Distribution of EHSA Classes – Weekday Morning (5–10am)",
    x = "EHSA Classification", 
    y = "Number of Hexagons"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", size = 14)
  )

```

The EHSA classification distribution for weekday mornings (5–10am) reveals that the majority of hexagons fall under no pattern detected, indicating that most areas do not exhibit significant or consistent hotspot or coldspot behaviour. However, a substantial number are classified as sporadic coldspots and sporadic hotspots, reflecting localized and irregular clustering of both low and high bus trip counts. Persistent coldspots are also evident, though persistent hotspots are largely absent, suggesting that consistently high-demand areas are rare in this period. Overall, the results highlight a spatially uneven demand pattern, with significant variability across different parts of the network.

#### EHSA map and Gi* Trends

```{r fig.width=16, fig.height=6}
# Create the EHSA map
tmap_mode("plot")
ehsa_map <- tm_shape(mpsz) + 
  tm_polygons(col = "grey90", border.col = "white") +
  tm_shape(ehsa_morning_sf) +
  tm_polygons(
    fill = "classification",
    fill.scale = tm_scale_categorical(values = ehsa_colors),
    fill.legend = tm_legend(
      title = "EHSA Classification",
      position = tm_pos_in("right", "bottom"))
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_shape(mpsz_planning) +
  tm_borders(col = "grey30", lwd = 1.5) +
  tm_layout(
    main.title = "EHSA Map",
    main.title.size = 1.2,
    main.title.fontface = "bold"
  )

# Get one representative hexagon from each significant EHSA classification
representative_hexagons <- ehsa_morning_sf %>%
  st_drop_geometry() %>%
  filter(classification != "no pattern detected") %>%
  group_by(classification) %>%
  slice(1) %>%
  ungroup() %>%
  select(HEX_ID, classification)

# Extract Gi* trends for these hexagons
hex_trends <- gi_morning %>%
  st_drop_geometry() %>%
  filter(HEX_ID %in% representative_hexagons$HEX_ID) %>%
  select(HEX_ID, HOUR_OF_DAY, gi_star) %>%
  left_join(representative_hexagons, by = "HEX_ID")

# Create individual plots for each classification type
plot_list <- list()

for (class in unique(hex_trends$classification)) {
  hex_data <- hex_trends %>% filter(classification == class)
  hex_id <- unique(hex_data$HEX_ID)
  
  p <- ggplot(data = hex_data, aes(x = HOUR_OF_DAY, y = gi_star)) +
    geom_line(color = ehsa_colors[class], linewidth = 1.2) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
    labs(
      title = paste0(class, "\n(", hex_id, ")"),
      x = "Hour of Day",
      y = "Gi*"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", size = 10, hjust = 0.5),
      axis.title = element_text(size = 9),
      axis.text = element_text(size = 8)
    )
  
  plot_list[[class]] <- p
}

# Combine trend plots
combined_trends <- wrap_plots(plot_list, ncol = 2) +
  plot_annotation(
    title = "Gi* Trends by EHSA Classification",
    theme = theme(
      plot.title = element_text(face = "bold", size = 12, hjust = 0.5)
    )
  )

# Convert tmap to grob
ehsa_map_grob <- tmap_grob(ehsa_map)

# Combine map and trends side by side
final_plot <- plot_grid(
  ehsa_map_grob,
  combined_trends,
  ncol = 2,
  rel_widths = c(1, 1),
  labels = NULL
)

# Add overall title
title <- ggdraw() + 
  draw_label(
    "Emerging Hot Spot Analysis (EHSA) – Weekday Morning (5–10am)",
    fontface = 'bold',
    size = 15,
    x = 0.5,
    hjust = 0.5
  )

# Combine title with plots
final_combined <- plot_grid(
  title,
  final_plot,
  ncol = 1,
  rel_heights = c(0.05, 0.95)
)

final_combined
```

The EHSA analysis reveals distinct spatial and temporal clustering patterns of weekday morning bus trips (5-10am). The map (left) shows blue coldspots concentrated in western and southwestern planning areas, indicating persistently low trip generation zones. Red and orange hotspots appear in central, eastern, and northern regions, representing areas with consistently high morning demand.

The Gi* trend plots (right) demonstrate diverse temporal behaviors across classifications. Consecutive hotspots (red, H0242) maintain strongly positive Gi* values (2-7) throughout the morning, peaking around 8am, indicating stable high-demand zones that intensify during peak hours. Sporadic hotspots (orange, H0171) show fluctuating positive values (1-3), suggesting intermittent clustering behavior.

Persistent coldspots (dark blue, H0007) remain consistently negative (-1.0 to -1.5), confirming stable low-demand areas throughout the period. Consecutive coldspots (light blue, H0118) show declining Gi* values (from -0.5 to -1.5), indicating areas becoming increasingly isolated from surrounding activity as the morning progresses. Sporadic coldspots (H0001) fluctuate near zero, reflecting irregular patterns.

Overall, the analysis confirms that certain planning areas exhibit stable, predictable temporal patterns (consecutive/persistent classifications), while others show irregular, time-varying behaviors (sporadic classifications)


# 5 Weekday afternoon peak

## 5.1 Overview

### Choropleth map of total trips

```{r}
weekday_afternoon <- trips %>%
  filter(DAY_TYPE == "WEEKDAY", HOUR_OF_DAY %in% 17:20) %>%
  group_by(HEX_ID) %>%
  summarise(TRIPS = sum(TRIPS), .groups = "drop") %>%
  left_join(hexagon, by = "HEX_ID") %>%
  st_as_sf()

tmap_mode("plot")

tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  
  tm_shape(weekday_afternoon) +
  tm_polygons(
    fill = "TRIPS",
    fill.scale = tm_scale_intervals(
      style = "quantile",  # Use style = "quantile" instead
      n = 5,
      values = "viridis"
    ),
    fill.legend = tm_legend(
      title = "Total Trips\n(Weekday Afternoon)",
      position = tm_pos_in("right", "bottom")
    )
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_shape(mpsz_area) +
  tm_borders(col = "black", lwd = 0.2) +
  tm_title("Distribution of Bus Trips - Weekday Afternoon Peak (5pm-8pm)") +
  tm_compass(position = tm_pos_in("left", "top")) +
  tm_scale_bar(position = tm_pos_in("left", "bottom"))
```

The visualisation shows that bus trips during the weekday afternoon peak (5–8pm) are highly concentrated in central and western regions such as Jurong East, Clementi, and Queenstown, as well as large residential towns including Tampines, Yishun, and Woodlands. These areas reflect strong homeward-bound commuting flows as people return from central business districts, universities, and industrial hubs. Moderate activity is evident in northern and northeastern towns such as Sengkang, Hougang, and Punggol, which act as important residential catchments. In contrast, peripheral zones, especially coastal and industrial areas, record relatively low trip counts. The overall spatial pattern highlights the dominance of evening return journeys, with trips dispersing across multiple residential clusters rather than converging on central areas.

### Trip Distribution

```{r}
# Filter out zero trips
weekday_afternoon_filtered <- weekday_afternoon %>% filter(TRIPS > 0)

# Calculate statistics for filtered data
mean_trips <- mean(weekday_afternoon_filtered$TRIPS)
median_trips <- median(weekday_afternoon_filtered$TRIPS)
q95 <- quantile(weekday_afternoon_filtered$TRIPS, 0.95)

# Plot 1: Original scale histogram (cut off at 95th percentile)
x_limit <- q95
p1 <- ggplot(weekday_afternoon_filtered, aes(x = TRIPS)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(aes(xintercept = mean_trips), 
             color = "red", linetype = "dashed", linewidth = 0.8) +
  geom_vline(aes(xintercept = median_trips), 
             color = "orange", linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = mean_trips, y = Inf, 
           label = paste0("Mean: ", format(round(mean_trips, 0), big.mark = ",")), 
           vjust = 1.5, hjust = -0.1, color = "red", size = 3.5, fontface = "bold") +
  annotate("text", x = median_trips, y = Inf, 
           label = paste0("Median: ", format(round(median_trips, 0), big.mark = ",")), 
           vjust = 3, hjust = -0.1, color = "orange", size = 3.5, fontface = "bold") +
  scale_x_continuous(
    labels = label_number(scale = 1/1000, suffix = "k"),
    breaks = seq(0, x_limit, by = 20000),
    limits = c(0, x_limit),
    expand = expansion(mult = c(0.02, 0.05))
  ) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "0-95th Percentile",
    x = "Number of Trips",
    y = "Frequency",
    caption = paste0("95th percentile: ", format(round(q95, 0), big.mark = ","), " trips")
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
    axis.title = element_text(size = 10),
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 9),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = "gray90"),
    plot.margin = margin(10, 5, 10, 10),
    plot.caption = element_text(size = 8, color = "gray50", hjust = 1, face = "italic")
  )

# Plot 2: Bar chart for 95th-100th percentile (top 5%)
top5_data <- weekday_afternoon_filtered %>% filter(TRIPS >= q95)

p2 <- ggplot(top5_data, aes(x = TRIPS)) +
  geom_histogram(bins = 25, fill = "coral", color = "white", alpha = 0.8) +
  scale_x_continuous(
    labels = label_number(scale = 1/1000, suffix = "k"),
    expand = expansion(mult = c(0.02, 0.05))
  ) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Top 5%",
    x = "Number of Trips",
    y = "Frequency",
    caption = paste0("n = ", nrow(top5_data), " hexagons | Max: ", 
                    format(max(top5_data$TRIPS), big.mark = ","))
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
    axis.title = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = "gray90"),
    plot.margin = margin(10, 5, 10, 5),
    plot.caption = element_text(size = 8, color = "gray50", hjust = 0.5)
  )

# Plot 3: Log transformation
p3 <- ggplot(weekday_afternoon_filtered, aes(x = log10(TRIPS))) +
  geom_histogram(bins = 50, fill = "darkgreen", color = "white", alpha = 0.8) +
  scale_x_continuous(
    breaks = 0:5,
    labels = c("1", "10", "100", "1k", "10k", "100k"),
    expand = expansion(mult = c(0.02, 0.05))
  ) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Log10 Transformed",
    x = "Number of Trips (Log Scale)",
    y = "Frequency",
    caption = "Distribution shape across all magnitudes"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
    axis.title = element_text(size = 10),
    axis.text.x = element_text(size = 9),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = "gray90"),
    plot.margin = margin(10, 10, 10, 5),
    plot.caption = element_text(size = 8, color = "gray50", hjust = 0.5)
  )

# Combine plots with adjusted widths
combined_plot <- p1 + p2 + p3 + 
  plot_layout(widths = c(2.5, 1.5, 2.5)) +
  plot_annotation(
    title = "Distribution of Trip Counts - Weekday Afternoon Peak (5pm-8pm)",
    subtitle = paste0("n = ", format(nrow(weekday_afternoon_filtered), big.mark = ","), 
                     " hexagons (zeros excluded)"),
    theme = theme(
      plot.title = element_text(face = "bold", size = 15, hjust = 0.5),
      plot.subtitle = element_text(size = 11, color = "gray40", hjust = 0.5)
    )
  )

combined_plot
```

This three-panel visualization reveals a highly right-skewed distribution of weekday afternoon peak trips (5pm-8pm) across 816 hexagonal zones in the study area. The main distribution (0-95th percentile) shows that the majority of hexagons have relatively low trip counts, with a median of 14,872 trips and mean of 30,263 trips. The substantial gap between median and mean indicates strong positive skew, suggesting most areas experience moderate afternoon traffic while a few zones generate disproportionately high trip volumes. The distribution shows a sharp peak near zero with a long tail extending rightward, similar to the morning peak pattern but with slightly lower central values.

The top 5% panel highlights the extreme values, where only 41 hexagons exceed the 95th percentile threshold of 100,900 trips, with the maximum reaching 315,613 trips. These outliers likely represent major transportation hubs, central business districts, or key transit corridors that concentrate afternoon commute activity as workers return home. Notably, the maximum afternoon trips (315,613) are lower than morning peaks (396,146), suggesting slightly more dispersed evening travel patterns.

The log-transformed view reveals a more normalized, approximately bell-shaped distribution, confirming the underlying data follows a log-normal distribution. The pattern demonstrates that afternoon trip generation spans multiple orders of magnitude, reflecting diverse urban zones from residential neighborhoods to major activity centers and employment hubs experiencing peak evening demand.

## 5.2 Local Indicators of Spatial Association(LISA)

The code below computes Local Moran’s I for TRIPS in the weekday-afternoon hexagons (5–8pm), using a distance-band neighbor list at 1984.313 meters.

```{r}
# Compute Local Moran's I
set.seed(1234)
lisa_afternoon <- weekday_afternoon %>%
  mutate(
    local_moran = local_moran(
      TRIPS, nb, wt, nsim = 99), 
    .before = 1) %>%
  unnest(local_moran)
```

### Local Moran’s I values

```{r}
# Filter for significant Local Moran's I (p-value < 0.1)
lisa_afternoon_sig <- lisa_afternoon %>%
  filter(p_ii < 0.1)

# Create the Local Moran's I map
ii_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  tm_shape(lisa_afternoon_sig) +
  tm_polygons(fill = "ii",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 5,
                values = "brewer.RdBu"
              ),
              fill.legend = tm_legend(
                title = "Local Moran's I",
                position = tm_pos_in("right", "bottom"),
                title.size = 0.8,
                text.size = 0.7
              )
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_shape(mpsz_area) +
  tm_borders(col = "grey30", lwd = 0.2) +
  tm_title("Local Moran's I - Afternoon Peak (p-value < 0.1)") +
  tm_compass(position = tm_pos_in("left", "top")) +
  tm_scale_bar(position = tm_pos_in("left", "bottom"))

# Display the map
ii_map
```

The Local Moran's I map reveals significant spatial clustering patterns of bus trip generation during weekday afternoon peak hours, with only statistically significant results (p < 0.1) displayed. Positive spatial autocorrelation (blue hexagons, values 0.5-1.5) appears more scattered compared to morning patterns, concentrated primarily in western, southwestern, and selected eastern planning areas. These high-high clusters indicate areas where high trip volumes are surrounded by similarly high-activity neighbors, representing residential zones receiving concentrated return commuter flows during evening peak hours.

Moderate positive clusters (light blue, 0-0.5) appear dispersed across various planning areas, suggesting more fragmented evening clustering patterns than morning peaks. Notably, negative spatial autocorrelation (orange to red, -1.0 to 0) appears extensively throughout the island, particularly in central, northern, and eastern regions. These represent spatial outliers—either high-activity hexagons surrounded by low-activity neighbors (high-low outliers) or vice versa (low-high outliers)—indicating transitional boundaries, edge effects between different land uses, or isolated activity centers.

The afternoon pattern differs markedly from morning peaks, showing more dispersed clustering as commuters return from centralized employment hubs to diverse residential locations. The fragmented spatial structure reflects Singapore's polycentric urban form and varied evening travel patterns, with some residential areas forming return-commute clusters while others exhibit irregular demand.

### Moran scatterplot with standardised variable

```{r fig.width=10, fig.height=6}
lisa_afternoon <- lisa_afternoon %>%
  mutate(lag_TRIPS = st_lag(
    TRIPS, nb, wt),
    .before = 1) %>%
  unnest(lag_TRIPS)

lisa_afternoon <- lisa_afternoon %>%
  mutate(z_TRIPS = scale(TRIPS),
         z_lag_TRIPS = scale(lag_TRIPS),
         .before = 1)

# Create the Moran scatterplot
moran_plot <- ggplot(data = lisa_afternoon,
       aes(x = z_TRIPS,
           y = z_lag_TRIPS,
           color = mean)) +
  geom_point(size = 2, alpha = 0.6) +
  geom_smooth(method = "lm",
              se = FALSE,
              color = "black") +
  geom_hline(yintercept = mean(lisa_afternoon$z_lag_TRIPS), lty = 2) +
  geom_vline(xintercept = mean(lisa_afternoon$z_TRIPS), lty = 2) +
  scale_color_manual(
    values = c("High-High" = "red",
               "Low-Low" = "blue",
               "Low-High" = "lightblue",
               "High-Low" = "pink")) +
  labs(x = "Standardised TRIPS",
       y = "Standardised Spatial Lag of TRIPS",
       title = "Standardised Moran Scatterplot with LISA Quadrants") +
  theme_minimal() +
  theme(legend.position = "bottom")

# Create the bar chart for quadrant counts
quadrant_counts <- lisa_afternoon %>%
  st_drop_geometry() %>%
  count(mean) %>%
  mutate(mean = factor(mean, levels = c("High-High", "Low-High", 
                                         "Low-Low", "High-Low")))

bar_plot <- ggplot(quadrant_counts, aes(x = mean, y = n, fill = mean)) +
  geom_bar(stat = "identity", alpha = 0.8) +
  geom_text(aes(label = n), vjust = -0.5, size = 4, fontface = "bold") +
  scale_fill_manual(
    values = c("High-High" = "red",
               "Low-Low" = "blue",
               "Low-High" = "lightblue",
               "High-Low" = "pink")) +
  labs(x = "LISA Quadrant",
       y = "Count of Hexagons",
       title = "Count by Quadrant") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

# Combine plots side by side
combined_plot <- moran_plot + bar_plot + 
  plot_layout(widths = c(2, 1)) +
  plot_annotation(
    title = "Local Moran's I Analysis - Weekday Afternoon Peak",
    theme = theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5))
  )

combined_plot
```

The Moran scatterplot reveals distinct spatial patterns in weekday afternoon bus trips, with the bar chart quantifying the distribution across LISA quadrants. The Low-Low quadrant (dark blue, 319 hexagons) dominates, representing cold spots where low-trip hexagons cluster together in peripheral or less populated regions, exhibiting strong positive spatial autocorrelation. However, this is notably lower than morning peaks (346), suggesting more dispersed evening patterns.

The Low-High quadrant (light blue, 248 hexagons) shows a significant increase from morning (178), representing underperforming hexagons within busier regions. This indicates substantial negative spatial autocorrelation where low-activity areas are surrounded by high-activity neighbors, reflecting transitional zones and spatial heterogeneity as evening commuters disperse to diverse residential locations.

The High-High quadrant (red, 192 hexagons) decreased from morning levels (239), showing fewer hot spots—high-trip hexagons surrounded by similarly high-trip neighbors. This reflects more fragmented afternoon clustering compared to concentrated morning residential-to-employment flows. The High-Low quadrant (pink, 57 hexagons) contains minimal outliers—isolated high-trip hexagons in low-activity surroundings.

The positive regression slope confirms overall positive spatial autocorrelation, though weaker than morning patterns. The more balanced distribution across quadrants reveals greater spatial complexity in afternoon trip patterns, with increased heterogeneity reflecting Singapore's polycentric structure and varied evening travel behaviors as commuters return from centralized employment hubs to dispersed residential areas.

### LISA Map

```{r fig.width=12, fig.height=5}
# Count significant quadrants
quadrant_counts <- lisa_afternoon_sig %>%
  st_drop_geometry() %>%
  count(mean) %>%
  mutate(mean = factor(mean, levels = c("High-High", "Low-High", 
                                         "Low-Low", "High-Low")))

# Bar chart
bar_chart <- ggplot(quadrant_counts, aes(x = mean, y = n, fill = mean)) +
  geom_bar(stat = "identity", width = 0.7) +
  geom_text(aes(label = n), vjust = -0.3, size = 5, fontface = "bold") +
  scale_fill_manual(
    values = c("High-High" = "red",
               "Low-Low" = "blue",
               "Low-High" = "lightblue",
               "High-Low" = "pink")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(x = "LISA Quadrant", 
       y = "Count of Hexagons", 
       title = "LISA Quadrant Counts") +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    axis.text.y = element_text(size = 11),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 13, face = "bold", hjust = 0.5),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.margin = margin(10, 10, 10, 10)
  )

# LISA map
tmap_mode("plot")
lisa_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  
  tm_shape(lisa_afternoon_sig) +
  tm_polygons(fill = "mean",
              fill.scale = tm_scale_categorical(
                values = c("High-High" = "red",
                          "Low-Low" = "blue",
                          "Low-High" = "lightblue",
                          "High-Low" = "pink")
              ),
              fill.legend = tm_legend(
                title = "LISA Quadrants",
                position = tm_pos_in("right", "bottom")
              )
  ) +
  tm_borders(col = "black", lwd = 0.5) +
  tm_shape(mpsz_area) +
  tm_borders(col = "grey30", lwd = 1.5) +
  tm_layout(
    main.title = "LISA Map (p-value < 0.1)",
    main.title.size = 1.2,
    main.title.fontface = "bold",
    legend.text.size = 0.8,
    legend.title.size = 1,
    inner.margins = c(0.05, 0.05, 0.05, 0.05)
  )

# Convert tmap to grob
lisa_map_grob <- tmap_grob(lisa_map)

# Combine
combined_plot <- plot_grid(
  lisa_map_grob, 
  bar_chart, 
  ncol = 2, 
  rel_widths = c(2,1.5),
  labels = NULL
)

title <- ggdraw() + 
  draw_label(
    "Local Indicators of Spatial Association (LISA) - Weekday Afternoon Peak",
    fontface = 'bold',
    size = 15,
    x = 0.5,
    hjust = 0.5
  )

final_plot <- plot_grid(
  title, 
  combined_plot,
  ncol = 1,
  rel_heights = c(0.06, 0.94)
)

final_plot
```

The LISA map reveals significantly different spatial clustering patterns for weekday afternoon bus trips compared to morning peaks. The distribution is remarkably balanced, with High-High clusters (red, 68 hexagons), Low-High outliers (light blue, 69 hexagons), and Low-Low clusters (dark blue, 69 hexagons) nearly equal in number. This contrasts sharply with morning patterns where Low-Low and High-High dominated, indicating greater spatial heterogeneity in afternoon travel.

High-High clusters appear more scattered across northern, central, and eastern planning areas, representing residential zones receiving concentrated return commuter flows. However, their reduced prevalence (68 vs 145 morning) reflects more dispersed evening destination patterns. Low-Low clusters remain present in peripheral and less populated regions but are substantially fewer (69 vs 150 morning), suggesting broader spatial distribution of afternoon activity.

The dramatic increase in Low-High outliers (69 vs 78 morning) indicates extensive spatial mixing where low-activity hexagons neighbor high-activity areas, reflecting transitional zones and fragmented clustering as evening commuters disperse from centralized employment hubs to diverse residential locations. High-Low outliers remain minimal (4 hexagons).

The balanced quadrant distribution reveals weaker positive spatial autocorrelation and greater spatial complexity compared to morning peaks. This reflects Singapore's polycentric structure and varied evening travel behaviors, with afternoon trip patterns exhibiting more decentralized, heterogeneous spatial organization.


## 5.3 Hot Spots and Cold Spots Analysis (HCSA)

The code below builds a fixed distance–band neighbourhood for each weekday-afternoon hexagon, creates row-standardised spatial weights, and then computes the Local Getis-Ord Gi* statistic for TRIPS.

```{r}
weekday_afternoon_fdw <- weekday_afternoon %>%
  mutate(nb = include_self(
    st_dist_band(
      st_geometry(geometry),
      upper = critical_dist)),
    wt = st_weights(nb,
                    style = "W"),
    .before = 1)

# Compute local Gi*
set.seed(1234)

HCSA_fdw <- weekday_afternoon_fdw %>%
  mutate(
    gistar = local_gstar_perm(
      TRIPS, nb, wt, nsim = 99),
    .before = 1) %>%
  unnest(gistar)
```

```{r}
# Filter for significant Gi* values (p < 0.1)
HCSA_sig <- HCSA_fdw %>%
  filter(p_sim < 0.1)

# Plot Gi* values for significant hexagons only
Gi_star_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  
  tm_shape(HCSA_sig) +
  tm_polygons(fill = "gi_star",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 5,
                values = "-brewer.rd_bu"),
              fill.legend = tm_legend(
                title = "local Gi*",
                position = tm_pos_in("right", "bottom")
              )
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_shape(mpsz_area) +
  tm_borders(col = "grey30", lwd = 0.2) +
  tm_title("Local Gi* of TRIPS (p-value < 0.1)") +
  tm_compass(position = tm_pos_in("left", "top")) +
  tm_scale_bar(position = tm_pos_in("left", "bottom"))

# HCSA cluster classification with p < 0.1
HCSA_fdw <- HCSA_fdw %>%
  mutate(HCSA_cluster = case_when(
    p_sim <= 0.1 & cluster == "High" ~ "Hot spot",
    p_sim <= 0.1 & cluster == "Low" ~ "Cold spot",
    TRUE ~ "Other"),
    HCSA_cluster = factor(
      HCSA_cluster,
      levels = c("Hot spot", "Cold spot")  # Reordered levels
    ),
    .before = 1
  )

# Filter to keep only significant hexagons
HCSA_sig <- HCSA_fdw %>%
  filter(HCSA_cluster != "Insignificant")

# Aggregate mpsz by PLN_AREA_N to get planning area boundaries
mpsz_planning <- mpsz %>%
  group_by(PLN_AREA_N) %>%
  summarise(geometry = st_union(geometry))

# HCSA cluster map
HCSA_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  
  tm_shape(HCSA_sig) +
  tm_polygons(
    fill = "HCSA_cluster",
    fill.scale = tm_scale_categorical(
      values = c(
        "Hot spot" = "red",
        "Cold spot" = "blue"
      )
    ),
    fill.legend = tm_legend(
      title = "HCSA Cluster",
      position = tm_pos_in("right", "bottom"))
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_shape(mpsz_area) +
  tm_borders(col = "grey30", lwd = 0.2) +
  tm_title("HCSA Clusters (p-value < 0.1)")

tmap_arrange(Gi_star_map, HCSA_map, asp = 1, ncol = 2)
```

The Local Gi* map identifies scattered hot spots (red/orange) across northern, central, and selected eastern regions indicating high afternoon trip clustering intensity, though notably more fragmented than morning patterns. Cold spots (blue) dominate western and peripheral areas with consistently low trips. The dispersed pattern suggests localized intensity variations reflecting diverse evening destination patterns rather than concentrated commuter flows.

The HCSA Clusters map reveals statistically significant spatial clustering of weekday afternoon bus trips (p < 0.1), showing markedly different patterns from morning peaks. Hot spots (red) appear as scattered, smaller clusters across northern, central, and eastern planning areas, representing residential zones receiving return commuter flows. However, these are less extensive and more fragmented than morning clusters, reflecting dispersed evening travel as workers return from centralized employment hubs to diverse residential locations. Cold spots (blue) remain prominent across western, northwestern, and peripheral regions, though their distribution is more varied, confirming spatial patterns of reduced evening transit demand in non-residential areas.

Together, the maps demonstrate more complex spatial organization in afternoon trip generation compared to morning peaks. The Gi* map shows fragmented clustering intensity, while the HCSA map reveals smaller, dispersed hot spot clusters. Both confirm weaker spatial autocorrelation with greater heterogeneity. The afternoon pattern lacks the clear residential-to-employment spatial segregation characteristic of morning commutes, instead showing decentralized trip generation across multiple activity centers.

## 5.4 Emerging Hot Spot Analysis (EHSA) - 5 to 10 pm

The code below prepares weekday afternoon (17–22h) trips for Emerging Hot Spot/space–time analysis. It aggregates trips by hexagon × hour, then ensure every hexagon has all hours (filling missing cells with 0) and orders IDs/time consistently. It converts IDs and hours to the right types, then builds a space–time cube (spacetime) by linking the hourly table to the hexagon geometries, and finally checks that the result is a valid space–time cube.

```{r}
hex_ids <- hexagon$HEX_ID

trips_afternoon_hourly <- trips %>%
  filter(DAY_TYPE == "WEEKDAY", HOUR_OF_DAY %in% 17:22) %>%
  group_by(HEX_ID, HOUR_OF_DAY) %>%
  summarise(TRIPS = sum(TRIPS), .groups = "drop") %>%
  tidyr::complete(HEX_ID = hex_ids, HOUR_OF_DAY = 17:22,
                  fill = list(TRIPS = 0)) %>%
  arrange(HEX_ID, HOUR_OF_DAY)

trips_afternoon_hourly$HEX_ID <- factor(trips_afternoon_hourly$HEX_ID, levels = hex_ids)
trips_afternoon_hourly$HOUR_OF_DAY <- as.integer(trips_afternoon_hourly$HOUR_OF_DAY)

# Build space–time cube
TRIPS_afternoon_st <- spacetime(
  trips_afternoon_hourly,
  hexagon %>% select(HEX_ID), 
  .loc_col  = "HEX_ID",
  .time_col = "HOUR_OF_DAY"
)

is_spacetime_cube(TRIPS_afternoon_st) 
```

The code below adds spatial neighbours/weights to the afternoon space–time cube and computes hourly Local Getis-Ord Gi* for TRIPS. It then flattens the results so each hexagon–hour has a Gi* score and significance.

```{r}
# Add neighbours and weights
TRIPS_afternoon_nb <- TRIPS_afternoon_st %>%
  activate("geometry") %>%
  mutate(
    nb = include_self(st_contiguity(geometry)), 
    wt = st_inverse_distance(nb, geometry, scale = 1, alpha = 1),
    .before = 1
  ) %>%
  set_nbs("nb") %>%
  set_wts("wt")

# Compute Gi* 
set.seed(1234)

gi_afternoon <- TRIPS_afternoon_nb %>%
  group_by(HOUR_OF_DAY) %>%
  mutate(gi_star = local_gstar_perm(TRIPS, nb, wt, nsim = 99)) %>%
  tidyr::unnest(gi_star) %>%
  ungroup()
```

### Mann-Kendall Test

The code below tests whether each hexagon’s Gi* values show a monotonic time trend (consistently increasing or decreasing) across the selected hours.

H0: No monotonic trend

H1: Monotonic trend is present

alpha: 0.1

```{r}
mk_hotspots_afternoon <- gi_afternoon %>%
  st_drop_geometry() %>%
  group_by(HEX_ID) %>%
  summarise(
    mk = if(n() >= 3) list(unclass(Kendall::MannKendall(gi_star))) else NA,
    .groups = "drop"
  ) %>%
  tidyr::unnest_wider(mk) %>%
  rename(tau = tau, p_value = sl)

mk_hotspots_afternoon_ranked <- mk_hotspots_afternoon %>%
  filter(!is.na(p_value)) %>%
  arrange(p_value, desc(abs(tau)))        

mk_hotspots_afternoon_ranked %>% slice(1:10)
```

With reference to the results, as p-value < alpha, we will reject the null hypothesis and infer that there is a significant trend.

### Performing Emerging Hotspot Analysis

The code below runs EHSA on the weekday–morning space-time cube to classify each hexagon into temporal hotspot/coldspot types.

```{r}
set.seed(1234)

ehsa_afternoon <- emerging_hotspot_analysis(
  x     = TRIPS_afternoon_nb,   
  .var  = "TRIPS",     
  k     = 1,      
  nsim  = 99   
)
```

The code below attaches EHSA results back to the hexagon layer and keep only statistically significant cells for mapping.

```{r}
ehsa_afternoon_sf <- hexagon %>%
  select(HEX_ID) %>%
  left_join(ehsa_afternoon, by = c("HEX_ID" = "location")) %>%
  filter(p_value < 0.1) %>%
  st_as_sf()
```

The code below applies the consistent legend order and color scheme to the data before plotting.

```{r}
ehsa_afternoon$classification   <- factor(ehsa_afternoon$classification,   levels = ehsa_levels)
ehsa_afternoon_sf$classification <- factor(ehsa_afternoon_sf$classification, levels = ehsa_levels)
```

#### Distribution of EHSA Classes

```{r}
ggplot(data = ehsa_afternoon, 
       aes(x = classification, fill = classification)) +
  geom_bar(color = "black", alpha = 0.8) +
  scale_fill_manual(values = ehsa_colors, drop = FALSE) +
  labs(
    title = "Distribution of EHSA Classes – Weekday Afternoon (5–10pm)",
    x = "EHSA Classification", 
    y = "Number of Hexagons"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", size = 14)
  )
```

The EHSA distribution for the weekday afternoon peak (5-10pm) shows that most hexagons fall under no pattern detected, suggesting that clustering of demand is limited across much of the city. Among significant categories, consecutive coldspots and sporadic coldspots are notable, indicating zones of consistently low demand, particularly in peripheral or less-connected areas. Sporadic hotspots also appear prominently, reflecting irregular but short-lived surges in demand at specific locations as evening commuters disperse. By contrast, persistent and consecutive hotspots are nearly absent, underscoring the lack of sustained high-demand clusters during this time window. Overall, the afternoon peak is defined by widespread irregular patterns and coldspot dominance, with hotspots emerging only intermittently and sporadically.

#### EHSA map and Gi* Trends

```{r fig.width=16, fig.height=6}
# Create the EHSA map
tmap_mode("plot")
ehsa_map <- tm_shape(mpsz) + 
  tm_polygons(col = "grey90", border.col = "white") +
  tm_shape(ehsa_afternoon_sf) +
  tm_polygons(
    fill = "classification",
    fill.scale = tm_scale_categorical(values = ehsa_colors),
    fill.legend = tm_legend(
      title = "EHSA Classification",
      position = tm_pos_in("right", "bottom"))
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_shape(mpsz_planning) +
  tm_borders(col = "grey30", lwd = 1.5) +
  tm_layout(
    main.title = "EHSA Map",
    main.title.size = 1.2,
    main.title.fontface = "bold"
  )

# Get one representative hexagon from each significant EHSA classification
representative_hexagons <- ehsa_afternoon_sf %>%
  st_drop_geometry() %>%
  filter(classification != "no pattern detected") %>%
  group_by(classification) %>%
  slice(1) %>%
  ungroup() %>%
  select(HEX_ID, classification)

# Extract Gi* trends for these hexagons
hex_trends <- gi_afternoon %>%
  st_drop_geometry() %>%
  filter(HEX_ID %in% representative_hexagons$HEX_ID) %>%
  select(HEX_ID, HOUR_OF_DAY, gi_star) %>%
  left_join(representative_hexagons, by = "HEX_ID")

# Create individual plots for each classification type
plot_list <- list()

for (class in unique(hex_trends$classification)) {
  hex_data <- hex_trends %>% filter(classification == class)
  hex_id <- unique(hex_data$HEX_ID)
  
  p <- ggplot(data = hex_data, aes(x = HOUR_OF_DAY, y = gi_star)) +
    geom_line(color = ehsa_colors[class], linewidth = 1.2) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
    labs(
      title = paste0(class, "\n(", hex_id, ")"),
      x = "Hour of Day",
      y = "Gi*"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", size = 10, hjust = 0.5),
      axis.title = element_text(size = 9),
      axis.text = element_text(size = 8)
    )
  
  plot_list[[class]] <- p
}

# Combine trend plots
combined_trends <- wrap_plots(plot_list, ncol = 2) +
  plot_annotation(
    title = "Gi* Trends by EHSA Classification",
    theme = theme(
      plot.title = element_text(face = "bold", size = 12, hjust = 0.5)
    )
  )

# Convert tmap to grob
ehsa_map_grob <- tmap_grob(ehsa_map)

# Combine map and trends side by side
final_plot <- plot_grid(
  ehsa_map_grob,
  combined_trends,
  ncol = 2,
  rel_widths = c(1, 1),
  labels = NULL
)

# Add overall title
title <- ggdraw() + 
  draw_label(
    "Emerging Hot Spot Analysis (EHSA) – Weekday Afternoon (5–10pm)",
    fontface = 'bold',
    size = 15,
    x = 0.5,
    hjust = 0.5
  )

# Combine title with plots
final_combined <- plot_grid(
  title,
  final_plot,
  ncol = 1,
  rel_heights = c(0.05, 0.95)
)

final_combined
```

The EHSA analysis reveals distinct spatial and temporal clustering patterns of weekday afternoon bus trips (5-10pm), showing markedly different characteristics from morning peaks. The map (left) displays blue coldspots scattered across western, northern, and peripheral planning areas, indicating persistently low evening trip generation zones. Red and orange hotspots appear more fragmented and dispersed across central and eastern regions, representing areas with elevated afternoon demand. However, these hotspots are notably less concentrated and extensive than morning patterns, reflecting the decentralized nature of evening commuter flows as workers return from centralized employment hubs to diverse residential locations.

The Gi* trend plots (right) demonstrate diverse temporal behaviors across classifications, revealing how clustering intensity evolves throughout the evening period. Consecutive hotspots (red, H0333) maintain strongly positive Gi* values (3-6) throughout the 5-10pm window, showing stable high-demand zones during return-commute hours at major residential destinations. Sporadic hotspots (orange, H0033) exhibit fluctuating positive values (2-4), suggesting intermittent clustering behavior as evening travel disperses across multiple activity centers and residential areas.

Persistent coldspots (dark blue, H0002) remain consistently negative (-0.5 to -1.0), confirming stable low-demand areas throughout the period in peripheral or non-residential zones. Consecutive coldspots (light blue, H0005) show declining Gi* values (from -0.5 to -1.0), indicating areas becoming increasingly isolated as evening activity diminishes. New coldspots (H0153) emerge with declining trends, representing areas that develop low-demand patterns specifically during evening hours. Sporadic coldspots (H0001) fluctuate near zero, reflecting irregular and unpredictable patterns.

Overall, the afternoon analysis demonstrates more fragmented spatial patterns and greater temporal variability compared to concentrated morning peaks. Certain planning areas exhibit stable, predictable patterns (consecutive/persistent classifications), while others show irregular, time-varying behaviors (sporadic classifications).

# 6 Weekend/holiday

## 6.1 Overview

### Choropleth map of total trips

```{r}
weekend <- trips %>%
  filter(DAY_TYPE == "WEEKENDS/HOLIDAY", HOUR_OF_DAY %in% 11:20) %>%
  group_by(HEX_ID) %>%
  summarise(TRIPS = sum(TRIPS), .groups = "drop") %>%
  left_join(hexagon, by = "HEX_ID") %>%
  st_as_sf()

tmap_mode("plot")

tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  
  tm_shape(weekend) +
  tm_polygons(
    fill = "TRIPS",
    fill.scale = tm_scale_intervals(
      style = "quantile",  # Use style = "quantile" instead
      n = 5,
      values = "viridis"
    ),
    fill.legend = tm_legend(
      title = "Total Trips\n(Weekend/holiday)",
      position = tm_pos_in("right", "bottom")
    )
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_shape(mpsz_area) +
  tm_borders(col = "black", lwd = 0.2) +
  tm_title("Distribution of Bus Trips - Weekend/holiday (11am-8pm)") +
  tm_compass(position = tm_pos_in("left", "top")) +
  tm_scale_bar(position = tm_pos_in("left", "bottom"))
```

The visualisation of weekend and holiday bus trips (11am–8pm) shows a more dispersed and leisure-oriented travel pattern compared to weekday peaks. High trip volumes are concentrated across large residential towns such as Tampines, Pasir Ris, Woodlands, and Yishun, as well as around central areas, reflecting strong demand for shopping, recreation, and social visits. Notably, trips also cluster around regional hubs like Jurong East and Tampines, which function as retail and entertainment centres.

Moderate activity is spread across northern and northeastern towns including Hougang, Sengkang, and Punggol, while the western corridor shows balanced demand, likely linked to weekend mall and park visits. In contrast, peripheral coastal and industrial zones record relatively low activity, as they are less associated with weekend travel purposes.

Overall, the spatial distribution suggests that weekend and holiday bus demand is polycentric and more evenly spread across residential and leisure destinations, rather than being heavily dominated by work-related commuting flows. This highlights the role of buses in supporting recreational mobility and connecting suburban households to key shopping, social, and leisure hubs across the island.

### Trip Distribution

```{r}
# Filter out zero trips
weekend_filtered <- weekend %>% filter(TRIPS > 0)

# Calculate statistics for filtered data
mean_trips <- mean(weekend_filtered$TRIPS)
median_trips <- median(weekend_filtered$TRIPS)
q95 <- quantile(weekend_filtered$TRIPS, 0.95)

# Plot 1: Original scale histogram (cut off at 95th percentile)
x_limit <- q95
p1 <- ggplot(weekend_filtered, aes(x = TRIPS)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(aes(xintercept = mean_trips), 
             color = "red", linetype = "dashed", linewidth = 0.8) +
  geom_vline(aes(xintercept = median_trips), 
             color = "orange", linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = mean_trips, y = Inf, 
           label = paste0("Mean: ", format(round(mean_trips, 0), big.mark = ",")), 
           vjust = 1.5, hjust = -0.1, color = "red", size = 3.5, fontface = "bold") +
  annotate("text", x = median_trips, y = Inf, 
           label = paste0("Median: ", format(round(median_trips, 0), big.mark = ",")), 
           vjust = 3, hjust = -0.1, color = "orange", size = 3.5, fontface = "bold") +
  scale_x_continuous(
    labels = label_number(scale = 1/1000, suffix = "k"),
    breaks = seq(0, x_limit, by = 20000),
    limits = c(0, x_limit),
    expand = expansion(mult = c(0.02, 0.05))
  ) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "0-95th Percentile",
    x = "Number of Trips",
    y = "Frequency",
    caption = paste0("95th percentile: ", format(round(q95, 0), big.mark = ","), " trips")
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
    axis.title = element_text(size = 10),
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 9),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = "gray90"),
    plot.margin = margin(10, 5, 10, 10),
    plot.caption = element_text(size = 8, color = "gray50", hjust = 1, face = "italic")
  )

# Plot 2: Bar chart for 95th-100th percentile (top 5%)
top5_data <- weekend_filtered %>% filter(TRIPS >= q95)

p2 <- ggplot(top5_data, aes(x = TRIPS)) +
  geom_histogram(bins = 25, fill = "coral", color = "white", alpha = 0.8) +
  scale_x_continuous(
    labels = label_number(scale = 1/1000, suffix = "k"),
    expand = expansion(mult = c(0.02, 0.05))
  ) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Top 5%",
    x = "Number of Trips",
    y = "Frequency",
    caption = paste0("n = ", nrow(top5_data), " hexagons | Max: ", 
                    format(max(top5_data$TRIPS), big.mark = ","))
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
    axis.title = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = "gray90"),
    plot.margin = margin(10, 5, 10, 5),
    plot.caption = element_text(size = 8, color = "gray50", hjust = 0.5)
  )

# Plot 3: Log transformation
p3 <- ggplot(weekend_filtered, aes(x = log10(TRIPS))) +
  geom_histogram(bins = 50, fill = "darkgreen", color = "white", alpha = 0.8) +
  scale_x_continuous(
    breaks = 0:5,
    labels = c("1", "10", "100", "1k", "10k", "100k"),
    expand = expansion(mult = c(0.02, 0.05))
  ) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Log10 Transformed",
    x = "Number of Trips (Log Scale)",
    y = "Frequency",
    caption = "Distribution shape across all magnitudes"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
    axis.title = element_text(size = 10),
    axis.text.x = element_text(size = 9),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = "gray90"),
    plot.margin = margin(10, 10, 10, 5),
    plot.caption = element_text(size = 8, color = "gray50", hjust = 0.5)
  )

# Combine plots with adjusted widths
combined_plot <- p1 + p2 + p3 + 
  plot_layout(widths = c(2.5, 1.5, 2.5)) +
  plot_annotation(
    title = "Distribution of Trip Counts - Weekend/holiday (11am-8pm)",
    subtitle = paste0("n = ", format(nrow(weekend_filtered), big.mark = ","), 
                     " hexagons (zeros excluded)"),
    theme = theme(
      plot.title = element_text(face = "bold", size = 15, hjust = 0.5),
      plot.subtitle = element_text(size = 11, color = "gray40", hjust = 0.5)
    )
  )

combined_plot
```

This three-panel visualization reveals a highly right-skewed distribution of weekend/holiday trips (11am-8pm) across 824 hexagonal zones in the study area. The main distribution (0-95th percentile) shows that the majority of hexagons have relatively low trip counts, with a median of 11,442 trips and mean of 25,297 trips. The substantial gap between median and mean indicates strong positive skew, suggesting most areas experience moderate weekend traffic while a few zones generate disproportionately high trip volumes. The distribution shows a sharp peak near zero with a long tail extending rightward, similar to weekday patterns but with notably lower central values, reflecting reduced overall transit demand during weekends and holidays.

The top 5% panel highlights the extreme values, where only 42 hexagons exceed the 95th percentile threshold of 89,450 trips, with the maximum reaching 369,851 trips. Notably, both the mean (25,297) and 95th percentile (89,450) are substantially lower than weekday patterns, confirming reduced commuter-driven demand. However, the maximum (369,851) remains high, suggesting certain weekend destinations attract concentrated visitor flows.

The log-transformed view reveals a more normalized, approximately bell-shaped distribution, confirming the underlying data follows a log-normal distribution typical for spatial phenomena. The pattern demonstrates that weekend trip generation spans multiple orders of magnitude, reflecting diverse urban zones from quiet residential neighborhoods to major leisure, retail, and recreational activity centers experiencing peak weekend demand.

## 6.2 Local Indicators of Spatial Association(LISA)

The code below computes Local Moran’s I for TRIPS in the weekend/holiday hexagons (11am–8pm), using a distance-band neighbor list at 1984.313 meters.

```{r}
nb <- st_dist_band(st_geometry(weekend), upper = critical_dist)
wt <- st_weights(nb, style = "W", allow_zero = TRUE)

# Compute Local Moran's I
set.seed(1234)
lisa_weekend <- weekend %>%
  mutate(
    local_moran = local_moran(
      TRIPS, nb, wt, nsim = 99), 
    .before = 1) %>%
  unnest(local_moran)
```

### Local Moran’s I values

```{r}
# Filter for significant Local Moran's I (p-value < 0.1)
lisa_weekend_sig <- lisa_weekend %>%
  filter(p_ii < 0.1)

# Create the Local Moran's I map
ii_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  tm_shape(lisa_weekend_sig) +
  tm_polygons(fill = "ii",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 5,
                values = "brewer.RdBu"
              ),
              fill.legend = tm_legend(
                title = "Local Moran's I",
                position = tm_pos_in("right", "bottom"),
                title.size = 0.8,
                text.size = 0.7
              )
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_shape(mpsz_area) +
  tm_borders(col = "grey30", lwd = 0.2) +
  tm_title("Local Moran's I - Weekend/holiday (p-value < 0.1)") +
  tm_compass(position = tm_pos_in("left", "top")) +
  tm_scale_bar(position = tm_pos_in("left", "bottom"))

# Display the map
ii_map
```

The Local Moran's I map reveals significant spatial clustering patterns of bus trip generation during weekend/holiday periods (11am-8pm), with only statistically significant results (p < 0.1) displayed. Positive spatial autocorrelation (blue hexagons, values 0.5-3) appears notably scattered and fragmented across the island, concentrated primarily in western, central, and selected eastern planning areas. These high-high clusters indicate areas where high trip volumes are surrounded by similarly high-activity neighbors, representing recreational destinations, shopping districts, or leisure zones attracting concentrated weekend visitor flows rather than commuter-driven patterns.

Moderate positive clusters (light blue, 0-1) appear sparsely dispersed across various planning areas, suggesting highly fragmented weekend clustering patterns distinct from weekday commuter flows. Notably, negative spatial autocorrelation (orange to red, -1 to 0) appears extensively throughout the island, particularly in northern, central, and peripheral regions. These represent spatial outliers—either high-activity leisure destinations surrounded by low-activity residential neighbors (high-low outliers) or vice versa (low-high outliers)—indicating transitional boundaries between recreational hubs and residential zones, or isolated weekend activity centers.

The weekend pattern differs fundamentally from weekday patterns, showing minimal concentrated clustering as leisure travel disperses across diverse recreational, retail, and tourist destinations rather than following predictable commuter routes. The highly fragmented spatial structure reflects Singapore's diverse weekend activity landscape and varied leisure travel behaviors, with some areas forming weekend attraction clusters while most exhibit irregular, destination-specific demand.

### Moran scatterplot with standardised variable

```{r fig.width=10, fig.height=6}
lisa_weekend <- lisa_weekend %>%
  mutate(lag_TRIPS = st_lag(
    TRIPS, nb, wt),
    .before = 1) %>%
  unnest(lag_TRIPS)

lisa_weekend <- lisa_weekend %>%
  mutate(z_TRIPS = scale(TRIPS),
         z_lag_TRIPS = scale(lag_TRIPS),
         .before = 1)

# Create the Moran scatterplot
moran_plot <- ggplot(data = lisa_weekend,
       aes(x = z_TRIPS,
           y = z_lag_TRIPS,
           color = mean)) +
  geom_point(size = 2, alpha = 0.6) +
  geom_smooth(method = "lm",
              se = FALSE,
              color = "black") +
  geom_hline(yintercept = mean(lisa_weekend$z_lag_TRIPS), lty = 2) +
  geom_vline(xintercept = mean(lisa_weekend$z_TRIPS), lty = 2) +
  scale_color_manual(
    values = c("High-High" = "red",
               "Low-Low" = "blue",
               "Low-High" = "lightblue",
               "High-Low" = "pink")) +
  labs(x = "Standardised TRIPS",
       y = "Standardised Spatial Lag of TRIPS",
       title = "Standardised Moran Scatterplot with LISA Quadrants") +
  theme_minimal() +
  theme(legend.position = "bottom")

# Create the bar chart for quadrant counts
quadrant_counts <- lisa_weekend %>%
  st_drop_geometry() %>%
  count(mean) %>%
  mutate(mean = factor(mean, levels = c("High-High", "Low-High", 
                                         "Low-Low", "High-Low")))

bar_plot <- ggplot(quadrant_counts, aes(x = mean, y = n, fill = mean)) +
  geom_bar(stat = "identity", alpha = 0.8) +
  geom_text(aes(label = n), vjust = -0.5, size = 4, fontface = "bold") +
  scale_fill_manual(
    values = c("High-High" = "red",
               "Low-Low" = "blue",
               "Low-High" = "lightblue",
               "High-Low" = "pink")) +
  labs(x = "LISA Quadrant",
       y = "Count of Hexagons",
       title = "Count by Quadrant") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

# Combine plots side by side
combined_plot <- moran_plot + bar_plot + 
  plot_layout(widths = c(2, 1)) +
  plot_annotation(
    title = "Local Moran's I Analysis - Weekend/holiday",
    theme = theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5))
  )

combined_plot
```

The Moran scatterplot reveals distinct spatial patterns in weekend/holiday bus trips, with the bar chart quantifying the distribution across LISA quadrants. The Low-Low quadrant (dark blue, 339 hexagons) dominates most strongly, representing cold spots where low-trip hexagons cluster together in peripheral or less populated regions, exhibiting strong positive spatial autocorrelation. This is the highest across all periods analyzed, reflecting reduced overall transit demand during weekends and holidays when commuter-driven patterns are absent.

The Low-High quadrant (light blue, 222 hexagons) remains substantial, representing underperforming hexagons within busier regions. This indicates considerable negative spatial autocorrelation where low-activity residential areas neighbor high-activity leisure destinations, reflecting transitional zones and spatial heterogeneity as weekend travelers concentrate at specific recreational, retail, or tourist hubs rather than dispersing across residential zones.

The High-High quadrant (red, 217 hexagons) shows moderate clustering—high-trip hexagons surrounded by similarly high-trip neighbors at major weekend destinations like shopping districts, recreational areas, and tourist attractions. This differs fundamentally from weekday patterns, reflecting leisure-driven rather than employment-related clustering. The High-Low quadrant (pink, 46 hexagons) contains minimal outliers—isolated high-trip leisure destinations surrounded by low-activity surroundings.

The positive regression slope confirms overall positive spatial autocorrelation, though the distribution reveals greater spatial complexity in weekend travel patterns. The dominance of Low-Low clusters (339 of 824 hexagons, 41%) alongside substantial Low-High outliers reflects Singapore's fragmented weekend travel landscape, with concentrated activity at specific leisure destinations contrasting sharply with widespread low-demand residential and peripheral areas lacking weekend commuter flows.

### LISA Map

```{r fig.width=12, fig.height=5}
# Count significant quadrants
quadrant_counts <- lisa_weekend_sig %>%
  st_drop_geometry() %>%
  count(mean) %>%
  mutate(mean = factor(mean, levels = c("High-High", "Low-High", 
                                         "Low-Low", "High-Low")))

# Bar chart
bar_chart <- ggplot(quadrant_counts, aes(x = mean, y = n, fill = mean)) +
  geom_bar(stat = "identity", width = 0.7) +
  geom_text(aes(label = n), vjust = -0.3, size = 5, fontface = "bold") +
  scale_fill_manual(
    values = c("High-High" = "red",
               "Low-Low" = "blue",
               "Low-High" = "lightblue",
               "High-Low" = "pink")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(x = "LISA Quadrant", 
       y = "Count of Hexagons", 
       title = "LISA Quadrant Counts") +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    axis.text.y = element_text(size = 11),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 13, face = "bold", hjust = 0.5),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.margin = margin(10, 10, 10, 10)
  )

# LISA map
tmap_mode("plot")
lisa_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  
  tm_shape(lisa_weekend_sig) +
  tm_polygons(fill = "mean",
              fill.scale = tm_scale_categorical(
                values = c("High-High" = "red",
                          "Low-Low" = "blue",
                          "Low-High" = "lightblue",
                          "High-Low" = "pink")
              ),
              fill.legend = tm_legend(
                title = "LISA Quadrants",
                position = tm_pos_in("right", "bottom")
              )
  ) +
  tm_borders(col = "black", lwd = 0.5) +
  tm_shape(mpsz_area) +
  tm_borders(col = "grey30", lwd = 1.5) +
  tm_layout(
    main.title = "LISA Map (p-value < 0.1)",
    main.title.size = 1.2,
    main.title.fontface = "bold",
    legend.text.size = 0.8,
    legend.title.size = 1,
    inner.margins = c(0.05, 0.05, 0.05, 0.05)
  )

# Convert tmap to grob
lisa_map_grob <- tmap_grob(lisa_map)

# Combine
combined_plot <- plot_grid(
  lisa_map_grob, 
  bar_chart, 
  ncol = 2, 
  rel_widths = c(2,1.5),
  labels = NULL
)

title <- ggdraw() + 
  draw_label(
    "Local Indicators of Spatial Association (LISA) - Weekend/holiday",
    fontface = 'bold',
    size = 15,
    x = 0.5,
    hjust = 0.5
  )

final_plot <- plot_grid(
  title, 
  combined_plot,
  ncol = 1,
  rel_heights = c(0.06, 0.94)
)

final_plot
```

The LISA map reveals significantly different spatial clustering patterns for weekday afternoon bus trips compared to morning peaks. The distribution is remarkably balanced, with High-High clusters (red, 68 hexagons), Low-High outliers (light blue, 69 hexagons), and Low-Low clusters (dark blue, 69 hexagons) nearly equal in number. This contrasts sharply with morning patterns where Low-Low and High-High dominated, indicating greater spatial heterogeneity in afternoon travel.

High-High clusters appear more scattered across northern, central, and eastern planning areas, representing residential zones receiving concentrated return commuter flows. However, their reduced prevalence (68 vs 145 morning) reflects more dispersed evening destination patterns. Low-Low clusters remain present in peripheral and less populated regions but are substantially fewer (69 vs 150 morning), suggesting broader spatial distribution of afternoon activity.

The dramatic increase in Low-High outliers (69 vs 78 morning) indicates extensive spatial mixing where low-activity hexagons neighbor high-activity areas, reflecting transitional zones and fragmented clustering as evening commuters disperse from centralized employment hubs to diverse residential locations. High-Low outliers remain minimal (4 hexagons).

The balanced quadrant distribution reveals weaker positive spatial autocorrelation and greater spatial complexity compared to morning peaks. This reflects Singapore's polycentric structure and varied evening travel behaviors, with afternoon trip patterns exhibiting more decentralized, heterogeneous spatial organization.

## 6.3 Hot Spots and Cold Spots Analysis (HCSA)

The code below builds a fixed distance–band neighbourhood for each weekend/holiday hexagon, creates row-standardised spatial weights, and then computes the Local Getis-Ord Gi* statistic for TRIPS.

```{r}
weekend_fdw <- weekend %>%
  mutate(nb = include_self(
    st_dist_band(
      st_geometry(geometry),
      upper = critical_dist)),
    wt = st_weights(nb,
                    style = "W"),
    .before = 1)

# Compute local Gi*
set.seed(1234)

HCSA_fdw <- weekend_fdw %>%
  mutate(
    gistar = local_gstar_perm(
      TRIPS, nb, wt, nsim = 99),
    .before = 1) %>%
  unnest(gistar)
```

```{r}
# Filter for significant Gi* values (p < 0.1)
HCSA_sig <- HCSA_fdw %>%
  filter(p_sim < 0.1)

# Plot Gi* values for significant hexagons only
Gi_star_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  
  tm_shape(HCSA_sig) +
  tm_polygons(fill = "gi_star",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 5,
                values = "-brewer.rd_bu"),
              fill.legend = tm_legend(
                title = "local Gi*",
                position = tm_pos_in("right", "bottom")
              )
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_shape(mpsz_area) +
  tm_borders(col = "grey30", lwd = 0.2) +
  tm_title("Local Gi* of TRIPS (p-value < 0.1)") +
  tm_compass(position = tm_pos_in("left", "top")) +
  tm_scale_bar(position = tm_pos_in("left", "bottom"))

# HCSA cluster classification with p < 0.1
HCSA_fdw <- HCSA_fdw %>%
  mutate(HCSA_cluster = case_when(
    p_sim <= 0.1 & cluster == "High" ~ "Hot spot",
    p_sim <= 0.1 & cluster == "Low" ~ "Cold spot",
    TRUE ~ "Other"),
    HCSA_cluster = factor(
      HCSA_cluster,
      levels = c("Hot spot", "Cold spot")  # Reordered levels
    ),
    .before = 1
  )

# Filter to keep only significant hexagons
HCSA_sig <- HCSA_fdw %>%
  filter(HCSA_cluster != "Insignificant")

# Aggregate mpsz by PLN_AREA_N to get planning area boundaries
mpsz_planning <- mpsz %>%
  group_by(PLN_AREA_N) %>%
  summarise(geometry = st_union(geometry))

# HCSA cluster map
HCSA_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  
  tm_shape(HCSA_sig) +
  tm_polygons(
    fill = "HCSA_cluster",
    fill.scale = tm_scale_categorical(
      values = c(
        "Hot spot" = "red",
        "Cold spot" = "blue"
      )
    ),
    fill.legend = tm_legend(
      title = "HCSA Cluster",
      position = tm_pos_in("right", "bottom"))
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_shape(mpsz_area) +
  tm_borders(col = "grey30", lwd = 0.2) +
  tm_title("HCSA Clusters (p-value < 0.1)")

tmap_arrange(Gi_star_map, HCSA_map, asp = 1, ncol = 2)
```

The Local Gi* map identifies scattered hot spots (red/orange) across central, eastern, and selected northern regions indicating high weekend trip clustering intensity at major recreational destinations, shopping districts, and tourist attractions. Cold spots (blue) dominate western and peripheral areas with consistently low trips. The dispersed pattern suggests destination-specific intensity variations reflecting leisure-driven travel to concentrated activity hubs rather than origin-based residential commuter flows characteristic of weekdays.

The HCSA Clusters map reveals statistically significant spatial clustering of weekend/holiday bus trips (p < 0.1), showing fundamentally different patterns from weekday periods. Hot spots (red) appear as scattered, moderately-sized clusters across northern, central, and eastern planning areas, representing major weekend destinations attracting concentrated visitor flows—shopping malls, recreational facilities, tourist attractions, and entertainment zones. These clusters are less extensive than weekday morning patterns but more defined than weekday afternoons, reflecting concentrated leisure activity at specific destinations. Cold spots (blue) remain prominent across western, peripheral, and predominantly residential regions, confirming spatial patterns of reduced weekend transit demand in areas lacking major recreational attractions.

Together, the maps demonstrate destination-focused spatial organization in weekend trip generation compared to weekday patterns. The Gi* map shows clustering intensity at leisure hubs, while the HCSA map reveals moderate-sized, destination-specific hot spot clusters. 

## 6.4 Emerging Hot Spot Analysis (EHSA)

The code below prepares weekend/holiday (11–20h) trips for Emerging Hot Spot/space–time analysis. It aggregates trips by hexagon × hour, then ensure every hexagon has all hours (filling missing cells with 0) and orders IDs/time consistently. It converts IDs and hours to the right types, then builds a space–time cube (spacetime) by linking the hourly table to the hexagon geometries, and finally checks that the result is a valid space–time cube.

```{r}
hex_ids <- hexagon$HEX_ID

trips_weekend_hourly <- trips %>%
  filter(DAY_TYPE == "WEEKENDS/HOLIDAY", HOUR_OF_DAY %in% 11:20) %>%
  group_by(HEX_ID, HOUR_OF_DAY) %>%
  summarise(TRIPS = sum(TRIPS), .groups = "drop") %>%
  tidyr::complete(HEX_ID = hex_ids, HOUR_OF_DAY = 11:20,
                  fill = list(TRIPS = 0)) %>%
  arrange(HEX_ID, HOUR_OF_DAY)

trips_weekend_hourly$HEX_ID <- factor(trips_weekend_hourly$HEX_ID, levels = hex_ids)
trips_weekend_hourly$HOUR_OF_DAY <- as.integer(trips_weekend_hourly$HOUR_OF_DAY)

# Build space–time cube
TRIPS_weekend_st <- spacetime(
  trips_weekend_hourly,
  hexagon %>% select(HEX_ID), 
  .loc_col  = "HEX_ID",
  .time_col = "HOUR_OF_DAY"
)

is_spacetime_cube(TRIPS_weekend_st) 
```

The code below adds spatial neighbours/weights to the weekend/holiday space–time cube and computes hourly Local Getis-Ord Gi* for TRIPS. It then flattens the results so each hexagon–hour has a Gi* score and significance.

```{r}
# Add neighbours and weights
TRIPS_weekend_nb <- TRIPS_weekend_st %>%
  activate("geometry") %>%
  mutate(
    nb = include_self(st_contiguity(geometry)), 
    wt = st_inverse_distance(nb, geometry, scale = 1, alpha = 1),
    .before = 1
  ) %>%
  set_nbs("nb") %>%
  set_wts("wt")

# Compute Gi* 
set.seed(1234)

gi_weekend <- TRIPS_weekend_nb %>%
  group_by(HOUR_OF_DAY) %>%
  mutate(gi_star = local_gstar_perm(TRIPS, nb, wt, nsim = 99)) %>%
  tidyr::unnest(gi_star) %>%
  ungroup()
```

### Mann-Kendall Test

The code below tests whether each hexagon’s Gi* values show a monotonic time trend (consistently increasing or decreasing) across the selected hours.

H0: No monotonic trend

H1: Monotonic trend is present

alpha: 0.1

```{r}
mk_hotspots_weekend <- gi_weekend %>%
  st_drop_geometry() %>%
  group_by(HEX_ID) %>%
  summarise(
    mk = if(n() >= 3) list(unclass(Kendall::MannKendall(gi_star))) else NA,
    .groups = "drop"
  ) %>%
  tidyr::unnest_wider(mk) %>%
  rename(tau = tau, p_value = sl)

mk_hotspots_weekend_ranked <- mk_hotspots_weekend %>%
  filter(!is.na(p_value)) %>%
  arrange(p_value, desc(abs(tau)))        

mk_hotspots_weekend_ranked %>% slice(1:10)
```

With reference to the results, as p-value < alpha, we will reject the null hypothesis and infer that there is a significant trend.

### Performing Emerging Hotspot Analysis

The code below runs EHSA on the weekday–morning space-time cube to classify each hexagon into temporal hotspot/coldspot types.

```{r}
set.seed(1234)

ehsa_weekend <- emerging_hotspot_analysis(
  x     = TRIPS_weekend_nb,   
  .var  = "TRIPS",     
  k     = 1,      
  nsim  = 99   
)
```

The code below attaches EHSA results back to the hexagon layer and keep only statistically significant cells for mapping.

```{r}
ehsa_weekend_sf <- hexagon %>%
  select(HEX_ID) %>%
  left_join(ehsa_weekend, by = c("HEX_ID" = "location")) %>%
  filter(p_value < 0.1) %>%
  st_as_sf()
```

The code below applies the consistent legend order and color scheme to the data before plotting.

```{r}
ehsa_weekend$classification   <- factor(ehsa_weekend$classification,   levels = ehsa_levels)
ehsa_weekend_sf$classification <- factor(ehsa_weekend_sf$classification, levels = ehsa_levels)
```

#### Distribution of EHSA Classes

```{r}
ggplot(data = ehsa_weekend, 
       aes(x = classification, fill = classification)) +
  geom_bar(color = "black", alpha = 0.8) +
  scale_fill_manual(values = ehsa_colors, drop = FALSE) +
  labs(
    title = "Distribution of EHSA Classes – Weekend/Holiday (11am–8pm)",
    x = "EHSA Classification", 
    y = "Number of Hexagons"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", size = 14)
  )
```

st hexagons are classified as no pattern detected, indicating that weekend/holiday demand does not form stable clusters across much of the city. Among the significant classes, persistent coldspots and sporadic coldspots are the next most common, pointing to areas where trip counts are consistently low or intermittently low. Sporadic hotspots also appear in meaningful numbers, reflecting localized, irregular bursts of higher demand. In contrast, persistent hotspots are few, and consecutive hotspots/coldspots plus new hotspots/coldspots are rare, showing that newly emerging or sustained high-demand clusters are uncommon in this period. A small NA group reflects locations without enough information to classify.

Overall, the distribution suggests a spatially uneven weekend/holiday pattern dominated by low-demand clusters, punctuated by isolated, short-lived hotspots. This aligns with discretionary travel behavior: trips are less governed by routine commutes and more by ad-hoc activities, producing irregular surges at select nodes while most areas remain stably low.

#### EHSA map and Gi* Trends

```{r fig.width=16, fig.height=6}
# Create the EHSA map
tmap_mode("plot")
ehsa_map <- tm_shape(mpsz) + 
  tm_polygons(col = "grey90", border.col = "white") +
  tm_shape(ehsa_weekend_sf) +
  tm_polygons(
    fill = "classification",
    fill.scale = tm_scale_categorical(values = ehsa_colors),
    fill.legend = tm_legend(
      title = "EHSA Classification",
      position = tm_pos_in("right", "bottom"))
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_shape(mpsz_planning) +
  tm_borders(col = "grey30", lwd = 1.5) +
  tm_layout(
    main.title = "EHSA Map",
    main.title.size = 1.2,
    main.title.fontface = "bold"
  )

# Get one representative hexagon from each significant EHSA classification
representative_hexagons <- ehsa_weekend_sf %>%
  st_drop_geometry() %>%
  filter(classification != "no pattern detected") %>%
  group_by(classification) %>%
  slice(1) %>%
  ungroup() %>%
  select(HEX_ID, classification)

# Extract Gi* trends for these hexagons
hex_trends <- gi_weekend %>%
  st_drop_geometry() %>%
  filter(HEX_ID %in% representative_hexagons$HEX_ID) %>%
  select(HEX_ID, HOUR_OF_DAY, gi_star) %>%
  left_join(representative_hexagons, by = "HEX_ID")

# Create individual plots for each classification type
plot_list <- list()

for (class in unique(hex_trends$classification)) {
  hex_data <- hex_trends %>% filter(classification == class)
  hex_id <- unique(hex_data$HEX_ID)
  
  p <- ggplot(data = hex_data, aes(x = HOUR_OF_DAY, y = gi_star)) +
    geom_line(color = ehsa_colors[class], linewidth = 1.2) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
    labs(
      title = paste0(class, "\n(", hex_id, ")"),
      x = "Hour of Day",
      y = "Gi*"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", size = 10, hjust = 0.5),
      axis.title = element_text(size = 9),
      axis.text = element_text(size = 8)
    )
  
  plot_list[[class]] <- p
}

# Combine trend plots
combined_trends <- wrap_plots(plot_list, ncol = 2) +
  plot_annotation(
    title = "Gi* Trends by EHSA Classification",
    theme = theme(
      plot.title = element_text(face = "bold", size = 12, hjust = 0.5)
    )
  )

# Convert tmap to grob
ehsa_map_grob <- tmap_grob(ehsa_map)

# Combine map and trends side by side
final_plot <- plot_grid(
  ehsa_map_grob,
  combined_trends,
  ncol = 2,
  rel_widths = c(1, 1),
  labels = NULL
)

# Add overall title
title <- ggdraw() + 
  draw_label(
    "Emerging Hot Spot Analysis (EHSA) – Weekend/holiday (11am–8pm)",
    fontface = 'bold',
    size = 15,
    x = 0.5,
    hjust = 0.5
  )

# Combine title with plots
final_combined <- plot_grid(
  title,
  final_plot,
  ncol = 1,
  rel_heights = c(0.05, 0.95)
)

final_combined
```

The EHSA analysis reveals distinct spatial and temporal clustering patterns of weekend/holiday bus trips (11am-8pm), showing fundamentally different characteristics from weekday commuter patterns. The map (left) displays blue coldspots scattered across western, northern, and peripheral planning areas, indicating persistently low weekend trip generation in predominantly residential zones. Red and orange hotspots appear fragmented and dispersed across central, eastern, and selected northern regions, representing major recreational destinations, shopping districts, and tourist attractions. These hotspots are less concentrated than weekday patterns, reflecting leisure-driven rather than employment-driven travel dynamics.

The Gi* trend plots (right) demonstrate diverse temporal behaviors across classifications throughout the 11am-8pm window. Persistent hotspots (dark red, H0003) maintain strongly positive Gi* values (4-5), showing stable high-demand zones at major weekend destinations. Sporadic hotspots (orange, H0017 and H0010) exhibit fluctuating positive values, suggesting intermittent clustering as visitors arrive at leisure destinations at varying times throughout the day.

Persistent coldspots (dark blue, H0006) remain consistently negative (-0.5 to -1.0), confirming stable low-demand areas in residential zones lacking recreational attractions. Sporadic coldspots (light blue, H0468) show variable patterns, while new hotspots (pink, H0513) emerge with positive trends, representing areas developing high-demand patterns specifically during weekend leisure hours. Consecutive hotspots (red, H0110) maintain elevated Gi* values (4-5), indicating sustained demand at popular weekend destinations.

# 7 Conclusion

This study shows that Singapore’s bus demand is highly uneven across space and time, with three consistent signals. First, trip volumes are strongly right-skewed in every period: most hexagons generate modest activity while a small set of locations—typically major residential towns, transfer nodes, and mixed-use hubs—account for a disproportionate share of trips. Second, spatial autocorrelation is present but varies by period. Weekday mornings display the clearest and most contiguous clustering: high–high clusters align with large housing corridors feeding central employment areas, while peripheral/industrial belts form low–low clusters. Third, evening and weekend patterns are more fragmented. In the afternoon peak, high–high clusters shrink and disperse, and low–high outliers rise—evidence of transitional edges and mixed land-use environments as commuters spread across diverse destinations. On weekends/holidays, demand becomes leisure-oriented: persistent cold spots dominate many peripheral zones, while hotspots concentrate around retail, recreation, and tourist areas, often as sporadic rather than persistent classes.

Methodologically, the combined use of Local Moran’s I, Gi * , HCSA and EHSA provides complementary lenses: LISA highlights where similarity clusters and outliers sit; Gi* quantifies the intensity of local concentration; HCSA filters Gi* to statistically robust hot/cold spots; and EHSA adds a temporal dimension, distinguishing persistent, consecutive, new, and sporadic patterns. The Mann–Kendall test then probes whether within-period Gi* trajectories trend monotonically. Together, these tools reveal that persistent hotspots are relatively rare outside the weekday morning window; many high-demand areas behave sporadically or shift through the peak, and only a limited number of hexagons exhibit clear monotonic trends within the examined hours.

Planning implications follow directly. For weekday mornings, strengthening trunk capacity and feeder reliability in established high–high corridors is likely to yield the greatest benefit, while targeted coverage and frequency rationalisation may suit persistent low–low belts. Afternoon operations should emphasise flexibility: interlined routes, dynamic short-turns, or headway-based control can better accommodate the dispersed, heterogeneous pattern and elevated share of outliers. On weekends/holidays, service design should pivot toward destination-driven mobility—timed frequencies to malls, parks, waterfronts, and event hubs—while maintaining baseline coverage in stable cold-spot neighborhoods. Across all periods, monitoring sporadic hotspots identified by EHSA is critical for event-responsive capacity and passenger information.