---
title: "Take home exercise 2"
author: "Tai Qiu Yan"
date: "October 8, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# 1 Introduction

# 2 Loading the R package

```{r}
pacman::p_load(tidyverse, sf, sfdep, tmap, knitr, kableExtra, DT)
```

# 3 Data Importing and Wrangling

## Importing Data

```{r}
#Bus Stop Location
BusStop = st_read(dsn = "data/LTADataMall/", 
                  layer = "BusStop") %>%
  st_transform(crs = 3414)

#Master Plan 2019 Subzone
mpsz = read_rds("data/rds/mpsz_sf.rds")

#O-D Bus Trips
odbus <- read_csv("data/LTADataMall/origin_destination_bus_202508.csv")
```

## Extracting Bus Stops located within Singapore

```{r}
BusStop_in_SG <- st_join(
  BusStop, mpsz, 
  join = st_within, 
  left = FALSE)
```

### Visualize

```{r}
tm_shape(mpsz) +
  tm_polygons() +
tm_shape(BusStop_in_SG) +
  tm_dots()
```

## Deriving Analytical Hexagon

```{r}
hexagon <- st_make_grid(mpsz, 
                        cellsize = 700,
                        what = "polygon",
                        square = FALSE) %>%
  st_sf()
```

## Selecting hexagons with bus stops

```{r}
hexagon$busstop_count =
  lengths(st_intersects(
    hexagon, BusStop_in_SG))

hexagon <- filter(
  hexagon, busstop_count > 0)
```

## Assigning ids to each hexagon

```{r}
hexagon <- hexagon %>% 
  select(, -busstop_count)

hexagon$HEX_ID <- sprintf(
  "H%04d", seq_len(
    nrow(hexagon))) %>% 
  as.factor()
```

## Preparing Trip Generation Data

```{r}
trips <- odbus %>%
  select(c(ORIGIN_PT_CODE, DAY_TYPE, TIME_PER_HOUR, TOTAL_TRIPS)) %>%
  rename(BUS_STOP_N = ORIGIN_PT_CODE) %>%
  rename(HOUR_OF_DAY = TIME_PER_HOUR) %>%
  rename(TRIPS = TOTAL_TRIPS)
trips$BUS_STOP_N <- as.factor(trips$BUS_STOP_N)
```

## Populating Hexagon IDs into BusStop data frame

```{r}
bs_hex <- st_intersection(
  BusStop, hexagon) %>%
  st_drop_geometry() %>%
  select(c(BUS_STOP_N, HEX_ID))
```

## Adding HEX_ID into bus trips data

```{r}
trips <- inner_join(trips, bs_hex)
```

## Aggregating TRIPS based on HEX_ID

```{r}
trips <- trips %>%
  group_by(
    HEX_ID,
    DAY_TYPE,
    HOUR_OF_DAY) %>%
  summarise(TRIPS = sum(TRIPS))
```

# 4 Weekday morning peak

## 4.1 Overview

### Choropleth map of total trips

```{r}
weekday_morning <- trips %>%
  filter(DAY_TYPE == "WEEKDAY", HOUR_OF_DAY %in% 6:8) %>%
  group_by(HEX_ID) %>%
  summarise(TRIPS = sum(TRIPS), .groups = "drop") %>%
  left_join(hexagon, by = "HEX_ID") %>%
  st_as_sf()

tmap_mode("plot")

tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  
  tm_shape(weekday_morning) +
  tm_polygons(
    fill = "TRIPS",
    fill.scale = tm_scale_intervals(
      style = "quantile",  # Use style = "quantile" instead
      n = 5,
      values = "viridis"
    ),
    fill.legend = tm_legend(
      title = "Total Trips\n(Weekday Morning)",
      position = tm_pos_in("right", "bottom")
    )
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_title("Distribution of Bus Trips - Weekday Morning Peak (6am-9am)") +
  tm_compass(position = tm_pos_in("left", "top")) +
  tm_scale_bar(position = tm_pos_in("left", "bottom"))
```

### Trip Distribution

```{r}
# Understanding the distribution of trips with improved formatting
library(scales)

# Calculate statistics
mean_trips <- mean(weekday_morning$TRIPS)
median_trips <- median(weekday_morning$TRIPS)
q75 <- quantile(weekday_morning$TRIPS, 0.75)
q95 <- quantile(weekday_morning$TRIPS, 0.95)

# Create improved histogram
ggplot(weekday_morning, aes(x = TRIPS)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(aes(xintercept = mean_trips), 
             color = "red", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = median_trips), 
             color = "orange", linetype = "dashed", size = 1) +
  # Add text labels for mean and median
  annotate("text", x = mean_trips, y = Inf, 
           label = paste0("Mean: ", round(mean_trips, 0)), 
           vjust = 2, hjust = -0.1, color = "red", size = 3.5) +
  annotate("text", x = median_trips, y = Inf, 
           label = paste0("Median: ", round(median_trips, 0)), 
           vjust = 4, hjust = -0.1, color = "orange", size = 3.5) +
  # Format x-axis with comma separators
  scale_x_continuous(
    labels = label_number(scale = 1/1000, suffix = "k"),
    breaks = seq(0, max(weekday_morning$TRIPS), by = 20000)
  ) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Distribution of Trip Counts - Weekday Morning Peak (6am-9am)",
    subtitle = paste0("n = ", nrow(weekday_morning), " hexagons | 95th percentile: ", 
                     round(q95, 0), " trips"),
    x = "Number of Trips",
    y = "Frequency (Number of Hexagons)") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10, color = "gray40"),
    axis.title = element_text(size = 11),
    panel.grid.minor = element_blank()
  )
```

## 4.2 Local Indicators of Spatial Association(LISA)

```{r}
# Build spatial weights (Queen contiguity)
nb <- st_contiguity(weekday_morning)
wt <- st_weights(nb, style = "W", allow_zero = TRUE)

# Compute Local Moran's I
set.seed(1234)
lisa_morning <- weekday_morning %>%
  mutate(local_moran = local_moran(
    TRIPS, nb, wt, nsim = 99
  ), .before = 1) %>%
  unnest(local_moran)

# Result
glimpse(lisa_morning)
```

### Local Moran’s I values and p-values

```{r}
ii_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +  # base layer

  tm_shape(lisa_morning) +
  tm_polygons(fill = "ii",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 5,
                values = "brewer.RdBu"
              ),
              fill.legend = tm_legend(
                title = "Local Moran's I",
                position = tm_pos_in("right", "bottom")
              )
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_title("Local Moran's I - morning peak")

p_ii_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +

  tm_shape(lisa_morning) +
  tm_polygons(fill = "p_ii",
              fill.scale = tm_scale_intervals(
                breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
                values = "-brewer.Reds"
              ),
              fill.legend = tm_legend(
                title = "p-value",
                position = tm_pos_in("right", "bottom")
              )
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_title("p-values of Local Moran's I")

tmap_arrange(ii_map, p_ii_map, asp = 1, ncol = 2)
```

### Moran scatterplot with standardised variable

```{r}
lisa_morning <- lisa_morning %>%
  mutate(lag_TRIPS = st_lag(
    TRIPS, nb, wt),
    .before = 1) %>%
  unnest(lag_TRIPS)

lisa_morning <- lisa_morning %>%
  mutate(z_TRIPS = scale(TRIPS),
         z_lag_TRIPS = scale(lag_TRIPS),
         .before = 1)

ggplot(data = lisa_morning,
       aes(x = z_TRIPS,
           y = z_lag_TRIPS,
           color = mean)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm",
              se = FALSE,
              color = "black") +
  geom_hline(yintercept = mean(lisa_morning$z_lag_TRIPS), lty = 2) +
  geom_vline(xintercept = mean(lisa_morning$z_TRIPS), lty = 2) +
  scale_color_manual(
    values = c("High-High" = "red",
               "Low-Low" = "blue",
               "Low-High" = "lightblue",
               "High-Low" = "pink")) +
  labs(x = "Standardised TRIPS",
       y = "Standardised Spatial Lag of TRIPS",
       title = "Standardised Moran Scatterplot with LISA Quadrants") +
  theme_minimal()
```


### LISA Map

```{r}
lisa_sig <- lisa_morning %>%
  filter(p_ii < 0.05)

tmap_mode("plot")

tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  
  tm_shape(lisa_morning) +
  tm_polygons(fill = "TRIPS",
              fill.scale = tm_scale_continuous(
                values = "greys"
              ),
              fill.legend = tm_legend(
                title = "Total TRIPS",
                position = tm_pos_in("right", "bottom")
              )
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  
  tm_shape(lisa_sig) +
  tm_polygons(fill = "mean",
              fill.scale = tm_scale_categorical(
                values = c("High-High" = "red",
                          "Low-Low" = "blue",
                          "Low-High" = "lightblue",
                          "High-Low" = "pink")
              ),
              fill.legend = tm_legend(
                title = "LISA Quadrants (p < 0.05)",
                position = tm_pos_in("left", "bottom")
              )
  ) +
  tm_borders(col = "black", lwd = 0.5) +
  tm_title("LISA Map - Weekday Morning Peak (6am-9am)")
```

## 4.3 Hot Spots and Cold Spots Analysis (HCSA)

### Gi* values and p-values

```{r}
# Compute fixed distance weight matrix
ct <- critical_threshold(st_geometry(weekday_morning))

weekday_morning_fdw <- weekday_morning %>%
  mutate(nb = include_self(
    st_dist_band(
      st_geometry(geometry),
      upper = ct)),
    wt = st_weights(nb,
                    style = "W"),
    .before = 1)

# Compute local Gi*

set.seed(1234)

HCSA_fdw <- weekday_morning_fdw %>%
  mutate(
    gistar = local_gstar_perm(
      TRIPS, nb, wt, nsim = 99),
    .before = 1) %>%
  unnest(gistar)

# Plot Gi* values
Gi_star_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  
  tm_shape(HCSA_fdw) +
  tm_polygons(fill = "gi_star",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 5,
                values = "-brewer.rd_bu"),
              fill.legend = tm_legend(
                title = "local Gi*",
                position = tm_pos_in(
                  "right", "bottom"))) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_title("Local Gi* of TRIPS")

# Plot p-values
p_values_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  
  tm_shape(HCSA_fdw) +
  tm_polygons(fill = "p_sim",
              fill.scale = tm_scale_intervals(
                breaks = c(0, 0.001, 0.01, 0.05, 0.1, 1),
                values = "-brewer.reds"),
              fill.legend = tm_legend(
                title = "p-value",
                position = tm_pos_in("right", "bottom"))) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_title("p-values of local Gi* of TRIPS (Fixed distance)")

# Display both maps side by side
tmap_arrange(Gi_star_map, p_values_map, asp = 1, ncol = 2)
```

### HCSA Map

```{r}
# HCSA cluster classification
HCSA_fdw <- HCSA_fdw %>%
  mutate(HCSA_cluster = case_when(
    p_sim > 0.05 ~ "Insignificant",
    p_sim <= 0.05 & cluster == "High" ~ "Hot spot",
    p_sim <= 0.05 & cluster == "Low" ~ "Cold spot",
    TRUE ~ "Other"),
    HCSA_cluster = factor(
      HCSA_cluster,
      levels = c("Insignificant", "Hot spot", "Cold spot")
    ),
    .before = 1
  )

# Plot HCSA cluster map
HCSA_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  
  tm_shape(HCSA_fdw) +
  tm_polygons(
    fill = "HCSA_cluster",
    fill.scale = tm_scale_categorical(
      values = c(
        "grey80",  # Insignificant
        "red",     # Hot spot
        "blue"     # Cold spot
      )
    ),
    fill.legend = tm_legend(
      title = "HCSA Cluster",
      position = tm_pos_in("right", "bottom"))
  ) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_title("HCSA Clusters (p < 0.05)")

# Gi* map and HCSA map
tmap_arrange(Gi_star_map, HCSA_map, asp = 1, ncol = 2)
```




### Emerging Hot Spot Analysis (EHSA)

```{r}

```







# comparasion

```{r}
# Prepare data for all three periods
weekday_afternoon <- trips %>%
  filter(DAY_TYPE == "WEEKDAY", HOUR_OF_DAY %in% 17:19) %>%
  group_by(HEX_ID) %>%
  summarise(TRIPS = sum(TRIPS), .groups = "drop") %>%
  left_join(hexagon, by = "HEX_ID") %>%
  st_as_sf()

weekend <- trips %>%
  filter(DAY_TYPE %in% c("WEEKENDS/HOLIDAY"), HOUR_OF_DAY %in% 11:19) %>%
  group_by(HEX_ID) %>%
  summarise(TRIPS = sum(TRIPS), .groups = "drop") %>%
  left_join(hexagon, by = "HEX_ID") %>%
  st_as_sf()

# Create comparative maps
morning_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  tm_shape(weekday_morning) +
  tm_polygons(fill = "TRIPS",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "viridis"
              ),
              fill.legend = tm_legend(title = "Trips")) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_title("Weekday Morning\n(6am-9am)")

afternoon_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  tm_shape(weekday_afternoon) +
  tm_polygons(fill = "TRIPS",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "viridis"
              ),
              fill.legend = tm_legend(title = "Trips")) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_title("Weekday Afternoon\n(5pm-8pm)")

weekend_map <- tm_shape(mpsz) +
  tm_polygons(col = "grey90", border.col = "white") +
  tm_shape(weekend) +
  tm_polygons(fill = "TRIPS",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "viridis"
              ),
              fill.legend = tm_legend(title = "Trips")) +
  tm_borders(col = "grey40", lwd = 0.5) +
  tm_title("Weekend/Holiday\n(11am-8pm)")

tmap_arrange(morning_map, afternoon_map, weekend_map, ncol = 3)
```

## Boxplot

```{r}
# Combine all three periods
all_periods <- bind_rows(
  weekday_morning %>% st_drop_geometry() %>% mutate(Period = "Morning"),
  weekday_afternoon %>% st_drop_geometry() %>% mutate(Period = "Afternoon"),
  weekend %>% st_drop_geometry() %>% mutate(Period = "Weekend")
)

ggplot(all_periods, aes(x = Period, y = TRIPS, fill = Period)) +
  geom_boxplot() +
  scale_fill_viridis_d() +
  labs(title = "Comparison of Trip Distributions Across Time Periods",
       y = "Number of Trips",
       x = "Time Period") +
  theme_minimal() +
  theme(legend.position = "none")
```