---
title: "Take home exercise 1"
author: "Tai Qiu Yan"
date: "September 12, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# 1 Loading the R packages

```{r}
pacman::p_load(tidyverse, sf, tmap, httr, 
               dplyr, ggplot2, patchwork, 
               terra, spatstat, rvest)
```

# 2 Data Importing and Wrangling

## 2.1 Importing and saving ACRA data

```{r}
folder_path <- "data/aspatial/acra"
file_list <- list.files(path = folder_path,
                        pattern = "^ACRA*.*\\.csv$",
                        full.names = TRUE)

acra_data <- file_list %>%
  map_dfr(read_csv)

write_rds(acra_data,
          "data/rds/acra_data.rds")
```

## 2.2 Tidying ACRA data

4 types of business that related to Singaporean's lifestyle are selected:

SSIC 93111 – Fitness centres and gymnasiums 

SSIC 47711 – Retail sale of clothing for adults

SSIC 56111 – Restaurants and Cafés

SSIC 82303 – Event/Concert Organisers

```{r}
#Fitness centres and gymnasiums 
biz_93111 <- acra_data %>%
  select(1:24) %>%
  filter(primary_ssic_code == 93111) %>%
  rename(date = registration_incorporation_date) %>%
  mutate(date = as.Date(date),
         YEAR = year(date),
         MONTH_NUM = month(date),
         MONTH_ABBR = month(date,
                            label = TRUE,
                            abbr = TRUE)) %>%
  mutate(
    postal_code = str_pad(postal_code, 
    width = 6, side = "left", pad = "0")) %>% 
           filter(date >= as.Date("2024-01-01") & date <= as.Date("2025-06-30"))

#Retail sale of clothing for adults
biz_47711 <- acra_data %>%
  select(1:24) %>%
  filter(primary_ssic_code == 47711) %>%
  rename(date = registration_incorporation_date) %>%
  mutate(date = as.Date(date),
         YEAR = year(date),
         MONTH_NUM = month(date),
         MONTH_ABBR = month(date,
                            label = TRUE,
                            abbr = TRUE)) %>%
  mutate(
    postal_code = str_pad(postal_code, 
    width = 6, side = "left", pad = "0")) %>% 
           filter(date >= as.Date("2024-01-01") & date <= as.Date("2025-06-30"))

#Restaurants and Cafés
biz_56111 <- acra_data %>%
  select(1:24) %>%
  filter(primary_ssic_code == 56111) %>%
  rename(date = registration_incorporation_date) %>%
  mutate(date = as.Date(date),
         YEAR = year(date),
         MONTH_NUM = month(date),
         MONTH_ABBR = month(date,
                            label = TRUE,
                            abbr = TRUE)) %>%
  mutate(
    postal_code = str_pad(postal_code, 
    width = 6, side = "left", pad = "0")) %>% 
           filter(date >= as.Date("2024-01-01") & date <= as.Date("2025-06-30"))

#Event/Concert Organisers
biz_82303 <- acra_data %>%
  select(1:24) %>%
  filter(primary_ssic_code == 82303) %>%
  rename(date = registration_incorporation_date) %>%
  mutate(date = as.Date(date),
         YEAR = year(date),
         MONTH_NUM = month(date),
         MONTH_ABBR = month(date,
                            label = TRUE,
                            abbr = TRUE)) %>%
  mutate(
    postal_code = str_pad(postal_code, 
    width = 6, side = "left", pad = "0")) %>% 
           filter(date >= as.Date("2024-01-01") & date <= as.Date("2025-06-30"))
```

## 2.3 Geocoding

found: successfully geocoded not found: unsuccessfully geocoded

::: panel-tabset

### Fitness centres and gymnasiums 

```{r}
#Geocoding
postcodes <- unique(biz_93111$postal_code)

url <- "https://onemap.gov.sg/api/common/elastic/search"

found <- data.frame()
not_found <- data.frame(postcode = character())

for (pc in postcodes) {
  query <- list(
    searchVal = pc,
    returnGeom = "Y",
    getAddrDetails = "Y",
    pageNum = "1"
  )

  res <- GET(url, query = query)
  json <- content(res)
  
  if (json$found != 0) {
    df <- as.data.frame(json$results, stringsAsFactors = FALSE)
    df$input_postcode <- pc
    found <- bind_rows(found, df)
  } else {
    not_found <- bind_rows(not_found, data.frame(postcode = pc))
  }
}

#Tidying the geocoded data
found <- found %>%
  select(1:10)

#Appending the location information
biz_93111 = biz_93111 %>%
  left_join(found,
            by = c('postal_code' = 'POSTAL'))

#Saving the data
write_rds(biz_93111, "data/rds/biz_93111.rds")

#Converting into SF data frame
biz_93111_sf <- st_as_sf(biz_93111,
                         coords = c("X", "Y"),
                         crs = 3414)
```

### Retail sale of clothing for adults

```{r}
#Geocoding
postcodes <- unique(biz_47711$postal_code)

url <- "https://onemap.gov.sg/api/common/elastic/search"

found <- data.frame()
not_found <- data.frame(postcode = character())

for (pc in postcodes) {
  query <- list(
    searchVal = pc,
    returnGeom = "Y",
    getAddrDetails = "Y",
    pageNum = "1"
  )

  res <- GET(url, query = query)
  json <- content(res)
  
  if (json$found != 0) {
    df <- as.data.frame(json$results, stringsAsFactors = FALSE)
    df$input_postcode <- pc
    found <- bind_rows(found, df)
  } else {
    not_found <- bind_rows(not_found, data.frame(postcode = pc))
  }
}

#Tidying the geocoded data
found <- found %>%
  select(1:10)

#Appending the location information
biz_47711 = biz_47711 %>%
  left_join(found,
            by = c('postal_code' = 'POSTAL'))

#Saving the data
write_rds(biz_47711, "data/rds/biz_47711.rds")

#Converting into SF data frame
biz_47711_sf <- st_as_sf(biz_47711,
                         coords = c("X", "Y"),
                         crs = 3414)
```

### Restaurants and Cafés

```{r}
#Geocoding
postcodes <- unique(biz_56111$postal_code)

url <- "https://onemap.gov.sg/api/common/elastic/search"

found <- data.frame()
not_found <- data.frame(postcode = character())

for (pc in postcodes) {
  query <- list(
    searchVal = pc,
    returnGeom = "Y",
    getAddrDetails = "Y",
    pageNum = "1"
  )

  res <- GET(url, query = query)
  json <- content(res)
  
  if (json$found != 0) {
    df <- as.data.frame(json$results, stringsAsFactors = FALSE)
    df$input_postcode <- pc
    found <- bind_rows(found, df)
  } else {
    not_found <- bind_rows(not_found, data.frame(postcode = pc))
  }
}

#Tidying the geocoded data
found <- found %>%
  select(1:10)

#Appending the location information
biz_56111 = biz_56111 %>%
  left_join(found,
            by = c('postal_code' = 'POSTAL'))

#Saving the data
write_rds(biz_56111, "data/rds/biz_56111.rds")

#Converting into SF data frame
biz_56111_sf <- st_as_sf(biz_56111,
                         coords = c("X", "Y"),
                         crs = 3414)
```

###Event/Concert Organisers

```{r}
#Geocoding
postcodes <- unique(biz_82303$postal_code)

url <- "https://onemap.gov.sg/api/common/elastic/search"

found <- data.frame()
not_found <- data.frame(postcode = character())

for (pc in postcodes) {
  query <- list(
    searchVal = pc,
    returnGeom = "Y",
    getAddrDetails = "Y",
    pageNum = "1"
  )

  res <- GET(url, query = query)
  json <- content(res)
  
  if (json$found != 0) {
    df <- as.data.frame(json$results, stringsAsFactors = FALSE)
    df$input_postcode <- pc
    found <- bind_rows(found, df)
  } else {
    not_found <- bind_rows(not_found, data.frame(postcode = pc))
  }
}

#Tidying the geocoded data
found <- found %>%
  select(1:10)

#Appending the location information
biz_82303 = biz_82303 %>%
  left_join(found,
            by = c('postal_code' = 'POSTAL'))

#Saving the data
write_rds(biz_82303, "data/rds/biz_82303.rds")

#Converting into SF data frame
biz_82303_sf <- st_as_sf(biz_82303,
                         coords = c("X", "Y"),
                         crs = 3414)
```

:::

### 2.3.1 Visualising the distribution

```{r fig.width=10, fig.height=6}
plot_monthly <- function(df, title_text, fill_color) {
  df %>%
    filter(date >= as.Date("2024-01-01") & date <= as.Date("2025-06-30")) %>%
    mutate(MonthYear = format(date, "%b %Y")) %>%
    mutate(MonthYear = factor(MonthYear, 
                              levels = format(seq(as.Date("2024-01-01"),
                                                  as.Date("2025-06-01"),
                                                  by = "1 month"), "%b %Y"),
                              ordered = TRUE)) %>%
    ggplot(aes(x = MonthYear)) +
    geom_bar(fill = fill_color) +   # use passed colour
    labs(
      title = title_text,
      x = "Month-Year",
      y = "Number of registrations"
    ) +
    ylim(0, 150) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title = element_text(hjust = 0.5, size = 13, face = "bold")
    )
}

# Assign unique colours
p1 <- plot_monthly(biz_93111, "Distribution of Fitness Centres and Gymnasiums", "#a6cee3")
p2 <- plot_monthly(biz_47711, "Distribution of Retail Sale of Clothing for Adults", "#fb9a99")
p3 <- plot_monthly(biz_56111, "Distribution of Restaurants and Cafés", "#b2df8a")
p4 <- plot_monthly(biz_82303, "Distribution of Event/Concert Organisers", "#fdbf6f")

# Combine 2×2
(p1 + p2) / (p3 + p4)
```


### 2.3.2 Visualising the business

```{r}
tmap_mode("view")

tm_shape(biz_93111_sf) +
  tm_dots(col = "#1f78b4", size = 0.15, alpha = 0.7, id = "entity_name", legend.show = TRUE) +
tm_shape(biz_47711_sf) +
  tm_dots(col = "#e31a1c", size = 0.15, alpha = 0.7, id = "entity_name", legend.show = TRUE) +
tm_shape(biz_56111_sf) +
  tm_dots(col = "#33a02c", size = 0.15, alpha = 0.7, id = "entity_name", legend.show = TRUE) +
tm_shape(biz_82303_sf) +
  tm_dots(col = "#ff7f00", size = 0.15, alpha = 0.7, id = "entity_name", legend.show = TRUE) +
tm_layout(legend.outside = TRUE,
          legend.title.size = 1.2,
          legend.text.size = 0.9)
```

## 2.4 Creating owin object

```{r}
mpsz_sf <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaKML.kml") %>% 
  st_zm(drop = TRUE, what = "ZM") %>% st_transform(crs = 3414)

extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}

mpsz_sf <- mpsz_sf %>%
  mutate(
    REGION_N = map_chr(Description, extract_kml_field, "REGION_N"),
    PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
    SUBZONE_N = map_chr(Description, extract_kml_field, "SUBZONE_N"),
    SUBZONE_C = map_chr(Description, extract_kml_field, "SUBZONE_C")
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())

mpsz_cl <- mpsz_sf %>%
  filter(SUBZONE_N != "SOUTHERN GROUP",
         PLN_AREA_N != "WESTERN ISLANDS",
         PLN_AREA_N != "NORTH-EASTERN ISLANDS")

write_rds(mpsz_cl, 
          "data/rds/mpsz_cl.rds")

sg_owin <- as.owin(mpsz_cl)
```

## 2.5 Combining point events object and owin object

```{r}
# Converting sf data frames to ppp class
biz_93111_ppp <- as.ppp(biz_93111_sf)
biz_47711_ppp <- as.ppp(biz_47711_sf)
biz_56111_ppp <- as.ppp(biz_56111_sf)
biz_82303_ppp <- as.ppp(biz_82303_sf)

# Combine
biz_93111_ppp = biz_93111_ppp[sg_owin]
biz_47711_ppp = biz_47711_ppp[sg_owin]
biz_56111_ppp = biz_56111_ppp[sg_owin]
biz_82303_ppp = biz_82303_ppp[sg_owin]
```


# 3 Analysis of Fitness centres and gymnasiums 

## 3.1 Clark-Evans test with CSR

The test hypotheses are:

Ho = The distribution of Fitness centres and gymnasiums are randomly distributed.

H1= The distribution of Fitness centres and gymnasiums are not randomly distributed.

The 95% confident interval will be used.

```{r}
clarkevans.test(biz_93111_ppp,
                correction="none",
                clipregion="sg_owin",
                alternative=c("clustered"),
                method="MonteCarlo",
                nsim=99)
```
Since p<0.05, we reject the null hypothesis (H0), and conclude that the distribution of Fitness centres and gymnasiums is significantly clustered.

## 3.2 Kernel Density Estimation Method

bw.ppl() algorithm is used as it tends to produce the more appropriate values when the pattern consists predominantly of tight clusters.

```{r}
# convert into km
biz_93111_ppp_km <- rescale.ppp(
  biz_93111_ppp, 1000, "km")

kde_biz_93111_km <- density(biz_93111_ppp_km,
                              sigma=bw.ppl,
                              edge=TRUE,
                              kernel="gaussian")

# Convert into raster
kde_93111_bw_terra <- rast(kde_biz_93111_km)

# Assign projection systems
crs(kde_93111_bw_terra) <- "EPSG:3414"

kde_93111_bw_terra
```

```{r}
tmap_mode('plot')

tm_shape(kde_93111_bw_terra) + 
  tm_raster(col.scale = 
              tm_scale_continuous(
                values = "viridis"),
            col.legend = tm_legend(
            title = "Density values",
            title.size = 0.7,
            text.size = 0.7,
            bg.color = "white",
            bg.alpha = 0.7,
            position = tm_pos_in(
              "right", "bottom"),
            frame = TRUE)) +
  tm_graticules(labels.size = 0.7) +
  tm_compass() +
  tm_layout(main.title = "KDE of Fitness Centres and Gymnasiums",
            scale = 1.0)
```

